### 2024-07-03 manuscript_analysis.Rmd
based on the supplementary for master thesis rmd. Will be extended with the new function etc. For now, only changes in the data analysis part

---
title: "manuscript_analysis"
author: "Sophie Moreau"
date: "2024-07-03"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(here)
library(readr)
library(viridis)

```

#Introduction
This is the supplementary information of the Master thesis. This document
covers the whole analysis performed to assess response diversity between fish communities
in Swiss peri-alpine lakes. 
This document has two parts: Data preparation (1) and Data analysis (2). 
Function to model general additive models is provided separately (see function_GAM.R). 
Code for figures used in the method figure is also provided separately (see supplementary_method_figure.R)

```{r}
# do we want to include zurich and constance separately?
projet_lac_data <- read.csv("PLDB_final_short_depth_climate_cleantaxa_20052022.csv")

str(projet_lac_data)
projet_lac_data$LakeBasin <- as.factor(projet_lac_data$LakeBasin)

projet_lac_data |> 
  distinct(LakeBasin)
```

#1. Data preparation
##Abundance data
```{r}
##read in the data: Simstrat model data and Projet Lac combined
projet_lac_data <- read.csv("PLDB_final_short_depth_climate_cleantaxa_20052022.csv")

str(projet_lac_data)

# preparing additional column with abundance data of each species
abundance_data <- projet_lac_data |> 
  # exclude rows without taxa information
  filter(!Taxa_latin_FINAL %in% "NO_FISH") |> 
  # exclude samples obtained by electro-fishing
  filter(!Protocol %in% "electro") |> 
  #select columns relevant for species individual occurrences
 group_by(Lake, Basin3, Fishec_action, mean_temp_per_day, mean_last_7days, Weightg_raw, Weightg_soak, Depth_sample, Taxa_latin_FINAL, Protocol) |> 
  # prepare column with abundance of each species
  tally() |> 
  # values_fill = 0 added to make all NAs into 0
  pivot_wider(names_from = Taxa_latin_FINAL, values_from = n, values_fill = 0)

# abundance of each species
abundance_data_long <- abundance_data |> 
  pivot_longer(cols = !(Lake : Protocol), names_to = "Species", values_to = "Abundance")

# add presence-absence column
abundance_data_long$Presence <- ifelse(abundance_data_long$Abundance >= 1, 1, 0)

# View(abundance_data_long)

max(abundance_data_long$Presence) #check if presence-absence only 0 and 1
max(abundance_data_long$Abundance)

#we only need information on species present in the lake, not all of them
abundance_data_long_lakepresence <- abundance_data_long |> 
  group_by(Lake, Species) |>
  # . stands for data being piped in do function
  do(LakePresence = max(.$Presence)) |> 
  unnest(cols = LakePresence)

# double-check 
abundance_data_long_lakepresence |>
  filter(LakePresence == "1") |> 
  group_by(Lake, LakePresence) |>
  count()

# join the two dfs
df_final <- abundance_data_long_lakepresence |> 
  filter(LakePresence == "1") |> 
  left_join(abundance_data_long )

#double-check if all lakepresence = 1
max(df_final$LakePresence)
min(df_final$LakePresence)

# save final df
saveRDS(df_final, "df_final.rds")

```

##Taxonomy and requirements
We pooled several species into ecomorphs to prevent loss of data-points. 
We also excluded species that were caught less than 10 times in the whole Projet Lac.

```{r}
# read in prepared df
df_final <- readRDS("df_final.rds")

#######refining errors in species' identity that were noticed upon examination of the
# data frame, following the table 16 in the Projet Lac "Synthesis report"

# C. brienzii should not be present in lake Thun, add to Coregonus sp.

df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Coregonus_brienzii") & Lake %in% c("Thun"),
                          "Coregonus_sp", as.character(Species)))

# Salvelinus umbla in Lake Thun should be Salvelinus sp.

df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Salvelinus_umbla") & Lake %in% c("Thun"),
                          "Salvelinus_sp", as.character(Species)))

# Squalius squalus in Biel and Neuchatel to Squalius cephalus

df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Squalius_squalus") & Lake %in% c("Biel", 
                                                                           "Neuchatel"),
                          "Squalius_cephalus", as.character(Species)))

#######Pooling species to be able to include more samples
# Profundal Cottus gobio are pooled

df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c(
    "Cottus_gobio_Profundal_Thun",
    "Cottus_gobio_Profundal_Lucerne", "Cottus_gobio_Profundal_Walen"
  ),
  "Cottus_sp_Profundal", as.character(Species)
  ))

# Salvelinus is pooled into "profundal" and "limnetic" habitats

# profundal

df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c( "Salvelinus_sp_Profundal_dwarf_Thun", 
                                          "Salvelinus_sp_Profundal_dwarf_VWS", 
                                          "Salvelinus_sp_Profundal_extreme_Thun",
    "Salvelinus_sp_Profundal_Walen_I", "Salvelinus_sp_Profundal_Walen_II"),
  "Salvelinus_sp_Profundal", as.character(Species)
  ))

# limnetic 
df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Salvelinus_sp_Limnetic_Thun", "Salvelinus_sp_Limnetic_VWS"),
                          "Salvelinus_sp_Limnetic", as.character(Species)))

# Coregonus species are pooled to new groups based on ecomorphs,
# following De-Dayne et al. 2022

# Albeli
df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Coregonus_albellus", "Coregonus_candidus", "Coregonus_confusus",
                                         "Coregonus_heglingus", "Coregonus_zugensis"),
                          "Coregonus_sp_albeli", as.character(Species)))
# Balchen
df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Coregonus_alpinus", "Coregonus_arenicolus", "Coregonus_duplex",
                        "Coregonus_helveticus", "Coregonus_palaea"),
                          "Coregonus_sp_balchen", as.character(Species)))

# Felchen

df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Coregonus_brienzii", "Coregonus_fatioi", "Coregonus_intermundia",
                                         "Coregonus_litoralis", "Coregonus_macrophthalmus", "Coregonus_zuerichensis"),
                          "Coregonus_sp_felchen", as.character(Species)))
# Large pelagic

df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Coregonus_acrinasus", "Coregonus_wartmanni"),
                          "Coregonus_sp_large_pelagic", as.character(Species)))

# Benthic profundal
df_final <- df_final |> 
  mutate(Species = ifelse(Species == "Coregonus_profundus",
                          "Coregonus_sp_benthic_profundal", as.character(Species)))
# Pelagic profundal
df_final <- df_final |> 
  mutate(Species = ifelse(Species == "Coregonus_nobilis",
                          "Coregonus_sp_pelagic_profundal", as.character(Species)))

# "Coregonus_sarnensis": could not be assigned to Albeli or Felchen

# check species list
species <- df_final |> 
  distinct(Species)

# check how many species have less than 10 overall observations: 30
df_final |> 
  group_by(Species) |> 
  summarize(tot_abu = sum(Presence)) |> 
  filter(tot_abu <= 10)

# exclusion of species with less than 10 overall observations
df_models <- df_final |> 
  group_by(Species) |> 
  mutate(tot_abu = sum(Presence)) |> 
  ungroup() |> 
  filter(tot_abu >= 10)

# save the df of the refined species, df will be used for the models
saveRDS(df_models, "df_models.rds")

df_models$Species <- as.factor(df_models$Species)
levels(df_models$Species)

species <- df_models |> 
  group_by(Species) |> 
  summarize(tot_obs = sum(Abundance))

```
##Data frames for GAMs
We compiled four data frames for the four types of general additive models that 
were run: Species with 1) binomial dataoccurring in one lake,
2) abundance data occurring in one lake, 3) binomial data 
occuring in multiple lakes, 4) abundance data and occuring in multiple lakes.

```{r}
###Overview
#1. Separation of species into binomial and abundance data: sample with abundance > 1, 0 and 1 only -> binomial data
#2. species that occur only in one lake: Lake not added as random effect

#1. binomial vs. abundance data

df_models$Species <- as.character(df_models$Species)

non_binomial_species <- df_models |> 
  filter(Abundance > 1) |> 
  group_by(Species) |>
  distinct(Species) |> 
  pull(Species)

binomial_species <- df_models |> 
  filter(!Species %in% non_binomial_species) |> 
  distinct(Species) |> 
  pull(Species)

#2. binomial vs. abundance species occuring only in one lake

#1: GAM without lake as random effect, binomial

bi_one_occurence <- df_models |> 
  filter(Species %in% binomial_species) |>
  group_by(Species) |>
  summarize(n_lake = n_distinct(Lake)) |> 
  filter(n_lake == 1) |>
  pull(Species)

# save df
df_bi <- df_models |>
  filter(Species %in% bi_one_occurence)

saveRDS(df_bi, "data_frame_models/df_binomial_gam")

#2 GAM abundance without lake as random effect
abu_one_occurence <- df_models |> 
  filter(Species %in% non_binomial_species) |> 
  group_by(Species) |> 
  summarize(n_lake = n_distinct(Lake)) |> 
  filter(n_lake == 1) |>
  pull(Species)

df_abu <- df_models |> 
  filter(Species %in% abu_one_occurence)

saveRDS(df_abu, "data_frame_models/df_abundance_gam")

#binomial vs. abundance species occuring in several lakes

#3 gam with random intercept, binomial
bi_multi_occurence <- df_models |> 
  filter(Species %in% binomial_species) |> 
  group_by(Species) |> 
  summarize(Lakes = n_distinct(Lake)) |> 
  filter(!Lakes == 1) |> 
  pull(Species)

df_binomial_re <- df_models |> 
  filter(Species %in% bi_multi_occurence)

saveRDS(df_binomial_re, "data_frame_models/df_binomial_re")

#4 abundance GAM with random intercept
abu_multi_occurence <- df_models |> 
  filter(Species %in% non_binomial_species) |> 
  group_by(Species) |> 
  summarize(Lakes = n_distinct(Lake)) |> 
  filter(!Lakes == 1) |> 
  pull(Species)

df_abundance_re <- df_models |> 
  filter(Species %in% abu_multi_occurence)

saveRDS(df_abundance_re, "data_frame_models/df_abundance_re")

```

#2. Data Analysis
##General Additive Models
Running the predictions for all GAMs with the smaller data frames to reduce running 
time.

```{r}
# packages relevant for data analysis
library(tidyverse)
library(here)
library(readr)
library(readxl)
library(gratia)
library(mgcv)
library(forcats)
library(ggpubr)
library(RColorBrewer)
library(broom)
library(grid)
library(DHARMa)
library(mgcViz)
library(gridExtra)
library(grid)

# function for the GAMs (old)
# source(here("function_GAM.R"))

# new function for temp and depth models 
source(here("function_temp_depth.R"))
```

###Predictions of GAMs

```{r}
# read in the dataframes
df_1 <- readRDS("data_frame_models/df_binomial_gam")
df_2 <- readRDS("data_frame_models/df_abundance_gam")
df_3 <- readRDS("data_frame_models/df_binomial_re")
df_4 <- readRDS("data_frame_models/df_abundance_re")

# use own function predictions() from function_GAM.R to predict species' models
# across all lakes

# predictions take VERY LONG! and two folders are needed to store the 
# DHARMa and prediction output inside: 
# total_models/gam_check
# total_models/predictions

#df1 and df2 are the fastest ones
# predictions() can also be used directly with df_models -> takes longer

predictions_depth_temp(df_1)

# predictions(df_2)
# predictions(df_3)
# predictions(df_4)

#  compile data frame with all predictions from the saved species' model predictions

df_predictions_all <- list.files(path = "total_models/predictions", pattern = ".rds", full.names = TRUE) |>
  map_dfr(readRDS)

# save total predictions as RDS
saveRDS(df_predictions_all, "total_models/df_pred_all.rds")

```

####Fig. 2
Overview of all model predictions

```{r, fig.show='hide'}
# Load total predictions
model_predictions <- readRDS("total_models/df_pred_all.rds")

# check the species list
model_predictions$species <- as.factor(model_predictions$species)
levels(model_predictions$species)

# Plot Figure 2. 
all_predictions <- model_predictions |>
  # exclude Lepomis gibbosus
  filter(species != "Lepomis_gibbosus") |> #lepomis gibbosus was excluded upon examination
  ggplot(aes(temp, fit)) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit), fill = factor(species)), alpha = 0.3) +
  geom_line(aes(color = factor(species))) +
  theme_bw() +
  facet_wrap(~species, scale = "free_y", labeller = labeller(species = label_wrap_gen(width = 10))) +
  theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
  scale_color_viridis(discrete = TRUE, guide = NULL, option = "G", aesthetics = c("color", "fill")) +
  xlab("Temperature") +
  ylab("Abundance")

all_predictions

# save the plot as TIFF
# tiff(paste("total_models/plots/all_predictions_models.tiff", sep = ""), units="in", width = 12, height=8, res=300)
# plot(all_predictions)
# 
# # Closing the graphical device
# dev.off()

```

##Sensitivity testing
Testing if depth-driven temperature models can adequately fit abundance data compared to
a temperature and depth model. 
Models were compared in the eleven most abundant species.

```{r}
# Read in abundance and depth data of all species included for modelling
df_models <- readRDS("df_models.rds")

# find 11 most abundant  species
species_list <- df_models |> 
  filter(tot_abu > 300) |> 
  distinct(Species) |> 
  pull(Species)

# most abundant species
 # "Alburnus_alburnus", "Coregonus_sp_albeli", "Coregonus_sp","Gasterosteus_aculeatus",
 # "Gobio_gobio", "Leuciscus_leuciscus", "Perca_fluviatilis", "Rutilus_rutilus", 
 # "Sander_lucioperca", "Scardinius_erythrophthalmus","Gymnocephalus_cernua"

# filter df into the most abundant species and select the columns of interest

df_sensitivity <- df_models |> 
  filter(tot_abu > 300) |> 
  select(Lake, Species, Protocol, Depth_sample, Abundance, Presence, mean_last_7days) |> 
  #there are some NAs in depth -> drop them
  drop_na(Depth_sample, mean_last_7days)

df_sensitivity$fLake <- as.factor(df_sensitivity$Lake)
df_sensitivity$fProtocol <- as.factor(df_sensitivity$Protocol)

# Loop through the eleven species and fit temp and temp-depth model

species_list <- df_sensitivity |> 
  distinct(Species) |> 
  pull(Species)

# loop through the species, model both GAMs, check DHARMa and predictions
for (i in species_list){

  data <- df_sensitivity |> 
    filter(Species == i)

  #Model fit for temp and depth
  model1 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(Depth_sample, k = 3)
                + s(fLake, bs = 're')
                +  s(fProtocol, bs = 're'), family = ziP())
  # Model fit for temp
  model2 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3)
                + s(fLake, bs = 're')
                +  s(fProtocol, bs = 're'), family = ziP())
  # validate model predictions for temp and depth model with dharma:
  # prepare residuals
  simulationOutput <- simulateResiduals(fittedModel = model1, plot = F)
  # tiff_filename <- paste("total_models/gam_check/temp_depth_check_", i, ".tiff", sep = "")
  # tiff(tiff_filename, width = 800, height = 600)
  print(plot(simulationOutput))
  # dev.off()
  # Plotting standardized residuals against predictors
  # tiff_file_2 <- paste("total_models/gam_check/temp_depth_predictor_", i, ".tiff", sep = "")
  # tiff(tiff_file_2, width = 800, height = 600)
  print(plotResiduals(model1, data$mean_last_7days, xlab = "temp", main=NULL))
  # dev.off()

  # for predictions with random effects in the GAMs, we select a random value
  # of the level of random effects and exclude the random effects in the prediction

  # lakes where the species is present
  unique_lakes <- distinct(data, Lake) |>
    pull()

  # select random lake for prediction
  random_lake <- sample(unique_lakes, 1)

  # for temp-depth model only predict for depth where the species is present
  presence_depth <- data |> 
    filter(Presence == 1) |>
    distinct(Depth_sample) |> 
    pull(Depth_sample)

  # take randomly 5 depth values for prediction
  random_depth <- sort(sample(presence_depth, 5))

  # Convert sorted values to factors with original order
  factor_random_depth <- factor(random_depth, levels = random_depth, ordered = TRUE)

  # grid for each of the models

  # temp-depth model predicted with random depths where the species is present
 grid1 <- expand.grid(mean_last_7days = seq(
    from = min(data$mean_last_7days, na.rm = TRUE),
    to = max(data$mean_last_7days, na.rm = TRUE), length = 200),
    fProtocol = factor("VERT"), fLake = factor(random_lake), 
    Depth_sample = factor_random_depth)

  grid2 <- expand.grid(mean_last_7days = seq(
    from = min(data$mean_last_7days, na.rm = TRUE),
    to = max(data$mean_last_7days, na.rm = TRUE), length = 1000),
    fProtocol = factor("VERT"), fLake = factor(random_lake))

  # print relevant stats
  print(paste(i))
  print(glance(model1))
  print(glance(model2))

  # manually predict both models
  model_prediction1 <- predict.gam(model1, newdata = grid1,
                                  exclude = c("s(fProtocol)", "s(fLake)"), #exclude the re
                                  term = "s(mean_last_7days)", type = "response",             
                                  se.fit = TRUE)

  model_bind1 <- cbind(grid1, as.data.frame(model_prediction1)) |>
    mutate(model = factor("temp_and_depth"))

# predict model2
  model_prediction2 <- predict.gam(model2, newdata = grid2,
                                   exclude = c("s(fProtocol)", "s(fLake)"),
                                   type = "response", se.fit = TRUE)

  model_bind2 <- cbind(grid2, as.data.frame(model_prediction2)) |>
    mutate(model = factor("temp"))

  df_pred <- bind_rows(model_bind1, model_bind2) |>
    rename(temp = mean_last_7days) |>
    select(-fProtocol, -fLake) |>
    mutate(Depth_sample = factor(ifelse(is.na(Depth_sample), "temp", as.character(Depth_sample))))

  # Extract numeric depth levels
  numeric_levels <- levels(df_pred$Depth_sample)[!levels(df_pred$Depth_sample) %in% "temp"]

  # Sort numeric depth levels
  sorted_numeric_levels <- sort(as.numeric(numeric_levels))

  # Combine sorted numeric levels with "temp" level
  sorted_levels <- c(as.character(sorted_numeric_levels), "temp")

  # Reorder levels
  df_pred$Depth_sample <- factor(df_pred$Depth_sample, levels = sorted_levels)

  # Check the levels to confirm they are ordered from lowest to highest
  levels(df_pred$Depth_sample)

  # rescale predictions of both models
  df_pred_rescaled <- df_pred |>
    group_by(model) |>
    mutate(fit_rescaled = (fit - min(fit)) / (max(fit) - min(fit))) |>
    mutate(se_fit_rescaled = (se.fit - min(se.fit)) / (max(se.fit) - min(se.fit))) |> 
    mutate(species = factor(i))

  # save predictions of both models in a rds
   saveRDS(df_pred_rescaled, paste0("total_models/sensitivity/rds/sensitivity_predictions_", i, ".rds"))

}

# prepare total predictions of all models
sensitivity_predictions <- list.files(path = "total_models/sensitivity/rds", pattern = ".rds", full.names = TRUE) |>
  map_dfr(readRDS)

# save total predictions as RDS
saveRDS(sensitivity_predictions, "total_models/sensitivity_predictions.rds")

```

###Fig. S1

```{r}
# read in the sensitivity testing data
sensitivity_predictions <- readRDS("total_models/sensitivity_predictions.rds")

levels(sensitivity_predictions$species)

# Plot both model predictions together per species
 figure_s1 <- sensitivity_predictions |>
    ggplot(aes(temp, fit_rescaled, color = model, fill = model)) +
    geom_line() +
    geom_ribbon(aes(ymin = (fit_rescaled - se.fit), ymax = (fit_rescaled + se.fit)), color = NA, alpha = 0.1) +
    theme_bw() +
    # labs(title = paste("Sensitivity Test", i)) +
    scale_color_manual(values = c("#08306B", "#D73027"), aesthetics = c("color", "fill")) +
    facet_wrap(~species, scale = "free") +
    ylab("Abundance") +
    xlab("Temperature") +
   theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        legend.position = "right",
        legend.title = element_blank())

  figure_s1
  # save as TIFF
# tiff(paste("total_models/plots/sensitivity_overview.tiff", sep = ""), units="in", width=11, height=7, res=300)
# 
# plot(figure_s1)
# 
# dev.off()

```
##Derivatives calculation

We calculated derivatives for each species per lake.

```{r}

# Derivatives() will ignore any random effect smooths it encounters in object.

# Table of pecies that were recorded with electro, fishbase or eawag, following Table 16 in Projet Lac Synthesis Report. 
# Plus the minimum and maximum temp in each lake -> needed for the new data for 
# calculating derivatives

species_lake <- read_xlsx("species_occurrences_lakes.xlsx") 
str(species_lake)

species_lake$fLake <- as.factor(species_lake$Lake)
species_lake$fProtocol <- as.factor(species_lake$Protocol)
species_lake$Species <- as.factor(species_lake$Species)

#Loading all four dfs to calculate derivatives

df_binomial_gam <- readRDS("data_frame_models/df_binomial_gam")
df_binomial_gam$fLake <- as.factor(df_binomial_gam$Lake)
df_binomial_gam$fProtocol <- as.factor(df_binomial_gam$Protocol)

df_abundance_gam <- readRDS("data_frame_models/df_abundance_gam")
df_abundance_gam$fProtocol <- as.factor(df_abundance_gam$Protocol)
df_abundance_gam$fLake <- as.factor(df_abundance_gam$Lake)

df_binomial_re <- readRDS("data_frame_models/df_binomial_re")
df_binomial_re$fLake <- as.factor(df_binomial_re$Lake)
df_binomial_re$fProtocol <- as.factor(df_binomial_re$Protocol)

df_abundance_re <- readRDS("data_frame_models/df_abundance_re")
df_abundance_re$fLake <- as.factor(df_abundance_re$Lake)
df_abundance_re$fProtocol <- as.factor(df_abundance_re$Protocol)

# Now four different loops are run to calculate derivatives of each species 
# in each lake

# Model 1: Binomial data without random intercepts

species_list <- df_binomial_gam |>
  distinct(Species) |>
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

# loop through the list of species
for (i in species_list) {
  data <- df_binomial_gam |>
    filter(Species == i)

  lake_data <- species_lake |>
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'), family = binomial)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  # loop through all lakes where the species is present, lake list is based on the information from
  # Projet Lac
  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    # using minimum and maximum temp in the lake for derivatives
     newdata <- tibble(mean_last_7days = seq(from = min(data_lake$temp, na.rm = TRUE),
                                             to = max(data_lake$temp, na.rm = TRUE), length = 200),
                       fProtocol = factor("VERT"))

     # calculate derivatives
     derivatives <- derivatives(gam_output[[i]], data = newdata) |>
       mutate(fLake = factor(j)) |>
       mutate(species = factor(i)) |>
       rename(temp = data)

     saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
  }
}

# Model 2.1: Abundance data and no random intercept for lake -> but binomial
# Coregonus_sp_benthic_profundal did not run with ZIP and was thus
# run with binomial.

species_list <- df_abundance_gam |> 
  # binomial one
  filter(Species == "Coregonus_sp_benthic_profundal") |>
  distinct(Species) |> 
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_abundance_gam |> 
    filter(Species == i)

  # binomial family instead of ZIP
  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) +
                           s(fProtocol, bs = 're'), family = binomial())

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), length = 200),
      fProtocol = factor("VERT"))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
  }
}

# Model 2.2: Abundance data and no random intercept for lake -> ZIP
# all other species in this group run with ZIP

species_list <- df_abundance_gam |>
  filter(!Species == "Coregonus_sp_benthic_profundal") |>
  distinct(Species) |> 
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_abundance_gam |> 
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'),
                         family = ziP())

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), length = 200),
      fProtocol = factor("VERT"))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
  }
}

# Model 3: Binomial data and random intercept for lake

species_list <- df_binomial_re |> 
  distinct(Species) |> 
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_binomial_re |> 
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = "re")
                         + s(fProtocol, bs = 're'), family = binomial)

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  unique_lakes <- distinct(data, Lake) |> 
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    # take all lakes where the species is present and take random one for 
    # the data included in derivatives()
    random_lake <- sample(unique_lakes, 1)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), lenqh = 200),
      fProtocol = factor("VERT"), fLake = factor(random_lake))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))

  }
}

# Model 4.1: Abundance data and random intercept for lake, simplified models
# multiple species in this group were simplified and run with binomial

species_list <- df_abundance_re |>
  # simplified models
  filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
                                "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta",
                        "Lepomis_gibbosus")) |>
  distinct(Species) |>
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_abundance_re |>
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
                         +  s(fProtocol, bs = 're'), family = binomial)

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  unique_lakes <- distinct(data, Lake) |>
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    random_lake <- sample(unique_lakes, 1)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), length = 200),
      fLake = factor(random_lake), fProtocol = factor("VERT"))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)
    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))

  }
}

# Model 4.1: Abundance data and random intercept for lake, complex models

species_list <- df_abundance_re |> 
  # exclusion of simplified models
  filter(!Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
                        "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta", 
                        "Lepomis_gibbosus")) |>
  distinct(Species) |> 
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_abundance_re |> 
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
                         +  s(fProtocol, bs = 're'), family = ziP())

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  unique_lakes <- distinct(data, Lake) |> 
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    random_lake <- sample(unique_lakes, 1)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), length = 200),
      fLake = factor(random_lake), fProtocol = factor("VERT"))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))

  }
}

# save all the derivatives in one file

df_deriv <- list.files(path = "total_models/derivatives", pattern = ".rds", full.names = TRUE) |> 
  map_dfr(readRDS)

saveRDS(df_deriv, "total_models/df_deriv_all.rds")

# save total derivatives as RDS (no exclusions)
# saveRDS(df_deriv, "total_models/df_deriv_all_no_excl.rds")

```

###Fig. S2 
Overview of derivatives in each lake community

```{r}
# Read in derivative data
all_models_derivatives <- readRDS("total_models/df_deriv_all.rds")
all_derivatives <- as_tibble(all_models_derivatives)

str(all_derivatives)

# Highlighting species with interesting derivatives: 
species_group <- all_derivatives |> 
  filter(species %in% c("Phoxinus_csikii", "Barbatula_sp_Lineage_II", "Cottus_sp_Po_profundal")) |> 
  distinct(species) |> 
  pull(species)

# Plot S2: derivatives included in each lake community
overview_derivatives <- all_derivatives |>
  arrange(fLake) |> 
  ggplot(aes(temp, derivative, color = species)) +
  geom_line() +
  facet_wrap(~fLake) +
  theme_bw(base_size = 16) +
  scale_color_viridis(discrete = TRUE, breaks = species_group, option = "G") +
  xlab("Temperature") +
  ylab("Derivatives") +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        legend.position = "right",
        legend.title = element_blank()) +
  geom_hline(yintercept = 0, color = "#7A7A7A")

overview_derivatives

# save as TIFF
# tiff(paste("total_models/plots/overview_derivatives_lake.tiff", sep = ""), units="in", width=11, height=7, res=300)
# 
# plot(overview_derivatives)
# 
# dev.off()
```

###Fig. S3

```{r}
# Data of the derivatives before simplifying multiple models
# without changes the 5 species were not run with simplified models. 
all_models_derivatives_no_excl <- readRDS("total_models/df_deriv_all_no_excl.rds")
df_deriv_no_excl <- as_tibble(all_models_derivatives_no_excl)

# Plot Figure S3a
#calculating maximum and mean derivatives for each species per lake

df_quantiles <- df_deriv_no_excl |> 
  group_by(species, fLake) |> 
  mutate(max_derivative = max(derivative)) |> 
  mutate(mean_derivative = mean(derivative)) |> 
  ungroup()

# calculate quantiles of derivatives
transform(quantile(df_quantiles$derivative,
                   c(0,0.01,0.05,0.1,0.25,0.5,0.75,0.85,0.9,0.95,0.99,1)))

# add column with the category of quantile each derivative is in
df_quantiles$groups <- cut(df_quantiles$derivative,              
                     breaks = c(-1.579943e+06, -2.342866e+00, -8.822087e-01,
                                -4.543748e-01, 3.454218e-04, 2.485142e-01,
                                5.527792e-01, 2.911987e+00, 2.018455e+01, 1.280898e+02,
                                1.231212e+07, 1.406101e+07),
                     labels = c("0-1%", "1-5%", "5-10%", "10-25%", "25-50%", "50-75%", "75-85%", "85-90%", "90-95%",
                                "95-99%", "99-100%"))

# Plot S3a: histogram of species' mean derivatives per lake
plot_percentiles <- df_quantiles |> 
  # Salmo trutta derivative is too large to be plotted
  filter(species != "Salmo_trutta") |> 
  ggplot(aes(x = mean_derivative, fill = factor(species))) + 
  geom_histogram(aes(y = after_stat(count / sum(count))), binwidth = 1.5) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Cyprinus_carpio" = "#80CDC1",
                                "Alburnus_arborella" = "#8C510A",
                                "Barbatula_sp_Lineage_I"="#DFC27D",
                               "Phoxinus_csikii" = "#01665E", "Salmo_trutta" = "#F46D43"))

plot_percentiles

# Plot S3a add lines for the percentiles
plot_percentiles_1 <- plot_percentiles +
  geom_vline(xintercept = 19.9, color = "red") +
  annotate("text", x= 18, y = 0.12, label="90th percentile", angle=90) +
  geom_vline(xintercept = 2.911987e+00, color = "red") +
  annotate("text", x= 1.5, y = 0.12, label="85th percentile", angle=90) +
  theme_classic(base_size = 13) +
  xlab("Mean species derivative per lake") +
  ylab("Percentage") +
 guides(fill = guide_legend(
    title = "Species")) +
  ylim(0, 0.2)

plot_percentiles_1

# save as TIFF
# tiff(paste("total_models/plots/plot_deriv_percentiles_1.tiff", sep = ""), units="in", width=12, height=6, res=300)
# 
# plot(plot_percentiles_1)
# 
# dev.off()

# above 90% = 2.018455e+01
df_quantiles |> 
  filter(derivative > 2.018455e+01) |> 
  distinct(species)

# 5 species above 90% -> those will be simplified to binomial 
# 1 Alburnus_arborella    
# 2 Barbatula_sp_Lineage_I
# 3 Cyprinus_carpio       
# 4 Phoxinus_csikii       
# 5 Salmo_trutta

# Plotting Figure S3b

# just keeping mean and maximum derivative values for each lake and species
data_quantiles <- df_quantiles |> 
  distinct(mean_derivative, max_derivative, fLake, species)

# Breaking mean derivatives in the same percentiles as before
data_quantiles$percentiles <- cut(data_quantiles$mean_derivative,              
                     breaks = c(-1.579943e+06, -2.342866e+00, -8.822087e-01,
                                -4.543748e-01, 3.454218e-04, 2.485142e-01,
                                5.527792e-01, 2.911987e+00, 2.018455e+01, 1.280898e+02,
                                1.231212e+07, 1.406101e+07),
                     labels = c("0-1%", "1-5%", "5-10%", "10-25%", "25-50%", "50-75%", "75-85%", "85-90%", "90-95%",
                                "95-99%", "99-100%"))

# Plot S3b tile plot of mean derivatives per species in each lake
mean_deriv_plot <- data_quantiles |> 
  ggplot(aes(fLake, y = fct_reorder(species, mean_derivative), fill= percentiles)) + 
  geom_tile() +
  scale_fill_manual(breaks = levels(data_quantiles$percentiles),
                    values = rev(brewer.pal(11, "BrBG")))

derivative_percentiles_2 <- mean_deriv_plot +
  guides(fill = guide_legend(title = "percentiles (mean derivative)", reverse = TRUE)) +
  xlab("") +
  ylab("") +
  theme_classic(base_size = 16)

derivative_percentiles_2

# save plot as TIFF
tiff(paste("total_models/plots/plot_deriv_percentiles_2.tiff", sep = ""), units="in", width=19, height=12, res=300)

plot(derivative_percentiles_2)

dev.off()

```

##Calculating thermal response diversity

```{r}
# Function provided by Ross et al. "How to measure response diversity" used to calculate
# response diversity

# read in function
source(here("functions.R"))

# load derivatives data
all_models_derivatives <- readRDS("total_models/df_deriv_all.rds")

all_deriv <- as_tibble(all_models_derivatives)

str(all_deriv)

levels(all_deriv$species)

#Calculating response diversity for each lake community

# Lepomis is excluded from the analysis
all_deriv <- all_deriv |>
  filter(!species %in% c("Lepomis_gibbosus"))

all_deriv$fLake <- as.character(all_deriv $fLake)

lakes_list <- all_deriv |> 
  distinct(fLake) |> 
  pull(fLake)

str(lakes_list)

lakes_list <- sort(lakes_list)

species_overview <- tibble()

all_deriv$species <- as.character(all_deriv$species)
all_deriv$species <- as.factor(all_deriv$species)

str(all_deriv )

# Loop to get response diversity measures for each lake 
# loop through list of lakes and calculate response diversity in the lake communities
for (i in lakes_list){

  data <- all_deriv |>
    select(temp, fLake, derivative, species) |> 
    filter(fLake == i)

  df_resp_div <- data |>
    pivot_wider(
      names_from = species,
      values_from = derivative)

  # apply the function from Ross et al. 
  # Dissimilarity
  df_resp_div$rdiv <- apply(df_resp_div[,-(1:2), drop = FALSE], 1, resp_div, sign_sens = F)
  # Divergence
  df_resp_div$sign <- apply(df_resp_div[,-(1:2), drop = FALSE], 1, resp_div, sign_sens = T)
  # Median Dissimilarity
  df_resp_div$Med <- median(df_resp_div$rdiv)

  # save response diversity metrics per lake
  saveRDS(df_resp_div, paste0("total_models/lakes_all_models/df_resp_div_", i, ".rds"))

}

# save the response diversity per lake into one file
resp_div_no_excl <- list.files(path = "total_models/lakes_all_models", pattern = ".rds", full.names = TRUE) |>
  map_dfr(readRDS) |>
  relocate(rdiv, Med, sign, .after = temp)

saveRDS(resp_div_no_excl,"total_models/resp_div_all.rds")
```

##Drivers of thermal response diversity

Preparing a df with the information on past eutrophication, species richness, and species categorization in each lake.
```{r}
# Df with properties of lakes
lake_info <- read_xlsx("lake_info.xlsx")

# select the columns we use
lake_selection <- lake_info |> 
  select(Lake, Phos_max, Max_depth, Lake_area)

# information about categorization in each lake for the subgroups
species_endemism <- read_excel("species_category_per_lake.xlsx") |> 
  rename(endemism = detail_category)

# exclude lepomis gibbosus and re-calculate species richness per lake
species_endemism_short <- species_endemism |> 
  filter(species != "Lepomis_gibbosus") |> 
  select(-sum_species) |> 
  group_by(fLake) |> 
  mutate(sum_species = sum(num_species))

species_endemism_short$endemism <- as.factor(species_endemism_short$endemism)

str(species_endemism_short)

# adding column with number of species per category in each lake
df_species_endemism <- species_endemism_short |> 
  select(-species, - num_species) |> 
  add_count(fLake, endemism) |> 
  rename(count = n) |> 
  distinct(fLake, count, sum_species, endemism)

str(df_species_endemism)

df_species_endemism$count <- as.numeric(df_species_endemism$count)

# making it wide 
df_species_endemism_long <- df_species_endemism |> 
  pivot_wider(names_from = endemism, values_from = count) |> 
  mutate_if(is.numeric, ~replace(., is.na(.), 0)) |> 
  rename(Lake = fLake)

str(df_species_endemism_long)

# merge endemism info and lake info

endemism_lake <- merge(df_species_endemism_long, lake_selection)

# Read in data on response diversity metrics in each lake
resp_div_all <- readRDS("total_models/resp_div_all.rds") |> 
  select(temp:fLake)

# select response diversity columns we need
rdiv <- resp_div_all |>  
  select(temp, fLake, rdiv, sign) |>
  drop_na()

# calculate mean and maximum values and rename Lake
df_rdiv <- rdiv |> 
  group_by(fLake) |> 
  mutate(mean_rdiv = mean(rdiv)) |> 
  mutate(max_rdiv = max(rdiv)) |> 
  mutate(mean_sign = mean(sign)) |> 
  mutate(max_sign = max(sign)) |> 
  distinct(fLake, mean_rdiv, max_rdiv, mean_sign, max_sign) |> 
  rename(Lake = fLake)

# combine lake properties, species subgroups and response diversity

df_lm <- merge(df_rdiv, endemism_lake)

str(df_lm)

# phosphorus values as numeric
df_lm$Phos_max <- as.numeric(df_lm$Phos_max)

```

###Species richness and eutrophication
Multiple linear regression to calculate the effect of species richness and
eutrophication on response diversity.

###Fig. 3

```{r}
# Multiple linear regression for mean dissimilarity explained by species richness and past eutrophication: 
# lm(mean_rdiv  ~ species richness + eutrophication)
# Predict manually with new data eutrophication value fixed at mean and species richness min() to max()
# do the same the other way around: mean_rdiv  ~  eutrophication + species richness, with species richness fixed at mean and eutrophication min() to max()species richness +
#   prepare these two plots

# Phosphorus data on Poschiavo is not available
df_lm_excl <- df_lm |> 
  filter(Lake != "Poschiavo")

# first lm for species richness ~ maximum phosphorus

lm1 <- lm(mean_rdiv ~ sum_species + Phos_max, data = df_lm_excl)

# checking model assumptions
summary(lm1)
shapiro.test(resid(lm1))
lmtest::bptest(lm1)
plot(lm1)

# model assumptions validated

# manually predict data
new_data <- tibble(sum_species = seq(from = min(df_lm_excl$sum_species), to = max(df_lm_excl$sum_species), length = 50), Phos_max = mean(df_lm_excl$Phos_max))

# predict lm
prediction_lm1 <- predict.lm(lm1, newdata = new_data, se.fit = TRUE, type = "response")

# combine new data and predictions
bind_model1 <- cbind(new_data, prediction_lm1) |> 
  mutate(category ="color")

# second lm for maximum phosphorus ~ species richness
lm2 <- lm(mean_rdiv ~ Phos_max + sum_species, data = df_lm_excl)

# checking model assumptions
summary(lm2)
shapiro.test(resid(lm2))
lmtest::bptest(lm2)
plot(lm2)

# new data
new_data2 <- tibble(Phos_max = seq(from = min(df_lm_excl$Phos_max), to = max(df_lm_excl$Phos_max),length = 50), sum_species = mean(df_lm_excl$sum_species))

# predict model
prediction_lm2 <- predict.lm(lm2, newdata = new_data2, se.fit = TRUE, type = "response")

# combine new data and predictions
bind_model2 <- cbind(new_data2, prediction_lm2) |> 
  mutate(category ="color")

# Plot Fig. 3
mycolor <- c("color" = "#A0E66A")

lm1_plot <- bind_model1 |> 
  ggplot(aes(sum_species, fit,  color = category, fill = category)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
  theme_classic(base_size = 20) +
  ylab("Mean dissimilarity") +
  xlab("Species richness") +
  ylim(1,3) +
  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)

lm2_plot <- bind_model2 |> 
  ggplot(aes(Phos_max, fit, color = category, fill = category)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
  theme_classic(base_size = 20) +
  ylab("Mean dissimilarity") +
  xlab("Maximum total phosphorus (µg/L)") +
  ylim(1,3) +
  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)

# combine both plots
figure_3 <- ggarrange(lm1_plot, lm2_plot)

# save as TIFF
# tiff(paste("total_models/plots/lm_species_richness_eutroph.tiff", sep = ""), units="in", width=12, height=5, res=300)
# 
# plot(figure_3)
# 
# # Closing the graphical device
# dev.off()

```
###Community subgroups
####Fig. 4a
We fitted simple linear models for the number of species in each category and the mean dissimilarity in each lake. 
```{r}
# Lms for all categories: endemic, translcated, non_native and non_endemic_native
# with data on all lakes
# Endemic
lm_endemic <- lm(mean_rdiv ~ endemic, data = df_lm)

# checking model assumptions
summary(lm_endemic)
shapiro.test(resid(lm_endemic))
lmtest::bptest(lm_endemic)
plot(lm_endemic)

new_data_a <- tibble(endemic = seq(from = min(df_lm$endemic), to = max(df_lm$endemic),
                                     length = 15))

prediction_endemic <- predict.lm(lm_endemic, newdata = new_data_a, se.fit = TRUE, type = "response")

df_lm_endemic <- cbind(new_data_a, prediction_endemic) |> 
  mutate(category = factor("endemic")) |> 
  rename(species_richness = endemic)

# Non-native

lm_nn <- lm(mean_rdiv ~ non_native, data = df_lm)

# checking model assumptions
summary(lm_nn)
shapiro.test(resid(lm_nn))
lmtest::bptest(lm_nn)
plot(lm_nn)

new_data_b <- tibble(non_native = seq(from = min(df_lm$non_native), to = max(df_lm$non_native),
                                   length = 15))

prediction_nn <- predict.lm(lm_nn, newdata = new_data_b, se.fit = TRUE, type = "response")

df_lm_nn <- cbind(new_data_b, prediction_nn)  |> 
  mutate(category = factor("non_native")) |> 
  rename(species_richness = non_native)

# Non-endemic native

lm_nne <- lm(mean_rdiv ~ native, data = df_lm)

# checking model assumptions
summary(lm_nne)
shapiro.test(resid(lm_nne))
lmtest::bptest(lm_nne)
plot(lm_nne)

new_data_c <- tibble(native = seq(from = min(df_lm$native), to = max(df_lm$native),
                                      length = 15))

prediction_nne <- predict.lm(lm_nne, newdata = new_data_c, se.fit = TRUE, type = "response")

df_lm_nne <- cbind(new_data_c, prediction_nne)  |> 
  mutate(category = factor("non_endemic_native")) |> 
  rename(species_richness = native)

# Translocated (non-native to region) species

lm_trans <- lm(mean_rdiv ~ non_native_region, data = df_lm)

# checking model assumptions
summary(lm_trans)
shapiro.test(resid(lm_trans))
lmtest::bptest(lm_trans)
plot(lm_trans)

new_data_d <- tibble(non_native_region = seq(from = min(df_lm$non_native_region), to = max(df_lm$non_native_region),
                                              length = 15))

prediction_trans <- predict.lm(lm_trans, newdata = new_data_d, se.fit = TRUE, type = "response")

df_lm_trans <- cbind(new_data_d, prediction_trans)  |> 
  mutate(category = factor("translocated")) |> 
  rename(species_richness = non_native_region)

# Combine all four lm predictions
df_all_lms <- rbind(df_lm_endemic, df_lm_nn, df_lm_nne, df_lm_trans)

mycolors <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
               "translocated" = "#F2AD63")

plot_all_lms <- df_all_lms|> 
  ggplot(aes(species_richness, fit, color = category, fill = category)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), color = NA, alpha = 0.3) +
  theme_classic(base_size = 16) +
  ylab("Mean dissimilarity") +
  xlab("Species richness") +
  scale_fill_manual(values = mycolors,
                    aesthetics = c("color", "fill"), 
                    labels = c("endemic", "non-native", "non-endemic native", "translocated"),
                    name = NULL)
plot_all_lms
# save as TIFF
# tiff(paste("total_models/plots/lm_all.tiff", sep = ""), units="in", width=11, height=7, res=300)
# # plot(ggarrange(depth1, depth2, ncol = 2))
# # plot science discussion
# plot(plot_all_lms)
# 
# dev.off()
```

#### Fig. 4b 
Predictions of all species shown inside the categories. 
```{r, fig.show='hide'}

# Plot species predictions inside categories across all lakes
# 5 categories;
# endemic: geographically constrained range
# non-native: from NA or Asia 
# non-endemic native: native to Switzerland but not endemic
# non-native region: native to Switzerland and surroundings but has been translocated
# to other Swiss lakes where the species was not native
# endemic translocated: endemic species that were translocated to other lakes

# read in table with the categories and rename the categories
species_category <- read_excel("species_category_across_lakes.xlsx") |> 
  select(-notes) |> 
  mutate(category = ifelse(category == "native", "non_endemic_native", category)) |> 
  mutate(category = ifelse(category == "non_native_region", "non_endemic_native_and_translocated", category)) |> 
  mutate(category2 = factor("all"))

species_category$category <- as.factor(species_category$category)
levels(species_category$category)

# merge predictions with categories
model_pred_categories <- merge(model_predictions, species_category)

# change category names for facet_wrap
category_names <- c(
  `endemic` = "endemic",
  `endemic_and_translocated` = "translocated endemic",
  `non_endemic_native` = "non-endemic native",
  `non_endemic_native_and_translocated` = "translocated non-endemic native",
  `non_native` = "non-native"
)

# colors for the plot
mycolors1 <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
                "endemic_and_translocated" = "#F2AD63", "non_endemic_native_and_translocated" = "#F2AD63")

# Rescale the predictions to 0-1 
model_pred_categories_rescaled <- model_pred_categories |> 
  group_by(species) |> 
  mutate(fit_rescaled = (fit - min(fit)) / (max(fit) - min(fit)))

# Plot the rescaled predictions
plot_pred <- model_pred_categories_rescaled |>
  # exclude Lepomis gibbosus
  filter(species != "Lepomis_gibbosus") |> 
  ggplot(aes(temp, fit_rescaled, group = species, color = category)) +
  geom_line(linewidth = 0.8) +
  theme_bw(base_size = 16) +
  ylab("Abundance") +
  xlab("Temperature") +
  scale_color_manual(values = mycolors1, guide = NULL) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        legend.position = "bottom",
        legend.title = element_blank())

plot_category_predictions <- plot_pred + facet_grid(~category, labeller = as_labeller(category_names))

plot_category_predictions

# save plot as TIFF
# tiff(paste("total_models/plots/plot_category_predictions.tiff", sep = ""), units="in", width=15, height=4, res=300)
# 
# plot(plot_category_predictions)
# 
# dev.off()

```

### 2024-07-03 function_temp_depth.R
added dharma prediction against depth

@@ -1,73 +1,4 @@
 
-# test the stats
-
-# temp and depth models
-# stats
-glance_summary <- glance(gam_output)
-tidy_summary <- tidy(gam_output)
-
-# Add species column to both summaries
-glance_summary <- glance_summary |> 
-  mutate(species = i)
-
-tidy_summary <- tidy_summary |> 
-  mutate(species = i)
-
-# combine both
-total_summary <- merge(tidy_summary, glance_summary)
-
-# Store the summary in the list
-summ <- summary(gam_output)
-
-dev_explained <- summ$dev.expl
-total_summary$deviance_explained <- dev_explained
-
-chi_sq <- summ[["chi.sq"]][["s(mean_last_7days)"]]
-
-total_summary$chi_square <- chi_sq
-
-p_value <- summ[["s.pv"]][1]
-total_summary$p_value <- p_value
-
-edf <- summ[["edf"]][1]
-total_summary$edf <- edf
-
-r_sq <- summ[["r.sq"]]
-total_summary$r_sq <- r_sq
-
-model_summary[[i]] <- total_summary
-# model_summaries_tidy[[i]] <- tidy_summary
-
-# Combine all summaries into a single data frame
-summary_df <- bind_rows(model_summary)
-
-# summary_df_tidy <- bind_rows(model_summaries_tidy)
-
-# # Export the summary data frame to a text file
-# write.table(summary_df, file = "summary_statistics.txt", sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE)
-write_xlsx(summary_df, path = "summary_statistics.xlsx")
-
-# write_xlsx(summary_df_tidy, path = "summary_statistics_tidy.xlsx")
-# summary_data <- as.data.frame(summary_df)
-# 
-# # Print the summary data frame
-# print(summary_df)
-
-df_1 <- readRDS("data_frame_models/df_binomial_gam")
-df_2 <- readRDS("data_frame_models/df_abundance_gam")
-df_3 <- readRDS("data_frame_models/df_binomial_re")
-df_4 <- readRDS("data_frame_models/df_abundance_re")
-
-new_df <- df_1 |> 
-  select(Species, tot_abu) |> 
-  distinct() |> 
-  filter(tot_abu > 20)
-
-df <- df_1 |> 
-  filter(tot_abu > 20)
-
-predictions_depth_temp(df)
-
 predictions_depth_temp <- function(df){
   require(broom)
   require(tidyverse)
[…]

### 2024-07-03 manuscript_analysis.Rmd
included multiple changes in the figures, also tried to figure out way to get derivatives of predicted model.

@@
 
+# DO NOT RUN NOW, RUN LATER!
 predictions_depth_temp(df_1)
-
-# predictions(df_2)
-# predictions(df_3)
-# predictions(df_4)
+predictions_depth_temp(df_2)
+predictions_depth_temp(df_3)
+# 4 not run yet, will wait with this
+predictions_depth_temp(df_4)
+
+# statistics of the gamms
+total_models <- bind_rows(df_1, df_2, df_3, df_4)
+
+# total_models |> 
+#   filter(tot_abu > 20) |> 
+#   distinct(Species, tot_abu)
+stats_gamms(total_models)
 
@@
 
-# Plot Figure 2. 
-all_predictions <- model_predictions |>
-  # exclude Lepomis gibbosus
-  filter(species != "Lepomis_gibbosus") |> #lepomis gibbosus was excluded upon examination
+# read in the category table and rename categories
+species_across_lakes <-  read_xlsx("species_category_across_lakes.xlsx") |> 
+  select(-notes) |> 
+    mutate(category = ifelse(category == "native", "non_endemic_native", category)) |>
+  mutate(category = ifelse(category == "non_native_region", "non_endemic_native_and_translocated", category)) |> 
+  mutate(category2 = factor("all"))
+
+species_across_lakes$category <- as.factor(species_across_lakes$category)
+levels(species_across_lakes$category)
+
+#merge predictions with categories
+predictions_categories <- merge(model_predictions, species_across_lakes)
+
+# EVTL NEW PLOT 2
+# change category names for facet_wrap
+category_names <- c(
+  `endemic` = "endemic",
+  `endemic_and_translocated` = "translocated endemic",
+  `non_endemic_native` = "non-endemic native",
+  `non_endemic_native_and_translocated` = "translocated non-endemic native",
+  `non_native` = "non-native"
+)
+
+# colors for the plot
+mycolors1 <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
+                "endemic_and_translocated" = "#F2AD63", "non_endemic_native_and_translocated" = "#F2AD63")
+
+# Rescale the predictions to 0-1 
+predictions_categories_rescaled <- predictions_categories |>
+  group_by(species) |>
+  mutate(fit_rescaled = (fit - min(fit)) / (max(fit) - min(fit)))
+
+plot_pred <- predictions_categories_rescaled |>
+  ggplot(aes(temp, fit_rescaled, group = species, color = category)) +
+  geom_line() +
+  # how can I get the right color?
+   # geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), alpha = 0.3) +
+  theme_bw(base_size = 16) +
+  ylab("Abundance") +
+  xlab("Temperature") +
+  scale_color_manual(values = mycolors1, guide = NULL) +
+  theme(panel.grid.major = element_blank(), 
+        panel.grid.minor = element_blank(), 
+        legend.position = "bottom",
+        legend.title = element_blank())
+
+plot_category_predictions <- plot_pred + facet_grid(~category, labeller = as_labeller(category_names))
+
+plot_category_predictions
+
+# trying to group by category and color
+predictions_categories |>
+  ggplot(aes(temp, fit)) +
+  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit), fill = factor(category)), alpha = 0.3) +
+  geom_line(aes(color = factor(category))) +
+  theme_bw() +
+  facet_wrap(~species + category, scale = "free_y", labeller = labeller(species = label_wrap_gen(width = 10))) +
+  theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
+  scale_color_manual(values = mycolors1, aesthetics = c("color", "fill")) +
+  xlab("Temperature") +
+  ylab("Abundance")
+
+# OLD FIGURE 2
+
+all_predictions <- predictions_categories |>
   ggplot(aes(temp, fit)) +
@@
 all_predictions
-
 # save the plot as TIFF
@@
 ##Sensitivity testing
+NOT CHANGED 
 Testing if depth-driven temperature models can adequately fit abundance data compared to
@@
 
+# separating the dfs to separate model types
+depth_binomial_gam <- df_binomial_gam |> 
+  filter(tot_abu > 20)
+
+temp_binomial_gam <- df_binomial_gam |> 
+  filter(tot_abu < 20)
+
+depth_abundance_gam <- df_abundance_gam |> 
+  filter(tot_abu > 20)
+
+temp_abundance_gam <- df_abundance_gam |> 
+  filter(tot_abu < 20)
+
+depth_binomial_re <- df_binomial_re |> 
+  filter(tot_abu > 20)
+
+temp_binomial_re <- df_binomial_re |> 
+  filter(tot_abu < 20)
+
+depth_abundance_re <- df_abundance_re |> 
+  filter(tot_abu > 20)
+
+temp_abundance_re <- df_abundance_re |> 
+  filter(tot_abu < 20)
+
+```
+
+###Depth and temp models
+```{r}
+# DEPTH AND TEMPERATURE
+# UNCLEAR HOW THIS CAN BE DONE
+# Model 1: Binomial data without random intercepts
+
+species_list <- depth_binomial_gam |>
+  distinct(Species) |>
+  pull(Species)
+
+species_list <- sort(species_list)
+
+derivatives <- list()
+gam_output <- list()
+model_prediction <- list()
+
+i <- "Salmo_sp_Blackspot"
+j <- "Poschiavo"
+# loop through the list of species
+for (i in species_list) {
+  data <- depth_binomial_gam |>
+    filter(Species == i)
+
+  lake_data <- species_lake |>
+    filter(Species == i)
+
+  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
+                           s(Depth_sample, k = 3) + s(fProtocol, bs = 're'),
+                         family = binomial)
+
+  lake_list <- distinct(lake_data, Lake) |>
+    pull()
+
+  # loop through all lakes where the species is present, lake list is based on the information from
+  # Projet Lac
+  for (j in lake_list){
+
+    data_lake <- species_lake |>
+      filter(Species == i) |>
+      filter(Lake == j)
+
+    # using minimum and maximum temp in the lake for derivatives
+     newdata <- tibble(mean_last_7days = seq(from = min(data_lake$temp, na.rm = TRUE),
+                                             to = max(data_lake$temp, na.rm = TRUE)
+                                             ,length = 200), Depth_sample =
+                         mean(data$Depth_sample, na.rm = TRUE),
+                       fProtocol = factor("VERT"))
+
+     # calculate derivatives
+     # derivatives <- derivatives(gam_output[[i]], data = newdata) |>
+     #   mutate(fLake = factor(j)) |>
+     #   mutate(species = factor(i))
+     #   rename(temp = data)
+     #
+
+       test_derivatives <- derivative_samples(gam_output[[i]]
+                                              , focal = "mean_last_7days", data = newdata)
+
+       test_derivatives |>
+         filter(.draw <= 200) |>
+         ggplot(aes(x = mean_last_7days, y = .derivative, group = .draw)) +
+         geom_line(alpha = 0.5)
+
+       derivatives <- test_derivatives |> 
+         filter(.draw == 500) |> 
+         mutate(fLake = factor(j)) |>
+         mutate(species = factor(i))
+
+       derivatives |> 
+         ggplot(aes(mean_last_7days, .derivative)) +
+         geom_line()
+
+     saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
+  }
+}
+
+# needs to be discussed
+
+# nothing changed starting from here
+############################################################################3##3
+# Model 2.1: Abundance data and no random intercept for lake -> but binomial
+# Coregonus_sp_benthic_profundal did not run with ZIP and was thus
+# run with binomial.
+
+species_list <- depth_abundance_gam |> 
+  # binomial one
+  filter(Species == "Coregonus_sp_benthic_profundal") |>
+  distinct(Species) |> 
+  pull(Species)
+
+species_list <- sort(species_list)
+
+derivatives <- list()
+gam_output <- list()
+model_prediction <- list()
+
+for (i in species_list) {
+
+  data <- df_abundance_gam |> 
+    filter(Species == i)
+
+  # binomial family instead of ZIP
+  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) +
+                           s(fProtocol, bs = 're'), family = binomial())
+
+  lake_data <- species_lake |>
+    filter(Species == i)
+
+  lake_list <- distinct(lake_data, Lake) |>
+    pull()
+
+  for (j in lake_list){
+
+    data_lake <- species_lake |>
+      filter(Species == i) |>
+      filter(Lake == j)
+
+    newdata <- tibble(mean_last_7days = seq(
+      from = min(data_lake$temp, na.rm = TRUE),
+      to = max(data_lake$temp, na.rm = TRUE), length = 200),
+      fProtocol = factor("VERT"))
+
+    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
+      mutate(fLake = factor(j)) |>
+      mutate(species = factor(i)) |>
+      rename(temp = data)
+
+    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
+  }
+}
+
+# Model 2.2: Abundance data and no random intercept for lake -> ZIP
+# all other species in this group run with ZIP
+
+species_list <- df_abundance_gam |>
+  filter(!Species == "Coregonus_sp_benthic_profundal") |>
+  distinct(Species) |> 
+  pull(Species)
+
+species_list <- sort(species_list)
+
+derivatives <- list()
+gam_output <- list()
+model_prediction <- list()
+
+for (i in species_list) {
+
+  data <- df_abundance_gam |> 
+    filter(Species == i)
+
+  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'),
+                         family = ziP())
+
+  lake_data <- species_lake |>
+    filter(Species == i)
+
+  lake_list <- distinct(lake_data, Lake) |>
+    pull()
+
+  for (j in lake_list){
+
+    data_lake <- species_lake |>
+      filter(Species == i) |>
+      filter(Lake == j)
+
+    newdata <- tibble(mean_last_7days = seq(
+      from = min(data_lake$temp, na.rm = TRUE),
+      to = max(data_lake$temp, na.rm = TRUE), length = 200),
+      fProtocol = factor("VERT"))
+
+    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
+      mutate(fLake = factor(j)) |>
+      mutate(species = factor(i)) |>
+      rename(temp = data)
+
+    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
+  }
+}
+
+# Model 3: Binomial data and random intercept for lake
+
+species_list <- df_binomial_re |> 
+  distinct(Species) |> 
+  pull(Species)
+
+species_list <- sort(species_list)
+
+derivatives <- list()
+gam_output <- list()
+model_prediction <- list()
+
+for (i in species_list) {
+
+  data <- df_binomial_re |> 
+    filter(Species == i)
+
+  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = "re")
+                         + s(fProtocol, bs = 're'), family = binomial)
+
+  lake_data <- species_lake |>
+    filter(Species == i)
+
+  lake_list <- distinct(lake_data, Lake) |>
+    pull()
+
+  unique_lakes <- distinct(data, Lake) |> 
+    pull()
+
+  for (j in lake_list){
+
+    data_lake <- species_lake |>
+      filter(Species == i) |>
+      filter(Lake == j)
+
+    # take all lakes where the species is present and take random one for 
+    # the data included in derivatives()
+    random_lake <- sample(unique_lakes, 1)
+
+    newdata <- tibble(mean_last_7days = seq(
+      from = min(data_lake$temp, na.rm = TRUE),
+      to = max(data_lake$temp, na.rm = TRUE), lenqh = 200),
+      fProtocol = factor("VERT"), fLake = factor(random_lake))
+
+    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
+      mutate(fLake = factor(j)) |>
+      mutate(species = factor(i)) |>
+      rename(temp = data)
+
+    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
+
+  }
+}
+
+# Model 4.1: Abundance data and random intercept for lake, simplified models
+# multiple species in this group were simplified and run with binomial
+
+species_list <- df_abundance_re |>
+  # simplified models
+  filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
+                                "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta",
+                        "Lepomis_gibbosus")) |>
+  distinct(Species) |>
+  pull(Species)
+
+species_list <- sort(species_list)
+
+derivatives <- list()
+gam_output <- list()
+model_prediction <- list()
+
+for (i in species_list) {
+
+  data <- df_abundance_re |>
+    filter(Species == i)
+
+  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
+                         +  s(fProtocol, bs = 're'), family = binomial)
+
+  lake_data <- species_lake |>
+    filter(Species == i)
+
+  lake_list <- distinct(lake_data, Lake) |>
+    pull()
+
+  unique_lakes <- distinct(data, Lake) |>
+    pull()
+
+  for (j in lake_list){
+
+    data_lake <- species_lake |>
+      filter(Species == i) |>
+      filter(Lake == j)
+
+    random_lake <- sample(unique_lakes, 1)
+
+    newdata <- tibble(mean_last_7days = seq(
+      from = min(data_lake$temp, na.rm = TRUE),
+      to = max(data_lake$temp, na.rm = TRUE), length = 200),
+      fLake = factor(random_lake), fProtocol = factor("VERT"))
+
+    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
+      mutate(fLake = factor(j)) |>
+      mutate(species = factor(i)) |>
+      rename(temp = data)
+    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
+
+  }
+}
+
+# Model 4.1: Abundance data and random intercept for lake, complex models
+
+species_list <- df_abundance_re |> 
+  # exclusion of simplified models
+  filter(!Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
+                        "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta", 
+                        "Lepomis_gibbosus")) |>
+  distinct(Species) |> 
+  pull(Species)
+
+species_list <- sort(species_list)
+
+derivatives <- list()
+gam_output <- list()
+model_prediction <- list()
+
+for (i in species_list) {
+
+  data <- df_abundance_re |> 
+    filter(Species == i)
+
+  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
+                         +  s(fProtocol, bs = 're'), family = ziP())
+
+  lake_data <- species_lake |>
+    filter(Species == i)
+
+  lake_list <- distinct(lake_data, Lake) |>
+    pull()
+
+  unique_lakes <- distinct(data, Lake) |> 
+    pull()
+
+  for (j in lake_list){
+
+    data_lake <- species_lake |>
+      filter(Species == i) |>
+      filter(Lake == j)
+
+    random_lake <- sample(unique_lakes, 1)
+
+    newdata <- tibble(mean_last_7days = seq(
+      from = min(data_lake$temp, na.rm = TRUE),
+      to = max(data_lake$temp, na.rm = TRUE), length = 200),
+      fLake = factor(random_lake), fProtocol = factor("VERT"))
+
+    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
+      mutate(fLake = factor(j)) |>
+      mutate(species = factor(i)) |>
+      rename(temp = data)
+
+    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
+
+  }
+}
+
+# save all the derivatives in one file
+
+df_deriv <- list.files(path = "total_models/derivatives", pattern = ".rds", full.names = TRUE) |> 
+  map_dfr(readRDS)
+
+saveRDS(df_deriv, "total_models/df_deriv_all.rds")
+
+# save total derivatives as RDS (no exclusions)
+# saveRDS(df_deriv, "total_models/df_deriv_all_no_excl.rds")
+
+```
+
+###Temp models
+```{r}
+
 # Now four different loops are run to calculate derivatives of each species 
@@
 # Plot Fig. 3
-mycolor <- c("color" = "#A0E66A")
+"#78BE63"
+mycolor <- c("color" = "#78BE63")
+# mycolor <- c("color" = "#A0E66A")
+"98D47C"
 
@@
 
+figure_3
 # save as TIFF
@@
 # plot(figure_3)
+# 
+# # Closing the graphical device
+# dev.off()
+
+```
+
+###Divergence figure 3
+
+```{r}
+# first lm for species richness ~ maximum phosphorus
+
+lm3 <- lm(mean_sign ~ sum_species + Phos_max, data = df_lm_excl)
+
+# checking model assumptions
+summary(lm3)
+shapiro.test(resid(lm3))
+lmtest::bptest(lm3)
+plot(lm3)
+
+# model assumptions validated
+
+# manually predict data
+new_data3 <- tibble(sum_species = seq(from = min(df_lm_excl$sum_species), to = max(df_lm_excl$sum_species), length = 50), Phos_max = mean(df_lm_excl$Phos_max))
+
+# predict lm
+prediction_lm3 <- predict.lm(lm3, newdata = new_data3, se.fit = TRUE, type = "response")
+
+# combine new data and predictions
+bind_model3 <- cbind(new_data3, prediction_lm3) |> 
+  mutate(category ="color")
+
+# second lm for maximum phosphorus ~ species richness
+lm4 <- lm(mean_sign ~ Phos_max + sum_species, data = df_lm_excl)
+
+# checking model assumptions
+summary(lm4)
+shapiro.test(resid(lm4))
+lmtest::bptest(lm4)
+plot(lm4)
+
+# new data
+new_data4 <- tibble(Phos_max = seq(from = min(df_lm_excl$Phos_max), to = max(df_lm_excl$Phos_max),length = 50), sum_species = mean(df_lm_excl$sum_species))
+
+# predict model
+prediction_lm4 <- predict.lm(lm4, newdata = new_data4, se.fit = TRUE, type = "response")
+
+# combine new data and predictions
+bind_model4 <- cbind(new_data4, prediction_lm4) |> 
+  mutate(category ="color")
+
+# Plot Fig. 3
+mycolor <- c("color" = "#88D44F")
+# mycolor <- c("color" = "#A0E66A")
+"98D47C"
+
+lm3_plot <- bind_model3 |> 
+  ggplot(aes(sum_species, fit,  color = category, fill = category)) +
+  geom_line(linewidth = 1.2) +
+  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
+  theme_classic(base_size = 20) +
+  ylab("Mean divergence") +
+  xlab("Species richness") +
+  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
+
+lm4_plot <- bind_model4 |> 
+  ggplot(aes(Phos_max, fit, color = category, fill = category)) +
+  geom_line(linewidth = 1.2) +
+  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
+  theme_classic(base_size = 20) +
+  ylab("Mean divergence") +
+  xlab("Maximum total phosphorus (µg/L)") +
+  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
+
+# combine both plots
+figure_divergence <- ggarrange(lm3_plot, lm4_plot)
+
+figure_divergence
+# save as TIFF
+# tiff(paste("total_models/plots/lm_divergence_multiple_linear.tiff", sep = ""), units="in", width=12, height=5, res=300)
+# 
+# plot(figure_divergence)
 # 
@@
 plot_all_lms <- df_all_lms|> 
-  ggplot(aes(species_richness, fit, color = category, fill = category)) +
+  ggplot(aes(species_richness, fit, color = category, fill = category, linetype = category)) +  
   geom_line(linewidth = 1) +
@@
                     aesthetics = c("color", "fill"),
-                    labels = c("endemic", "non-native", "non-endemic native", "translocated"),
-                    name = NULL)
+                    labels = c("endemic", "non-native (ns)", "non-endemic native", "translocated (ns)")) +
+   scale_linetype_manual(values = c("endemic" = "solid",
+                                   "non_native" = "dashed",
+                                   "non_endemic_native" = "solid",
+                                   "translocated" = "dashed"), labels = c("endemic", "non-native (ns)", "non-endemic native", "translocated (ns)"))
+
 plot_all_lms
@@
 
+####Divergence Fig. 4a
+
+```{r}
+# Lms for all categories: endemic, translcated, non_native and non_endemic_native
+# with data on all lakes
+# Endemic
+deriv_lm1 <- lm(mean_sign ~ endemic, data = df_lm)
+
+# checking model assumptions
+summary(deriv_lm1)
+shapiro.test(resid(deriv_lm1))
+lmtest::bptest(deriv_lm1)
+plot(deriv_lm1)
+
+new_data_1 <- tibble(endemic = seq(from = min(df_lm$endemic), to = max(df_lm$endemic),
+                                     length = 15))
+
+prediction_endemic_1 <- predict.lm(deriv_lm1, newdata = new_data_1, se.fit = TRUE, type = "response")
+
+df_deriv_endemic <- cbind(new_data_1, prediction_endemic_1) |> 
+  mutate(category = factor("endemic")) |> 
+  rename(species_richness = endemic)
+
+# Non-native
+
+deriv_lm2 <- lm(mean_sign ~ non_native, data = df_lm)
+
+# checking model assumptions
+summary(deriv_lm2)
+shapiro.test(resid(deriv_lm2))
+lmtest::bptest(deriv_lm2)
+plot(deriv_lm2)
+
+new_data_2 <- tibble(non_native = seq(from = min(df_lm$non_native), to = max(df_lm$non_native),
+                                   length = 15))
+
+prediction_2 <- predict.lm(deriv_lm2, newdata = new_data_2, se.fit = TRUE, type = "response")
+
+df_deriv_nn <- cbind(new_data_2, prediction_2)  |> 
+  mutate(category = factor("non_native")) |> 
+  rename(species_richness = non_native)
+
+# Non-endemic native
+
+deriv_lm3 <- lm(mean_sign ~ native, data = df_lm)
+
+# checking model assumptions
+summary(deriv_lm3)
+shapiro.test(resid(deriv_lm3))
+lmtest::bptest(deriv_lm3)
+plot(deriv_lm3)
+
+new_data_3 <- tibble(native = seq(from = min(df_lm$native), to = max(df_lm$native),
+                                      length = 15))
+
+prediction_3 <- predict.lm(deriv_lm3, newdata = new_data_3, se.fit = TRUE, type = "response")
+
+df_deriv_nne <- cbind(new_data_3, prediction_3)  |> 
+  mutate(category = factor("non_endemic_native")) |> 
+  rename(species_richness = native)
+
+# Translocated (non-native to region) species
+
+deriv_lm4 <- lm(mean_sign ~ non_native_region, data = df_lm)
+
+# checking model assumptions
+summary(deriv_lm4)
+shapiro.test(resid(deriv_lm4))
+lmtest::bptest(deriv_lm4)
+plot(deriv_lm4)
+
+new_data_4 <- tibble(non_native_region = seq(from = min(df_lm$non_native_region), to = max(df_lm$non_native_region),
+                                              length = 15))
+
+prediction_4 <- predict.lm(deriv_lm4, newdata = new_data_4, se.fit = TRUE, type = "response")
+
+df_deriv_trans <- cbind(new_data_4, prediction_4)  |> 
+  mutate(category = factor("translocated")) |> 
+  rename(species_richness = non_native_region)
+
+# Combine all four lm predictions
+df_deriv_lms <- rbind(df_deriv_endemic, df_deriv_nn, df_deriv_nne, df_deriv_trans)
+
+mycolors <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
+               "translocated" = "#F2AD63")
+
+lm_divergence_all <- df_deriv_lms|> 
+  ggplot(aes(species_richness, fit, color = category, fill = category)) +
+  geom_line(linewidth = 1, linetype = "dashed") +
+  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), color = NA, alpha = 0.3) +
+  theme_classic(base_size = 16) +
+  ylab("Mean divergence") +
+  xlab("Species richness") +
+  scale_fill_manual(values = mycolors,
+                    aesthetics = c("color", "fill"), 
+                    labels = c("endemic", "non-native", "non-endemic native", "translocated"),
+                    name = NULL)
+
+lm_divergence_all
+
+# save as TIFF
+# tiff(paste("total_models/plots/lm_divergence_all.tiff", sep = ""), units="in", width=11, height=7, res=300)
+# # plot(ggarrange(depth1, depth2, ncol = 2))
+# # plot science discussion
+# plot(lm_divergence_all)
+# 
+# dev.off()
+```
 #### Fig. 4b 

### 2024-07-03 function_temp_depth.R
stats are now separate, typo solved -> models should work.

@@ -56,7 +56,7 @@ predictions_depth_temp <- function(df){
           
           print(paste(data$tot_abu[1], i))
           
-          gam_output <- gam(data = df, Presence ~ s(mean_last_7days, k = 3) + s(Depth_sample, k = 3)
+          gam_output <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) + s(Depth_sample, k = 3)
                             +  s(fProtocol, bs = 're'), family = binomial)
           
           # prepare residuals
@@ -73,7 +73,7 @@ predictions_depth_temp <- function(df){
           dev.off()
           
           # Plotting standardized residuals against predictors: depth
-          tiff_file_3 <- paste("total_models/gam_check/depth_predictor_depth", i, ".tiff", sep = "")
+          tiff_file_3 <- paste("total_models/gam_check/depth_predictor_", i, ".tiff", sep = "")
           tiff(tiff_file_3, width = 800, height = 600)
           print(plotResiduals(simulationOutput, depth_data$Depth_sample, xlab = "depth", main=NULL))
           dev.off()
@@ -89,47 +89,47 @@ predictions_depth_temp <- function(df){
           # saving the predictions
           saveRDS(pred_df, paste0("total_models/predictions/predictions_",i,".rds"))
           
-          # stats
-          glance_summary <- glance(gam_output)
-          tidy_summary <- tidy(gam_output)
-          
-          # Add species column to both summaries
-          glance_summary <- glance_summary |> 
-            mutate(species = i)
-          
-          tidy_summary <- tidy_summary |> 
-            mutate(species = i)
-          
-          # combine both
-          total_summary <- merge(tidy_summary, glance_summary)
-          
-          # Store the summary in the list
-          summ <- summary(gam_output)
-          
-          dev_explained <- summ$dev.expl
-          total_summary$deviance_explained <- dev_explained
-          
-          chi_sq <- summ[["chi.sq"]][["s(mean_last_7days)"]]
-          
-          total_summary$chi_square <- chi_sq
-          
-          p_value <- summ[["s.pv"]][1]
-          total_summary$p_value <- p_value
-          
-          edf <- summ[["edf"]][1]
-          total_summary$edf <- edf
-          
-          r_sq <- summ[["r.sq"]]
-          total_summary$r_sq <- r_sq
-          
-          model_summary[[i]] <- total_summary
+          # # stats
+          # glance_summary <- glance(gam_output)
[…]

### 2024-07-03 manuscript_analysis.Rmd
rerunning for typo models from 1-3

@@
 # predictions() can also be used directly with df_models -> takes longer
-
+# BECAUSE OF TYPO: RE RUN TEMP + DEPTH models from df1-3, df4 still needs to be run & stats overview too
+
+df_1to3_dept_temp <- bind_rows(df_1, df_2, df_3) |> 
+  filter(tot_abu > 20)
+
+df_1to3_dept_temp |> 
+  distinct(Species)
+
+predictions_depth_temp(df_1to3_dept_temp)
 # DO NOT RUN NOW, RUN LATER!
@@
 
+stats_gamms(df_test)
+
 # statistics of the gamms
@@
 
-# total_models |> 
-#   filter(tot_abu > 20) |> 
-#   distinct(Species, tot_abu)
 stats_gamms(total_models)

### 2024-07-03 function_temp_depth.R
models are running smoothly. stats will be tested

@@ -286,8 +286,9 @@ predictions_depth_temp <- function(df){
           fProtocol = factor("VERT"), fLake = factor(random_lake))
         
         # special cases
-        if (i %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
-                     "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta")){
+        if (i %in% c("Alosa_agone", "Alburnus_arborella", "Barbatula_sp_Lineage_I",
+                     "Coregonus_sp_large_pelagic", 
+                     "Cyprinus_carpio", "Lepomis_gibbosus", "Phoxinus_csikii", "Salmo_trutta")){
           
           print(paste(data$tot_abu[1], i))
           
@@ -736,8 +737,9 @@ predictions_depth_temp <- function(df){
           fProtocol = factor("VERT"), fLake = factor(random_lake))
         
         # special cases
-        if (i %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
-                     "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta")){
+        if (i %in% c("Alosa_agone", "Alburnus_arborella", "Barbatula_sp_Lineage_I",
+                     "Coregonus_sp_large_pelagic", 
+                     "Cyprinus_carpio", "Lepomis_gibbosus", "Phoxinus_csikii", "Salmo_trutta")){
           print(i)
           
           gam_output<- gam(data = data, Presence ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
@@ -1151,8 +1153,9 @@ stats_gamms <- function(df){
         
       
         # special cases
-        if (i %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
-                     "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta")){
+        if (i %in% c("Alosa_agone", "Alburnus_arborella", "Barbatula_sp_Lineage_I",
+                     "Coregonus_sp_large_pelagic", 
+                     "Cyprinus_carpio", "Lepomis_gibbosus", "Phoxinus_csikii", "Salmo_trutta")){
           
           print(paste(data$tot_abu[1], i))
           
@@ -1473,8 +1476,9 @@ stats_gamms <- function(df){
         # the prediction anyway
       
         # special cases
-        if (i %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
[…]

### 2024-07-03 manuscript_analysis.Rmd
solved problems with some of the models, checked AICs and simplified them.

@@
 
-df_1to3_dept_temp <- bind_rows(df_1, df_2, df_3) |> 
-  filter(tot_abu > 20)
-
-df_1to3_dept_temp |> 
-  distinct(Species)
-
-predictions_depth_temp(df_1to3_dept_temp)
 # DO NOT RUN NOW, RUN LATER!
@@
 predictions_depth_temp(df_3)
-# 4 not run yet, will wait with this
 predictions_depth_temp(df_4)
-
-stats_gamms(df_test)
 
@@
 
+###AIC comparisons in special cases
+
+```{r}
+
+AIC_testing <- function(data2, i){
+
+  data <- data2 |> 
+    filter(Species == i)
+
+  data$fProtocol <- as.factor(data$Protocol)
+  data$fLake <- as.factor(data$Lake)
+
+  model_0 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
+                   s(Depth_sample, k = 3)
+                            + s(fLake, bs = 're')
+                            +  s(fProtocol, bs = 're'), family = ziP())
+  print(glance(model_0))
+# new models
+
+model1 <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) + s(Depth_sample, k = 3)
+                            + s(fLake, bs = 're')
+                            +  s(fProtocol, bs = 're'), family = binomial)
+print(glance(model1))
+
+model2 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
+                            + s(fLake, bs = 're')
+                            +  s(fProtocol, bs = 're'), family = ziP())
+print(glance(model2))
+
+}
+
+AIC_testing(df_4, "Lepomis_gibbosus")
+
+# in Alosa agone: model_0 not running, smaller AIC for binomial depth + temp model -> will be modelled as binomial depth + temp
+# same for Coregonus_sp_large_pelagic
+# lepomis gibbosus also has the smallest AIC for binomial depth + temp model
+
+# also testing the other special cases:
+ # "Alburnus_arborella", "Barbatula_sp_Lineage_I","Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta"
+
+AIC_testing(df_4, "Alburnus_arborella") 
+# not running with model0, Binomial with depth and temp best
+AIC_testing(df_4, "Barbatula_sp_Lineage_I")
+# binomial depth and temp has lowest AIC
+AIC_testing(df_4, "Cyprinus_carpio")
+# binomial depth and temp has lowest AIC 
+AIC_testing(df_4, "Phoxinus_csikii")
+#binomial depth and temp has lowest AIC 
+AIC_testing(df_4, "Salmo_trutta")
+# binomial depth and temp lowest AIC
+
+# 8 models are run with binomial temp and depth
+
+```
 ####Fig. 2

### 2024-07-03 function_temp_depth.R
same problem also in the stadt function, not running smoothly. Cottu sp Po has an error

@@ -952,6 +952,12 @@ predictions_depth_temp <- function(df){
   
 }
 
+
+subs <- total_models |> 
+  filter(Species == "Cottus_sp_Po")
+df <- subs
+
+i <- "Cottus_sp_Po"
 stats_gamms <- function(df){
   require(broom)
   require(tidyverse)
@@ -987,10 +993,7 @@ stats_gamms <- function(df){
     data <- df |> 
       filter(Species == i) |> 
       mutate(n_lake = n_distinct(Lake))
-    
-    unique_lakes <- unique(data$fLake)
-    
-    random_lake <- sample(levels(unique_lakes), 1)
+  
     
     # temp depth models
     if(max(data$tot_abu) > 20){
@@ -1001,7 +1004,7 @@ stats_gamms <- function(df){
           
           print(paste(data$tot_abu[1], i))
           
-          gam_output <- gam(data = df, Presence ~ s(mean_last_7days, k = 3) + s(Depth_sample, k = 3)
+          gam_output <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) + s(Depth_sample, k = 3)
                             +  s(fProtocol, bs = 're'), family = binomial)
           
           # stats
@@ -1051,7 +1054,7 @@ stats_gamms <- function(df){
         else if (max(data$Abundance) > 1)  {
           print(paste(data$tot_abu[1], i))
           
-          gam_output <- gam(data = df, Abundance ~ s(mean_last_7days, k = 3) + s(Depth_sample, k = 3)
+          gam_output <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(Depth_sample, k = 3)
                             +  s(fProtocol, bs = 're'), family = ziP())
           
           # stats
@@ -1101,7 +1104,7 @@ stats_gamms <- function(df){
         else{
           print(paste(data$tot_abu[1], i))
           
-          gam_output <- gam(data = df, Abundance ~ s(mean_last_7days, k = 3) + s(Depth_sample, k = 3)
+          gam_output <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(Depth_sample, k = 3)
                             +  s(fProtocol, bs = 're'), family = binomial)
           
           # stats
@@ -1159,7 +1162,7 @@ stats_gamms <- function(df){
           
           print(paste(data$tot_abu[1], i))
           
-          gam_output <- gam(data = df, Presence ~ s(mean_last_7days, k = 3) + s(Depth_sample, k = 3)
[…]

### 2024-07-03 manuscript_analysis.Rmd
stats function has been run

@@
 
-#df1 and df2 are the fastest ones
-# predictions() can also be used directly with df_models -> takes longer
-# BECAUSE OF TYPO: RE RUN TEMP + DEPTH models from df1-3, df4 still needs to be run & stats overview too
-
 # DO NOT RUN NOW, RUN LATER!
@@
 
+# issue with cottus sp po
 # statistics of the gamms
@@
   print(glance(model_0))
-# new models
 
@@
 # binomial depth and temp lowest AIC
+
+# AIC_testing(df_4, "Cottus_sp_Po")
+# # best model is the temp, depth model
 
@@
 # trying to group by category and color
-predictions_categories |>
+figure_test_2 <- predictions_categories |>
   ggplot(aes(temp, fit)) +
@@
   ylab("Abundance")
+
+# tiff(paste("total_models/plots/draft_all_predictions.tiff", sep = ""), units="in", width = 12, height=8, res=300)
+# plot(figure_test_2)
+# 
+# # Closing the graphical device
+# dev.off()
 

### 2024-07-05 models_derivatives_response_diversity.Rmd
separate rmd for models, derivatives and response diversity calculattion

--- 2024-07-03 manuscript_analysis.Rmd
@@
 ---
-title: "manuscript_analysis"
+title: "Models, derivatives, response diversity"
 author: "Sophie Moreau"
-date: "2024-07-03"
+date: "2024-07-05"
 output: html_document
@@
 ---
-
-```{r}
-library(tidyverse)
-library(here)
-library(readr)
-library(viridis)
-
-```
-
-#Introduction
-This is the supplementary information of the Master thesis. This document
-covers the whole analysis performed to assess response diversity between fish communities
-in Swiss peri-alpine lakes. 
-This document has two parts: Data preparation (1) and Data analysis (2). 
-Function to model general additive models is provided separately (see function_GAM.R). 
-Code for figures used in the method figure is also provided separately (see supplementary_method_figure.R)
-
-```{r}
-# do we want to include zurich and constance separately?
-projet_lac_data <- read.csv("PLDB_final_short_depth_climate_cleantaxa_20052022.csv")
-
-str(projet_lac_data)
-projet_lac_data$LakeBasin <- as.factor(projet_lac_data$LakeBasin)
-
-projet_lac_data |> 
-  distinct(LakeBasin)
-```
-
-#1. Data preparation
-##Abundance data
-```{r}
-##read in the data: Simstrat model data and Projet Lac combined
-projet_lac_data <- read.csv("PLDB_final_short_depth_climate_cleantaxa_20052022.csv")
-
-str(projet_lac_data)
-
-# preparing additional column with abundance data of each species
-abundance_data <- projet_lac_data |> 
-  # exclude rows without taxa information
-  filter(!Taxa_latin_FINAL %in% "NO_FISH") |> 
-  # exclude samples obtained by electro-fishing
-  filter(!Protocol %in% "electro") |> 
-  #select columns relevant for species individual occurrences
- group_by(Lake, Basin3, Fishec_action, mean_temp_per_day, mean_last_7days, Weightg_raw, Weightg_soak, Depth_sample, Taxa_latin_FINAL, Protocol) |> 
-  # prepare column with abundance of each species
-  tally() |> 
-  # values_fill = 0 added to make all NAs into 0
-  pivot_wider(names_from = Taxa_latin_FINAL, values_from = n, values_fill = 0)
-
-# abundance of each species
-abundance_data_long <- abundance_data |> 
-  pivot_longer(cols = !(Lake : Protocol), names_to = "Species", values_to = "Abundance")
-
-# add presence-absence column
-abundance_data_long$Presence <- ifelse(abundance_data_long$Abundance >= 1, 1, 0)
-
-# View(abundance_data_long)
-
-max(abundance_data_long$Presence) #check if presence-absence only 0 and 1
-max(abundance_data_long$Abundance)
-
-#we only need information on species present in the lake, not all of them
-abundance_data_long_lakepresence <- abundance_data_long |> 
-  group_by(Lake, Species) |>
-  # . stands for data being piped in do function
-  do(LakePresence = max(.$Presence)) |> 
-  unnest(cols = LakePresence)
-
-# double-check 
-abundance_data_long_lakepresence |>
-  filter(LakePresence == "1") |> 
-  group_by(Lake, LakePresence) |>
-  count()
-
-# join the two dfs
-df_final <- abundance_data_long_lakepresence |> 
-  filter(LakePresence == "1") |> 
-  left_join(abundance_data_long )
-
-#double-check if all lakepresence = 1
-max(df_final$LakePresence)
-min(df_final$LakePresence)
-
-# save final df
-saveRDS(df_final, "df_final.rds")
-
-```
-
-##Taxonomy and requirements
-We pooled several species into ecomorphs to prevent loss of data-points. 
-We also excluded species that were caught less than 10 times in the whole Projet Lac.
-
-```{r}
-# read in prepared df
-df_final <- readRDS("df_final.rds")
-
-#######refining errors in species' identity that were noticed upon examination of the
-# data frame, following the table 16 in the Projet Lac "Synthesis report"
-
-# C. brienzii should not be present in lake Thun, add to Coregonus sp.
-
-df_final <- df_final |>
-  mutate(Species = ifelse(Species %in% c("Coregonus_brienzii") & Lake %in% c("Thun"),
-                          "Coregonus_sp", as.character(Species)))
-
-# Salvelinus umbla in Lake Thun should be Salvelinus sp.
-
-df_final <- df_final |>
-  mutate(Species = ifelse(Species %in% c("Salvelinus_umbla") & Lake %in% c("Thun"),
-                          "Salvelinus_sp", as.character(Species)))
-
-# Squalius squalus in Biel and Neuchatel to Squalius cephalus
-
-df_final <- df_final |>
-  mutate(Species = ifelse(Species %in% c("Squalius_squalus") & Lake %in% c("Biel", 
-                                                                           "Neuchatel"),
-                          "Squalius_cephalus", as.character(Species)))
-
-#######Pooling species to be able to include more samples
-# Profundal Cottus gobio are pooled
-
-df_final <- df_final |>
-  mutate(Species = ifelse(Species %in% c(
-    "Cottus_gobio_Profundal_Thun",
-    "Cottus_gobio_Profundal_Lucerne", "Cottus_gobio_Profundal_Walen"
-  ),
-  "Cottus_sp_Profundal", as.character(Species)
-  ))
-
-# Salvelinus is pooled into "profundal" and "limnetic" habitats
-
-# profundal
-
-df_final <- df_final |>
-  mutate(Species = ifelse(Species %in% c( "Salvelinus_sp_Profundal_dwarf_Thun", 
-                                          "Salvelinus_sp_Profundal_dwarf_VWS", 
-                                          "Salvelinus_sp_Profundal_extreme_Thun",
-    "Salvelinus_sp_Profundal_Walen_I", "Salvelinus_sp_Profundal_Walen_II"),
-  "Salvelinus_sp_Profundal", as.character(Species)
-  ))
-
-# limnetic 
-df_final <- df_final |>
-  mutate(Species = ifelse(Species %in% c("Salvelinus_sp_Limnetic_Thun", "Salvelinus_sp_Limnetic_VWS"),
-                          "Salvelinus_sp_Limnetic", as.character(Species)))
-
-# Coregonus species are pooled to new groups based on ecomorphs,
-# following De-Dayne et al. 2022
-
-# Albeli
-df_final <- df_final |>
-  mutate(Species = ifelse(Species %in% c("Coregonus_albellus", "Coregonus_candidus", "Coregonus_confusus",
-                                         "Coregonus_heglingus", "Coregonus_zugensis"),
-                          "Coregonus_sp_albeli", as.character(Species)))
-# Balchen
-df_final <- df_final |>
-  mutate(Species = ifelse(Species %in% c("Coregonus_alpinus", "Coregonus_arenicolus", "Coregonus_duplex",
-                        "Coregonus_helveticus", "Coregonus_palaea"),
-                          "Coregonus_sp_balchen", as.character(Species)))
-
-# Felchen
-
-df_final <- df_final |>
-  mutate(Species = ifelse(Species %in% c("Coregonus_brienzii", "Coregonus_fatioi", "Coregonus_intermundia",
-                                         "Coregonus_litoralis", "Coregonus_macrophthalmus", "Coregonus_zuerichensis"),
-                          "Coregonus_sp_felchen", as.character(Species)))
-# Large pelagic
-
-df_final <- df_final |>
-  mutate(Species = ifelse(Species %in% c("Coregonus_acrinasus", "Coregonus_wartmanni"),
-                          "Coregonus_sp_large_pelagic", as.character(Species)))
-
-# Benthic profundal
-df_final <- df_final |> 
-  mutate(Species = ifelse(Species == "Coregonus_profundus",
-                          "Coregonus_sp_benthic_profundal", as.character(Species)))
-# Pelagic profundal
-df_final <- df_final |> 
-  mutate(Species = ifelse(Species == "Coregonus_nobilis",
-                          "Coregonus_sp_pelagic_profundal", as.character(Species)))
-
-# "Coregonus_sarnensis": could not be assigned to Albeli or Felchen
-
-# check species list
-species <- df_final |> 
-  distinct(Species)
-
-# check how many species have less than 10 overall observations: 30
-df_final |> 
-  group_by(Species) |> 
-  summarize(tot_abu = sum(Presence)) |> 
-  filter(tot_abu <= 10)
-
-# exclusion of species with less than 10 overall observations
-df_models <- df_final |> 
-  group_by(Species) |> 
-  mutate(tot_abu = sum(Presence)) |> 
-  ungroup() |> 
-  filter(tot_abu >= 10)
-
-# save the df of the refined species, df will be used for the models
-saveRDS(df_models, "df_models.rds")
-
-df_models$Species <- as.factor(df_models$Species)
-levels(df_models$Species)
-
-species <- df_models |> 
-  group_by(Species) |> 
-  summarize(tot_obs = sum(Abundance))
-
-```
-##Data frames for GAMs
-We compiled four data frames for the four types of general additive models that 
-were run: Species with 1) binomial dataoccurring in one lake,
-2) abundance data occurring in one lake, 3) binomial data 
-occuring in multiple lakes, 4) abundance data and occuring in multiple lakes.
-
-```{r}
-###Overview
-#1. Separation of species into binomial and abundance data: sample with abundance > 1, 0 and 1 only -> binomial data
-#2. species that occur only in one lake: Lake not added as random effect
-
-#1. binomial vs. abundance data
-
-df_models$Species <- as.character(df_models$Species)
-
-non_binomial_species <- df_models |> 
-  filter(Abundance > 1) |> 
-  group_by(Species) |>
-  distinct(Species) |> 
-  pull(Species)
-
-binomial_species <- df_models |> 
-  filter(!Species %in% non_binomial_species) |> 
-  distinct(Species) |> 
-  pull(Species)
-
-#2. binomial vs. abundance species occuring only in one lake
-
-#1: GAM without lake as random effect, binomial
-
-bi_one_occurence <- df_models |> 
-  filter(Species %in% binomial_species) |>
-  group_by(Species) |>
-  summarize(n_lake = n_distinct(Lake)) |> 
-  filter(n_lake == 1) |>
-  pull(Species)
-
-# save df
-df_bi <- df_models |>
-  filter(Species %in% bi_one_occurence)
-
-saveRDS(df_bi, "data_frame_models/df_binomial_gam")
-
-#2 GAM abundance without lake as random effect
-abu_one_occurence <- df_models |> 
-  filter(Species %in% non_binomial_species) |> 
-  group_by(Species) |> 
-  summarize(n_lake = n_distinct(Lake)) |> 
-  filter(n_lake == 1) |>
-  pull(Species)
-
-df_abu <- df_models |> 
-  filter(Species %in% abu_one_occurence)
-
-saveRDS(df_abu, "data_frame_models/df_abundance_gam")
-
-#binomial vs. abundance species occuring in several lakes
-
-#3 gam with random intercept, binomial
-bi_multi_occurence <- df_models |> 
-  filter(Species %in% binomial_species) |> 
-  group_by(Species) |> 
-  summarize(Lakes = n_distinct(Lake)) |> 
-  filter(!Lakes == 1) |> 
-  pull(Species)
-
-df_binomial_re <- df_models |> 
-  filter(Species %in% bi_multi_occurence)
-
-saveRDS(df_binomial_re, "data_frame_models/df_binomial_re")
-
-#4 abundance GAM with random intercept
-abu_multi_occurence <- df_models |> 
-  filter(Species %in% non_binomial_species) |> 
-  group_by(Species) |> 
-  summarize(Lakes = n_distinct(Lake)) |> 
-  filter(!Lakes == 1) |> 
-  pull(Species)
-
-df_abundance_re <- df_models |> 
-  filter(Species %in% abu_multi_occurence)
-
-saveRDS(df_abundance_re, "data_frame_models/df_abundance_re")
-
-```
-
 #2. Data Analysis
@@
 
-# function for the GAMs (old)
-# source(here("function_GAM.R"))
-
-# new function for temp and depth models 
+# function for the temperature only GAMs
+source(here("function_GAM.R"))
+
+# function for temp and depth models 
 source(here("function_temp_depth.R"))
@@
 # total_models/predictions
-
-# DO NOT RUN NOW, RUN LATER!
+# with three subfolders: temp, temp_test, depth_temp (total_models/predictions_depth_temp)
+
+df_exclusion <- df_4 |> 
+  filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
+                   "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta", "Lepomis_gibbosus"))
+
+# all temperature models, no simplificaiton
+
+predictions(df_exclusion)
+
+predictions(df_1)
+predictions(df_2)
+predictions(df_3)
+predictions(df_4)
+
+# temperature models with simplified models
+predictions_temp(df_1)
+predictions_temp(df_2)
+predictions_temp(df_3)
+predictions_temp(df_4)
+
+# sensitivity
 predictions_depth_temp(df_1)
@@
 
+df_predictions_all <- list.files(path = "total_models/depth_temp/predictions", pattern = ".rds", full.names = TRUE) |>
+  map_dfr(readRDS)
+
+# save total predictions as RDS
+saveRDS(df_predictions_all, "total_models/df_pred_all.rds")
+
 df_predictions_all <- list.files(path = "total_models/predictions", pattern = ".rds", full.names = TRUE) |>
@@
 
-# save total predictions as RDS
-saveRDS(df_predictions_all, "total_models/df_pred_all.rds")
-
 ```
@@
 
-  model_0 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
-                   s(Depth_sample, k = 3)
-                            + s(fLake, bs = 're')
-                            +  s(fProtocol, bs = 're'), family = ziP())
-  print(glance(model_0))
+  # model_0 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
+  #                  s(Depth_sample, k = 3)
+  #                           + s(fLake, bs = 're')
+  #                           +  s(fProtocol, bs = 're'), family = ziP())
+  # print(glance(model_0))
 
@@
 
+AIC_testing(total_models, "Alburnus_arborella")
 # 8 models are run with binomial temp and depth
@@
 ```
-
-##Drivers of thermal response diversity
-
-Preparing a df with the information on past eutrophication, species richness, and species categorization in each lake.
-```{r}
-# Df with properties of lakes
-lake_info <- read_xlsx("lake_info.xlsx")
-
-# select the columns we use
-lake_selection <- lake_info |> 
-  select(Lake, Phos_max, Max_depth, Lake_area)
-
-# information about categorization in each lake for the subgroups
-species_endemism <- read_excel("species_category_per_lake.xlsx") |> 
-  rename(endemism = detail_category)
-
-# exclude lepomis gibbosus and re-calculate species richness per lake
-species_endemism_short <- species_endemism |> 
-  filter(species != "Lepomis_gibbosus") |> 
-  select(-sum_species) |> 
-  group_by(fLake) |> 
-  mutate(sum_species = sum(num_species))
-
-species_endemism_short$endemism <- as.factor(species_endemism_short$endemism)
-
-str(species_endemism_short)
-
-# adding column with number of species per category in each lake
-df_species_endemism <- species_endemism_short |> 
-  select(-species, - num_species) |> 
-  add_count(fLake, endemism) |> 
-  rename(count = n) |> 
-  distinct(fLake, count, sum_species, endemism)
-
-str(df_species_endemism)
-
-df_species_endemism$count <- as.numeric(df_species_endemism$count)
-
-# making it wide 
-df_species_endemism_long <- df_species_endemism |> 
-  pivot_wider(names_from = endemism, values_from = count) |> 
-  mutate_if(is.numeric, ~replace(., is.na(.), 0)) |> 
-  rename(Lake = fLake)
-
-str(df_species_endemism_long)
-
-# merge endemism info and lake info
-
-endemism_lake <- merge(df_species_endemism_long, lake_selection)
-
-# Read in data on response diversity metrics in each lake
-resp_div_all <- readRDS("total_models/resp_div_all.rds") |> 
-  select(temp:fLake)
-
-# select response diversity columns we need
-rdiv <- resp_div_all |>  
-  select(temp, fLake, rdiv, sign) |>
-  drop_na()
-
-# calculate mean and maximum values and rename Lake
-df_rdiv <- rdiv |> 
-  group_by(fLake) |> 
-  mutate(mean_rdiv = mean(rdiv)) |> 
-  mutate(max_rdiv = max(rdiv)) |> 
-  mutate(mean_sign = mean(sign)) |> 
-  mutate(max_sign = max(sign)) |> 
-  distinct(fLake, mean_rdiv, max_rdiv, mean_sign, max_sign) |> 
-  rename(Lake = fLake)
-
-# combine lake properties, species subgroups and response diversity
-
-df_lm <- merge(df_rdiv, endemism_lake)
-
-str(df_lm)
-
-# phosphorus values as numeric
-df_lm$Phos_max <- as.numeric(df_lm$Phos_max)
-
-```
-
-###Species richness and eutrophication
-Multiple linear regression to calculate the effect of species richness and
-eutrophication on response diversity.
-
-###Fig. 3
-
-```{r}
-# Multiple linear regression for mean dissimilarity explained by species richness and past eutrophication: 
-# lm(mean_rdiv  ~ species richness + eutrophication)
-# Predict manually with new data eutrophication value fixed at mean and species richness min() to max()
-# do the same the other way around: mean_rdiv  ~  eutrophication + species richness, with species richness fixed at mean and eutrophication min() to max()species richness +
-#   prepare these two plots
-
-# Phosphorus data on Poschiavo is not available
-df_lm_excl <- df_lm |> 
-  filter(Lake != "Poschiavo")
-
-# first lm for species richness ~ maximum phosphorus
-
-lm1 <- lm(mean_rdiv ~ sum_species + Phos_max, data = df_lm_excl)
-
-# checking model assumptions
-summary(lm1)
-shapiro.test(resid(lm1))
-lmtest::bptest(lm1)
-plot(lm1)
-
-# model assumptions validated
-
-# manually predict data
-new_data <- tibble(sum_species = seq(from = min(df_lm_excl$sum_species), to = max(df_lm_excl$sum_species), length = 50), Phos_max = mean(df_lm_excl$Phos_max))
-
-# predict lm
-prediction_lm1 <- predict.lm(lm1, newdata = new_data, se.fit = TRUE, type = "response")
-
-# combine new data and predictions
-bind_model1 <- cbind(new_data, prediction_lm1) |> 
-  mutate(category ="color")
-
-# second lm for maximum phosphorus ~ species richness
-lm2 <- lm(mean_rdiv ~ Phos_max + sum_species, data = df_lm_excl)
-
-# checking model assumptions
-summary(lm2)
-shapiro.test(resid(lm2))
-lmtest::bptest(lm2)
-plot(lm2)
-
-# new data
-new_data2 <- tibble(Phos_max = seq(from = min(df_lm_excl$Phos_max), to = max(df_lm_excl$Phos_max),length = 50), sum_species = mean(df_lm_excl$sum_species))
-
-# predict model
-prediction_lm2 <- predict.lm(lm2, newdata = new_data2, se.fit = TRUE, type = "response")
-
-# combine new data and predictions
-bind_model2 <- cbind(new_data2, prediction_lm2) |> 
-  mutate(category ="color")
-
-# Plot Fig. 3
-"#78BE63"
-mycolor <- c("color" = "#78BE63")
-# mycolor <- c("color" = "#A0E66A")
-"98D47C"
-
-lm1_plot <- bind_model1 |> 
-  ggplot(aes(sum_species, fit,  color = category, fill = category)) +
-  geom_line(linewidth = 1.2) +
-  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
-  theme_classic(base_size = 20) +
-  ylab("Mean dissimilarity") +
-  xlab("Species richness") +
-  ylim(1,3) +
-  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
-
-lm2_plot <- bind_model2 |> 
-  ggplot(aes(Phos_max, fit, color = category, fill = category)) +
-  geom_line(linewidth = 1.2) +
-  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
-  theme_classic(base_size = 20) +
-  ylab("Mean dissimilarity") +
-  xlab("Maximum total phosphorus (µg/L)") +
-  ylim(1,3) +
-  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
-
-# combine both plots
-figure_3 <- ggarrange(lm1_plot, lm2_plot)
-
-figure_3
-# save as TIFF
-# tiff(paste("total_models/plots/lm_species_richness_eutroph.tiff", sep = ""), units="in", width=12, height=5, res=300)
-# 
-# plot(figure_3)
-# 
-# # Closing the graphical device
-# dev.off()
-
-```
-
-###Divergence figure 3
-
-```{r}
-# first lm for species richness ~ maximum phosphorus
-
-lm3 <- lm(mean_sign ~ sum_species + Phos_max, data = df_lm_excl)
-
-# checking model assumptions
-summary(lm3)
-shapiro.test(resid(lm3))
-lmtest::bptest(lm3)
-plot(lm3)
-
-# model assumptions validated
-
-# manually predict data
-new_data3 <- tibble(sum_species = seq(from = min(df_lm_excl$sum_species), to = max(df_lm_excl$sum_species), length = 50), Phos_max = mean(df_lm_excl$Phos_max))
-
-# predict lm
-prediction_lm3 <- predict.lm(lm3, newdata = new_data3, se.fit = TRUE, type = "response")
-
-# combine new data and predictions
-bind_model3 <- cbind(new_data3, prediction_lm3) |> 
-  mutate(category ="color")
-
-# second lm for maximum phosphorus ~ species richness
-lm4 <- lm(mean_sign ~ Phos_max + sum_species, data = df_lm_excl)
-
-# checking model assumptions
-summary(lm4)
-shapiro.test(resid(lm4))
-lmtest::bptest(lm4)
-plot(lm4)
-
-# new data
-new_data4 <- tibble(Phos_max = seq(from = min(df_lm_excl$Phos_max), to = max(df_lm_excl$Phos_max),length = 50), sum_species = mean(df_lm_excl$sum_species))
-
-# predict model
-prediction_lm4 <- predict.lm(lm4, newdata = new_data4, se.fit = TRUE, type = "response")
-
-# combine new data and predictions
-bind_model4 <- cbind(new_data4, prediction_lm4) |> 
-  mutate(category ="color")
-
-# Plot Fig. 3
-mycolor <- c("color" = "#88D44F")
-# mycolor <- c("color" = "#A0E66A")
-"98D47C"
-
-lm3_plot <- bind_model3 |> 
-  ggplot(aes(sum_species, fit,  color = category, fill = category)) +
-  geom_line(linewidth = 1.2) +
-  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
-  theme_classic(base_size = 20) +
-  ylab("Mean divergence") +
-  xlab("Species richness") +
-  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
-
-lm4_plot <- bind_model4 |> 
-  ggplot(aes(Phos_max, fit, color = category, fill = category)) +
-  geom_line(linewidth = 1.2) +
-  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
-  theme_classic(base_size = 20) +
-  ylab("Mean divergence") +
-  xlab("Maximum total phosphorus (µg/L)") +
-  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
-
-# combine both plots
-figure_divergence <- ggarrange(lm3_plot, lm4_plot)
-
-figure_divergence
-# save as TIFF
-# tiff(paste("total_models/plots/lm_divergence_multiple_linear.tiff", sep = ""), units="in", width=12, height=5, res=300)
-# 
-# plot(figure_divergence)
-# 
-# # Closing the graphical device
-# dev.off()
-
-```
-###Community subgroups
-####Fig. 4a
-We fitted simple linear models for the number of species in each category and the mean dissimilarity in each lake. 
-```{r}
-# Lms for all categories: endemic, translcated, non_native and non_endemic_native
-# with data on all lakes
-# Endemic
-lm_endemic <- lm(mean_rdiv ~ endemic, data = df_lm)
-
-# checking model assumptions
-summary(lm_endemic)
-shapiro.test(resid(lm_endemic))
-lmtest::bptest(lm_endemic)
-plot(lm_endemic)
-
-new_data_a <- tibble(endemic = seq(from = min(df_lm$endemic), to = max(df_lm$endemic),
-                                     length = 15))
-
-prediction_endemic <- predict.lm(lm_endemic, newdata = new_data_a, se.fit = TRUE, type = "response")
-
-df_lm_endemic <- cbind(new_data_a, prediction_endemic) |> 
-  mutate(category = factor("endemic")) |> 
-  rename(species_richness = endemic)
-
-# Non-native
-
-lm_nn <- lm(mean_rdiv ~ non_native, data = df_lm)
-
-# checking model assumptions
-summary(lm_nn)
-shapiro.test(resid(lm_nn))
-lmtest::bptest(lm_nn)
-plot(lm_nn)
-
-new_data_b <- tibble(non_native = seq(from = min(df_lm$non_native), to = max(df_lm$non_native),
-                                   length = 15))
-
-prediction_nn <- predict.lm(lm_nn, newdata = new_data_b, se.fit = TRUE, type = "response")
-
-df_lm_nn <- cbind(new_data_b, prediction_nn)  |> 
-  mutate(category = factor("non_native")) |> 
-  rename(species_richness = non_native)
-
-# Non-endemic native
-
-lm_nne <- lm(mean_rdiv ~ native, data = df_lm)
-
-# checking model assumptions
-summary(lm_nne)
-shapiro.test(resid(lm_nne))
-lmtest::bptest(lm_nne)
-plot(lm_nne)
-
-new_data_c <- tibble(native = seq(from = min(df_lm$native), to = max(df_lm$native),
-                                      length = 15))
-
-prediction_nne <- predict.lm(lm_nne, newdata = new_data_c, se.fit = TRUE, type = "response")
-
-df_lm_nne <- cbind(new_data_c, prediction_nne)  |> 
-  mutate(category = factor("non_endemic_native")) |> 
-  rename(species_richness = native)
-
-# Translocated (non-native to region) species
-
-lm_trans <- lm(mean_rdiv ~ non_native_region, data = df_lm)
-
-# checking model assumptions
-summary(lm_trans)
-shapiro.test(resid(lm_trans))
-lmtest::bptest(lm_trans)
-plot(lm_trans)
-
-new_data_d <- tibble(non_native_region = seq(from = min(df_lm$non_native_region), to = max(df_lm$non_native_region),
-                                              length = 15))
-
-prediction_trans <- predict.lm(lm_trans, newdata = new_data_d, se.fit = TRUE, type = "response")
-
-df_lm_trans <- cbind(new_data_d, prediction_trans)  |> 
-  mutate(category = factor("translocated")) |> 
-  rename(species_richness = non_native_region)
-
-# Combine all four lm predictions
-df_all_lms <- rbind(df_lm_endemic, df_lm_nn, df_lm_nne, df_lm_trans)
-
-mycolors <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
-               "translocated" = "#F2AD63")
-
-plot_all_lms <- df_all_lms|> 
-  ggplot(aes(species_richness, fit, color = category, fill = category, linetype = category)) +  
-  geom_line(linewidth = 1) +
-  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), color = NA, alpha = 0.3) +
-  theme_classic(base_size = 16) +
-  ylab("Mean dissimilarity") +
-  xlab("Species richness") +
-  scale_fill_manual(values = mycolors,
-                    aesthetics = c("color", "fill"),
-                    labels = c("endemic", "non-native (ns)", "non-endemic native", "translocated (ns)")) +
-   scale_linetype_manual(values = c("endemic" = "solid",
-                                   "non_native" = "dashed",
-                                   "non_endemic_native" = "solid",
-                                   "translocated" = "dashed"), labels = c("endemic", "non-native (ns)", "non-endemic native", "translocated (ns)"))
-
-plot_all_lms
-# save as TIFF
-# tiff(paste("total_models/plots/lm_all.tiff", sep = ""), units="in", width=11, height=7, res=300)
-# # plot(ggarrange(depth1, depth2, ncol = 2))
-# # plot science discussion
-# plot(plot_all_lms)
-# 
-# dev.off()
-```
-
-####Divergence Fig. 4a
-
-```{r}
-# Lms for all categories: endemic, translcated, non_native and non_endemic_native
-# with data on all lakes
-# Endemic
-deriv_lm1 <- lm(mean_sign ~ endemic, data = df_lm)
-
-# checking model assumptions
-summary(deriv_lm1)
-shapiro.test(resid(deriv_lm1))
-lmtest::bptest(deriv_lm1)
-plot(deriv_lm1)
-
-new_data_1 <- tibble(endemic = seq(from = min(df_lm$endemic), to = max(df_lm$endemic),
-                                     length = 15))
-
-prediction_endemic_1 <- predict.lm(deriv_lm1, newdata = new_data_1, se.fit = TRUE, type = "response")
-
-df_deriv_endemic <- cbind(new_data_1, prediction_endemic_1) |> 
-  mutate(category = factor("endemic")) |> 
-  rename(species_richness = endemic)
-
-# Non-native
-
-deriv_lm2 <- lm(mean_sign ~ non_native, data = df_lm)
-
-# checking model assumptions
-summary(deriv_lm2)
-shapiro.test(resid(deriv_lm2))
-lmtest::bptest(deriv_lm2)
-plot(deriv_lm2)
-
-new_data_2 <- tibble(non_native = seq(from = min(df_lm$non_native), to = max(df_lm$non_native),
-                                   length = 15))
-
-prediction_2 <- predict.lm(deriv_lm2, newdata = new_data_2, se.fit = TRUE, type = "response")
-
-df_deriv_nn <- cbind(new_data_2, prediction_2)  |> 
-  mutate(category = factor("non_native")) |> 
-  rename(species_richness = non_native)
-
-# Non-endemic native
-
-deriv_lm3 <- lm(mean_sign ~ native, data = df_lm)
-
-# checking model assumptions
-summary(deriv_lm3)
-shapiro.test(resid(deriv_lm3))
-lmtest::bptest(deriv_lm3)
-plot(deriv_lm3)
-
-new_data_3 <- tibble(native = seq(from = min(df_lm$native), to = max(df_lm$native),
-                                      length = 15))
-
-prediction_3 <- predict.lm(deriv_lm3, newdata = new_data_3, se.fit = TRUE, type = "response")
-
-df_deriv_nne <- cbind(new_data_3, prediction_3)  |> 
-  mutate(category = factor("non_endemic_native")) |> 
-  rename(species_richness = native)
-
-# Translocated (non-native to region) species
-
-deriv_lm4 <- lm(mean_sign ~ non_native_region, data = df_lm)
-
-# checking model assumptions
-summary(deriv_lm4)
-shapiro.test(resid(deriv_lm4))
-lmtest::bptest(deriv_lm4)
-plot(deriv_lm4)
-
-new_data_4 <- tibble(non_native_region = seq(from = min(df_lm$non_native_region), to = max(df_lm$non_native_region),
-                                              length = 15))
-
-prediction_4 <- predict.lm(deriv_lm4, newdata = new_data_4, se.fit = TRUE, type = "response")
-
-df_deriv_trans <- cbind(new_data_4, prediction_4)  |> 
-  mutate(category = factor("translocated")) |> 
-  rename(species_richness = non_native_region)
-
-# Combine all four lm predictions
-df_deriv_lms <- rbind(df_deriv_endemic, df_deriv_nn, df_deriv_nne, df_deriv_trans)
-
-mycolors <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
-               "translocated" = "#F2AD63")
-
-lm_divergence_all <- df_deriv_lms|> 
-  ggplot(aes(species_richness, fit, color = category, fill = category)) +
-  geom_line(linewidth = 1, linetype = "dashed") +
-  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), color = NA, alpha = 0.3) +
-  theme_classic(base_size = 16) +
-  ylab("Mean divergence") +
-  xlab("Species richness") +
-  scale_fill_manual(values = mycolors,
-                    aesthetics = c("color", "fill"), 
-                    labels = c("endemic", "non-native", "non-endemic native", "translocated"),
-                    name = NULL)
-
-lm_divergence_all
-
-# save as TIFF
-# tiff(paste("total_models/plots/lm_divergence_all.tiff", sep = ""), units="in", width=11, height=7, res=300)
-# # plot(ggarrange(depth1, depth2, ncol = 2))
-# # plot science discussion
-# plot(lm_divergence_all)
-# 
-# dev.off()
-```
-#### Fig. 4b 
-Predictions of all species shown inside the categories. 
-```{r, fig.show='hide'}
-
-# Plot species predictions inside categories across all lakes
-# 5 categories;
-# endemic: geographically constrained range
-# non-native: from NA or Asia 
-# non-endemic native: native to Switzerland but not endemic
-# non-native region: native to Switzerland and surroundings but has been translocated
-# to other Swiss lakes where the species was not native
-# endemic translocated: endemic species that were translocated to other lakes
-
-# read in table with the categories and rename the categories
-species_category <- read_excel("species_category_across_lakes.xlsx") |> 
-  select(-notes) |> 
-  mutate(category = ifelse(category == "native", "non_endemic_native", category)) |> 
-  mutate(category = ifelse(category == "non_native_region", "non_endemic_native_and_translocated", category)) |> 
-  mutate(category2 = factor("all"))
-
-species_category$category <- as.factor(species_category$category)
-levels(species_category$category)
-
-# merge predictions with categories
-model_pred_categories <- merge(model_predictions, species_category)
-
-# change category names for facet_wrap
-category_names <- c(
-  `endemic` = "endemic",
-  `endemic_and_translocated` = "translocated endemic",
-  `non_endemic_native` = "non-endemic native",
-  `non_endemic_native_and_translocated` = "translocated non-endemic native",
-  `non_native` = "non-native"
-)
-
-# colors for the plot
-mycolors1 <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
-                "endemic_and_translocated" = "#F2AD63", "non_endemic_native_and_translocated" = "#F2AD63")
-
-# Rescale the predictions to 0-1 
-model_pred_categories_rescaled <- model_pred_categories |> 
-  group_by(species) |> 
-  mutate(fit_rescaled = (fit - min(fit)) / (max(fit) - min(fit)))
-
-# Plot the rescaled predictions
-plot_pred <- model_pred_categories_rescaled |>
-  # exclude Lepomis gibbosus
-  filter(species != "Lepomis_gibbosus") |> 
-  ggplot(aes(temp, fit_rescaled, group = species, color = category)) +
-  geom_line(linewidth = 0.8) +
-  theme_bw(base_size = 16) +
-  ylab("Abundance") +
-  xlab("Temperature") +
-  scale_color_manual(values = mycolors1, guide = NULL) +
-  theme(panel.grid.major = element_blank(), 
-        panel.grid.minor = element_blank(), 
-        legend.position = "bottom",
-        legend.title = element_blank())
-
-plot_category_predictions <- plot_pred + facet_grid(~category, labeller = as_labeller(category_names))
-
-plot_category_predictions
-
-# save plot as TIFF
-# tiff(paste("total_models/plots/plot_category_predictions.tiff", sep = ""), units="in", width=15, height=4, res=300)
-# 
-# plot(plot_category_predictions)
-# 
-# dev.off()
-
-```

### 2024-07-05 models_derivatives_response_diversity.Rmd
run all the models to get the simplified and complex models, derivatives now

@@
 # total_models/predictions
-# with three subfolders: temp, temp_test, depth_temp (total_models/predictions_depth_temp)
-
-df_exclusion <- df_4 |> 
-  filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
-                   "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta", "Lepomis_gibbosus"))
+# with three subfolders: temp, temp_test, depth_temp (total_models/predictions/depth_temp)
 
 # all temperature models, no simplificaiton
-
-predictions(df_exclusion)
-
-predictions(df_1)
-predictions(df_2)
-predictions(df_3)
-predictions(df_4)
+# predictions(df_1)
+# predictions(df_2)
+# predictions(df_3)
+# predictions(df_4)
+# 
+# no_exclusion_predictions_all <- list.files(path = "total_models/predictions/temp_test", pattern = ".rds", full.names = TRUE) |>
+#   map_dfr(readRDS)
+# 
+# # save total predictions as RDS
+# saveRDS(no_exclusion_predictions_all, "total_models/df_pred_no_exclusion.rds")
 
 # temperature models with simplified models
+
 predictions_temp(df_1)
@@
 
+no_exclusion_predictions_all <- list.files(path = "total_models/predictions/temp", pattern = ".rds", full.names = TRUE) |>
+  map_dfr(readRDS)
+
+# save total predictions as RDS
+saveRDS(no_exclusion_predictions_all, "total_models/df_pred_all.rds")
+
 # sensitivity
-predictions_depth_temp(df_1)
-predictions_depth_temp(df_2)
-predictions_depth_temp(df_3)
-predictions_depth_temp(df_4)
-
-# issue with cottus sp po
-# statistics of the gamms
-total_models <- bind_rows(df_1, df_2, df_3, df_4)
-
-stats_gamms(total_models)
-
-#  compile data frame with all predictions from the saved species' model predictions
-
-df_predictions_all <- list.files(path = "total_models/depth_temp/predictions", pattern = ".rds", full.names = TRUE) |>
-  map_dfr(readRDS)
-
-# save total predictions as RDS
-saveRDS(df_predictions_all, "total_models/df_pred_all.rds")
-
-df_predictions_all <- list.files(path = "total_models/predictions", pattern = ".rds", full.names = TRUE) |>
-  map_dfr(readRDS)
+# predictions_depth_temp(df_1)
+# predictions_depth_temp(df_2)
+# predictions_depth_temp(df_3)
+# predictions_depth_temp(df_4)
+# 
+# # issue with cottus sp po
+# # statistics of the gamms
+# total_models <- bind_rows(df_1, df_2, df_3, df_4)
+# 
+# stats_gamms(total_models)
+# 
+# #  compile data frame with all predictions from the saved species' model predictions
+# 
+# depth_temp_predictions_all <- list.files(path = "total_models/predictions/depth_temp", pattern = ".rds", full.names = TRUE) |>
+#   map_dfr(readRDS)
+# 
+# # save total predictions as RDS
+# saveRDS(depth_temp_predictions_all, "total_models/depth_temp_pred_all.rds")
 
@@
 
-###AIC comparisons in special cases
-
-```{r}
-
-AIC_testing <- function(data2, i){
-
-  data <- data2 |> 
-    filter(Species == i)
-
-  data$fProtocol <- as.factor(data$Protocol)
-  data$fLake <- as.factor(data$Lake)
-
-  # model_0 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
-  #                  s(Depth_sample, k = 3)
-  #                           + s(fLake, bs = 're')
-  #                           +  s(fProtocol, bs = 're'), family = ziP())
-  # print(glance(model_0))
-
-model1 <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) + s(Depth_sample, k = 3)
-                            + s(fLake, bs = 're')
-                            +  s(fProtocol, bs = 're'), family = binomial)
-print(glance(model1))
-
-model2 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
-                            + s(fLake, bs = 're')
-                            +  s(fProtocol, bs = 're'), family = ziP())
-print(glance(model2))
-
-}
-
-AIC_testing(df_4, "Lepomis_gibbosus")
-
-# in Alosa agone: model_0 not running, smaller AIC for binomial depth + temp model -> will be modelled as binomial depth + temp
-# same for Coregonus_sp_large_pelagic
-# lepomis gibbosus also has the smallest AIC for binomial depth + temp model
-
-# also testing the other special cases:
- # "Alburnus_arborella", "Barbatula_sp_Lineage_I","Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta"
-
-AIC_testing(df_4, "Alburnus_arborella") 
-# not running with model0, Binomial with depth and temp best
-AIC_testing(df_4, "Barbatula_sp_Lineage_I")
-# binomial depth and temp has lowest AIC
-AIC_testing(df_4, "Cyprinus_carpio")
-# binomial depth and temp has lowest AIC 
-AIC_testing(df_4, "Phoxinus_csikii")
-#binomial depth and temp has lowest AIC 
-AIC_testing(df_4, "Salmo_trutta")
-# binomial depth and temp lowest AIC
-
-# AIC_testing(df_4, "Cottus_sp_Po")
-# # best model is the temp, depth model
-
-AIC_testing(total_models, "Alburnus_arborella")
-# 8 models are run with binomial temp and depth
-
-```
 ####Fig. 2
@@
 
-# separating the dfs to separate model types
-depth_binomial_gam <- df_binomial_gam |> 
-  filter(tot_abu > 20)
-
-temp_binomial_gam <- df_binomial_gam |> 
-  filter(tot_abu < 20)
-
-depth_abundance_gam <- df_abundance_gam |> 
-  filter(tot_abu > 20)
-
-temp_abundance_gam <- df_abundance_gam |> 
-  filter(tot_abu < 20)
-
-depth_binomial_re <- df_binomial_re |> 
-  filter(tot_abu > 20)
-
-temp_binomial_re <- df_binomial_re |> 
-  filter(tot_abu < 20)
-
-depth_abundance_re <- df_abundance_re |> 
-  filter(tot_abu > 20)
-
-temp_abundance_re <- df_abundance_re |> 
-  filter(tot_abu < 20)
-
 ```
 
-###Depth and temp models
+###Temp models
 ```{r}
-# DEPTH AND TEMPERATURE
-# UNCLEAR HOW THIS CAN BE DONE
+
+# Now four different loops are run to calculate derivatives of each species 
+# in each lake
+
 # Model 1: Binomial data without random intercepts
 
-species_list <- depth_binomial_gam |>
+species_list <- df_binomial_gam |>
   distinct(Species) |>
@@
 
-i <- "Salmo_sp_Blackspot"
-j <- "Poschiavo"
 # loop through the list of species
 for (i in species_list) {
-  data <- depth_binomial_gam |>
+  data <- df_binomial_gam |>
     filter(Species == i)
@@
 
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
-                           s(Depth_sample, k = 3) + s(fProtocol, bs = 're'),
-                         family = binomial)
+  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'), family = binomial)
 
@@
      newdata <- tibble(mean_last_7days = seq(from = min(data_lake$temp, na.rm = TRUE),
-                                             to = max(data_lake$temp, na.rm = TRUE)
-                                             ,length = 200), Depth_sample =
-                         mean(data$Depth_sample, na.rm = TRUE),
+                                             to = max(data_lake$temp, na.rm = TRUE), length = 200),
                        fProtocol = factor("VERT"))
@@
      # calculate derivatives
-     # derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-     #   mutate(fLake = factor(j)) |>
-     #   mutate(species = factor(i))
-     #   rename(temp = data)
-     #
-
-       test_derivatives <- derivative_samples(gam_output[[i]]
-                                              , focal = "mean_last_7days", data = newdata)
-
-       test_derivatives |>
-         filter(.draw <= 200) |>
-         ggplot(aes(x = mean_last_7days, y = .derivative, group = .draw)) +
-         geom_line(alpha = 0.5)
-
-       derivatives <- test_derivatives |> 
-         filter(.draw == 500) |> 
+     derivatives <- derivatives(gam_output[[i]], data = newdata) |>
        mutate(fLake = factor(j)) |>
-         mutate(species = factor(i))
-
-       derivatives |> 
-         ggplot(aes(mean_last_7days, .derivative)) +
-         geom_line()
-
-     saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-  }
-}
-
-# needs to be discussed
-
-# nothing changed starting from here
-############################################################################3##3
+       mutate(species = factor(i)) |>
+       rename(temp = mean_last_7days)
+
+     saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))
+  }
+}
+
 # Model 2.1: Abundance data and no random intercept for lake -> but binomial
@@
 
-species_list <- depth_abundance_gam |> 
+species_list <- df_abundance_gam |> 
   # binomial one
@@
       mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
+      mutate(species = factor(i))
+      rename(temp = mean_last_7days)
+
+    saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))
   }
@@
       mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
+      rename(temp = mean_last_7days)
+
+    saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))
   }
@@
       mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
+      rename(temp = mean_last_7days)
+
+    saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))
 
@@
       mutate(species = factor(i)) |>
-      rename(temp = data)
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
+      rename(temp = mean_last_7days)
+    saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))
 
@@
       mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-
-  }
-}
-
-# save all the derivatives in one file
-
-df_deriv <- list.files(path = "total_models/derivatives", pattern = ".rds", full.names = TRUE) |> 
-  map_dfr(readRDS)
-
-saveRDS(df_deriv, "total_models/df_deriv_all.rds")
-
-# save total derivatives as RDS (no exclusions)
-# saveRDS(df_deriv, "total_models/df_deriv_all_no_excl.rds")
-
-```
-
-###Temp models
-```{r}
-
-# Now four different loops are run to calculate derivatives of each species 
-# in each lake
-
-# Model 1: Binomial data without random intercepts
-
-species_list <- df_binomial_gam |>
-  distinct(Species) |>
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-# loop through the list of species
-for (i in species_list) {
-  data <- df_binomial_gam |>
-    filter(Species == i)
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'), family = binomial)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  # loop through all lakes where the species is present, lake list is based on the information from
-  # Projet Lac
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    # using minimum and maximum temp in the lake for derivatives
-     newdata <- tibble(mean_last_7days = seq(from = min(data_lake$temp, na.rm = TRUE),
-                                             to = max(data_lake$temp, na.rm = TRUE), length = 200),
-                       fProtocol = factor("VERT"))
-
-     # calculate derivatives
-     derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-       mutate(fLake = factor(j)) |>
-       mutate(species = factor(i)) |>
-       rename(temp = data)
-
-     saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-  }
-}
-
-# Model 2.1: Abundance data and no random intercept for lake -> but binomial
-# Coregonus_sp_benthic_profundal did not run with ZIP and was thus
-# run with binomial.
-
-species_list <- df_abundance_gam |> 
-  # binomial one
-  filter(Species == "Coregonus_sp_benthic_profundal") |>
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_gam |> 
-    filter(Species == i)
-
-  # binomial family instead of ZIP
-  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) +
-                           s(fProtocol, bs = 're'), family = binomial())
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-  }
-}
-
-# Model 2.2: Abundance data and no random intercept for lake -> ZIP
-# all other species in this group run with ZIP
-
-species_list <- df_abundance_gam |>
-  filter(!Species == "Coregonus_sp_benthic_profundal") |>
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_gam |> 
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'),
-                         family = ziP())
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-  }
-}
-
-# Model 3: Binomial data and random intercept for lake
-
-species_list <- df_binomial_re |> 
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_binomial_re |> 
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = "re")
-                         + s(fProtocol, bs = 're'), family = binomial)
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  unique_lakes <- distinct(data, Lake) |> 
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    # take all lakes where the species is present and take random one for 
-    # the data included in derivatives()
-    random_lake <- sample(unique_lakes, 1)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), lenqh = 200),
-      fProtocol = factor("VERT"), fLake = factor(random_lake))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-
-  }
-}
-
-# Model 4.1: Abundance data and random intercept for lake, simplified models
-# multiple species in this group were simplified and run with binomial
-
-species_list <- df_abundance_re |>
-  # simplified models
-  filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
-                                "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta",
-                        "Lepomis_gibbosus")) |>
-  distinct(Species) |>
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_re |>
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
-                         +  s(fProtocol, bs = 're'), family = binomial)
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  unique_lakes <- distinct(data, Lake) |>
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    random_lake <- sample(unique_lakes, 1)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fLake = factor(random_lake), fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-
-  }
-}
-
-# Model 4.1: Abundance data and random intercept for lake, complex models
-
-species_list <- df_abundance_re |> 
-  # exclusion of simplified models
-  filter(!Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
-                        "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta", 
-                        "Lepomis_gibbosus")) |>
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_re |> 
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
-                         +  s(fProtocol, bs = 're'), family = ziP())
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  unique_lakes <- distinct(data, Lake) |> 
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    random_lake <- sample(unique_lakes, 1)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fLake = factor(random_lake), fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
+      rename(temp = mean_last_7days)
+
+    saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))
 

### 2024-07-05 models_derivatives_response_diversity.Rmd
rerun with lepomis gibbosus

@@
 library(grid)
+library(viridis)
 
@@
 ```{r, fig.show='hide'}
+
 # Load total predictions
@@
 
-     saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))
+     saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
   }
@@
       mutate(fLake = factor(j)) |>
-      mutate(species = factor(i))
+      mutate(species = factor(i)) |> 
       rename(temp = mean_last_7days)
 
-    saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))
+    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
   }
@@
 
-    saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))
+    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
   }
@@
       from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), lenqh = 200),
+      to = max(data_lake$temp, na.rm = TRUE), length = 200),
       fProtocol = factor("VERT"), fLake = factor(random_lake))
@@
 
-    saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))
+    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
+
+  }
+}
+
+# Model 4: before simplification
+
+species_list <- df_abundance_re |> 
+  # exclusion of simplified models
+  filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
+                        "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta",
+                        "Lepomis_gibbosus")) |>
+  distinct(Species) |> 
+  pull(Species)
+
+species_list <- sort(species_list)
+
+derivatives <- list()
+gam_output <- list()
+model_prediction <- list()
+
+for (i in species_list) {
+
+  data <- df_abundance_re |> 
+    filter(Species == i)
+
+  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
+                         +  s(fProtocol, bs = 're'), family = ziP())
+
+  lake_data <- species_lake |>
+    filter(Species == i)
+
+  lake_list <- distinct(lake_data, Lake) |>
+    pull()
+
+  unique_lakes <- distinct(data, Lake) |> 
+    pull()
+
+  for (j in lake_list){
+
+    data_lake <- species_lake |>
+      filter(Species == i) |>
+      filter(Lake == j)
+
+    random_lake <- sample(unique_lakes, 1)
+
+    newdata <- tibble(mean_last_7days = seq(
+      from = min(data_lake$temp, na.rm = TRUE),
+      to = max(data_lake$temp, na.rm = TRUE), length = 200),
+      fLake = factor(random_lake), fProtocol = factor("VERT"))
+
+    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
+      mutate(fLake = factor(j)) |>
+      mutate(species = factor(i)) |>
+      rename(temp = mean_last_7days)
+
+    saveRDS(derivatives, paste0("total_models/derivatives_no_simplification/derivatives_", i, "_",  j, ".rds"))
 
@@
       rename(temp = mean_last_7days)
-    saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))
+    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
 
@@
 
-    saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))
+    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
 
@@
 # save all the derivatives in one file
-
+# 
 df_deriv <- list.files(path = "total_models/derivatives", pattern = ".rds", full.names = TRUE) |>
@@
 # save total derivatives as RDS (no exclusions)
-# saveRDS(df_deriv, "total_models/df_deriv_all_no_excl.rds")
+# 
+# df_deriv_no_excl <- list.files(path = "total_models/derivatives_no_simplification", pattern = ".rds", full.names = TRUE) |>
+#   map_dfr(readRDS)
+# 
+# saveRDS(df_deriv_no_excl, "total_models/df_deriv_all_no_excl.rds")
 
@@
   arrange(fLake) |> 
-  ggplot(aes(temp, derivative, color = species)) +
+  ggplot(aes(temp, .derivative, color = species)) +
   geom_line() +
@@
 ```{r}
+# cyprinus carpio and lepomis should not be simplified
+
 # Data of the derivatives before simplifying multiple models
@@
   group_by(species, fLake) |> 
-  mutate(max_derivative = max(derivative)) |> 
-  mutate(mean_derivative = mean(derivative)) |> 
+  mutate(max_derivative = max(.derivative)) |> 
+  mutate(mean_derivative = mean(.derivative)) |> 
   ungroup()
@@
 # calculate quantiles of derivatives
-transform(quantile(df_quantiles$derivative,
+transform(quantile(df_quantiles$.derivative,
                    c(0,0.01,0.05,0.1,0.25,0.5,0.75,0.85,0.9,0.95,0.99,1)))
@@
 # add column with the category of quantile each derivative is in
-df_quantiles$groups <- cut(df_quantiles$derivative,              
-                     breaks = c(-1.579943e+06, -2.342866e+00, -8.822087e-01,
-                                -4.543748e-01, 3.454218e-04, 2.485142e-01,
-                                5.527792e-01, 2.911987e+00, 2.018455e+01, 1.280898e+02,
-                                1.231212e+07, 1.406101e+07),
+df_quantiles$groups <- cut(df_quantiles$.derivative,              
+                     breaks = c(-1.579943e+06, -2.706252e+00, -1.269492e+00,
+                                -3.348207e-01, 1.844665e-03, 2.262231e-01,
+                                9.864616e-01, 2.018455e+01, 3.013806e+01, 1.648215e+06,
+                                1.303406e+07, 1.406101e+07),
                      labels = c("0-1%", "1-5%", "5-10%", "10-25%", "25-50%", "50-75%", "75-85%", "85-90%", "90-95%",
@@
 plot_percentiles_1 <- plot_percentiles +
-  geom_vline(xintercept = 19.9, color = "red") +
-  annotate("text", x= 18, y = 0.12, label="90th percentile", angle=90) +
-  geom_vline(xintercept = 2.911987e+00, color = "red") +
-  annotate("text", x= 1.5, y = 0.12, label="85th percentile", angle=90) +
+  geom_vline(xintercept = 3.013806e+01, color = "red") +
+  annotate("text", x= 26, y = 0.12, label="90th percentile", angle=90) +
+  geom_vline(xintercept = 2.018455e+01, color = "red") +
+  annotate("text", x= 16, y = 0.12, label="85th percentile", angle=90) +
   theme_classic(base_size = 13) +
@@
 df_quantiles |> 
-  filter(derivative > 2.018455e+01) |> 
+  filter(.derivative > 2.018455e+01) |> 
   distinct(species)
@@
 # Breaking mean derivatives in the same percentiles as before
+
 data_quantiles$percentiles <- cut(data_quantiles$mean_derivative,              
-                     breaks = c(-1.579943e+06, -2.342866e+00, -8.822087e-01,
-                                -4.543748e-01, 3.454218e-04, 2.485142e-01,
-                                5.527792e-01, 2.911987e+00, 2.018455e+01, 1.280898e+02,
-                                1.231212e+07, 1.406101e+07),
+                     breaks = c(-1.579943e+06, -2.706252e+00, -1.269492e+00,
+                                -3.348207e-01, 1.844665e-03, 2.262231e-01,
+                                9.864616e-01, 2.018455e+01, 3.013806e+01, 1.648215e+06,
+                                1.303406e+07, 1.406101e+07),
                      labels = c("0-1%", "1-5%", "5-10%", "10-25%", "25-50%", "50-75%", "75-85%", "85-90%", "90-95%",
@@
 
-# Lepomis is excluded from the analysis
-all_deriv <- all_deriv |>
-  filter(!species %in% c("Lepomis_gibbosus"))
-
 all_deriv$fLake <- as.character(all_deriv $fLake)
@@
 # loop through list of lakes and calculate response diversity in the lake communities
+
 for (i in lakes_list){
@@
   data <- all_deriv |>
-    select(temp, fLake, derivative, species) |> 
+    select(.derivative, temp, species, fLake) |> 
     filter(fLake == i)
@@
       names_from = species,
-      values_from = derivative)
+      values_from = .derivative)
 
@@
   # save response diversity metrics per lake
-  saveRDS(df_resp_div, paste0("total_models/lakes_all_models/df_resp_div_", i, ".rds"))
+  saveRDS(df_resp_div, paste0("total_models/resp_div_lakes/df_resp_div_", i, ".rds"))
 
@@
 # save the response diversity per lake into one file
-resp_div_no_excl <- list.files(path = "total_models/lakes_all_models", pattern = ".rds", full.names = TRUE) |>
+resp_div_no_excl <- list.files(path = "total_models/resp_div_lakes", pattern = ".rds", full.names = TRUE) |>
   map_dfr(readRDS) |>

### 2024-07-05 function_temp_depth.R
function for modelling temp + depth

@@ -87,7 +87,7 @@ predictions_depth_temp <- function(df){
             mutate(species = factor(i))
           
           # saving the predictions
-          saveRDS(pred_df, paste0("total_models/predictions/predictions_",i,".rds"))
+          saveRDS(pred_df, paste0("total_models/predictions/depth_temp/predictions_",i,".rds"))
           
           # # stats
           # glance_summary <- glance(gam_output)
@@ -160,7 +160,7 @@ predictions_depth_temp <- function(df){
             mutate(species = factor(i))
           
           # saving the predictions
-          saveRDS(pred_df, paste0("total_models/predictions/predictions_",i,".rds"))
+          saveRDS(pred_df, paste0("total_models/predictions/depth_temp/predictions_",i,".rds"))
           
           # # stats
           # glance_summary <- glance(gam_output)
@@ -232,7 +232,7 @@ predictions_depth_temp <- function(df){
             mutate(species = factor(i))
           
           # saving the predictions
-          saveRDS(pred_df, paste0("total_models/predictions/predictions_",i,".rds"))
+          saveRDS(pred_df, paste0("total_models/predictions/depth_temp/predictions_",i,".rds"))
           
           # # stats
           # glance_summary <- glance(gam_output)
@@ -323,7 +323,7 @@ predictions_depth_temp <- function(df){
             rename(temp = mean_last_7days) |>
             mutate(species = factor(i))
           
-          saveRDS(pred_df, paste0("total_models/predictions/predictions_",i,".rds"))
+          saveRDS(pred_df, paste0("total_models/predictions/depth_temp/predictions_",i,".rds"))
           
           # # stats
           # glance_summary <- glance(gam_output)
@@ -468,7 +468,7 @@ predictions_depth_temp <- function(df){
             rename(temp = mean_last_7days) |>
             mutate(species = factor(i))
           
-          saveRDS(pred_df, paste0("total_models/predictions/predictions_",i,".rds"))
+          saveRDS(pred_df, paste0("total_models/predictions/depth_temp/predictions_",i,".rds"))
           
           # # stats
           # glance_summary <- glance(gam_output)
@@ -549,7 +549,7 @@ predictions_depth_temp <- function(df){
             mutate(species = factor(i))
           
           # saving the predictions
-          saveRDS(pred_df, paste0("total_models/predictions/predictions_",i,".rds"))
[…]

### 2024-07-05 function_GAM.R
modelling temp only models

@@ -1,0 +1,743 @@
+################ Function GAMs
+# Function to model species' abundances along temperature with generalized additive models. 
+# Prediction of GAMs with mgcv and validation of model assumptions with DHARMa. 
+# predictions() function is a loop to separate the species
+# into the type of GAM they are predicted with based on 1) the number of lakes 
+# the species occurs in, 2) the distribution of their abundance data. 
+
+# We use 4 general types of models:
+# 1) binomial with net-type as random intercept
+# 2) ZIP with net-type as random intercept
+# 3) binomial with net-type and lake as random intercepts
+# 4) ZIP with net-type and lake as random intercepts
+
+# and had 6 exceptions: 
+# Coregonus_sp_benthic_profundal: Model did not run with ZIP and was thus modelled
+# with binomial
+
+# 5 models were simplified:
+# 1 Alburnus_arborella    
+# 2 Barbatula_sp_Lineage_I
+# 3 Cyprinus_carpio       
+# 4 Phoxinus_csikii       
+# 5 Salmo_trutta  
+# those 5 species have derivatives above the 90th percentile
+# and modelled as binomial with net-type and lake as random intercepts
+
+
+# When predicting the GAMs with random effects we followed the suggestion here: 
+# https://fromthebottomoftheheap.net/2021/02/02/random-effects-in-gams/
+# (see comment section)
+# In the new data for the prediction a random level of the random effect is selected
+# and then the effect of the random effect smooth(s) is excluded with the exclude
+# argument.
+# Here, "VERT" is always used as the level for the random effect of the net-type. 
+# In the random effect for lakes, a random lake is chosen where the species is present
+# and used in the prediction.
+
+predictions_temp <- function(df){
+  require(broom)
+  require(tidyverse)
+  require(DHARMa)
+  require(mgcv)
+  require(gratia)
+  require(mgcViz)
+
+  
+  species_list <- df |> 
+    distinct(Species) |> 
+    pull(Species)
+  
+  species_list <- sort(species_list)
+  
+  gam_output <- list()
+  model_prediction <- list()
+  derivatives <- list()
+  grid <- list()
+  pred_df <- list()
+  tiff_filename <- list()
+  tiff_file_2 <- list()
+  temp_data <- list()
+  summary <- list()
+  
+  df$fLake <- as.factor(df$Lake)
+  
+  df$fProtocol <- as.factor(df$Protocol)
+  
+  for (i in species_list) {
+    data <- df |> 
+      filter(Species == i) |> 
[…]

### 2024-07-05 drivers_response_diversity.Rmd
part three of manuscript: drivers of response diversity

--- 2024-07-03 manuscript_analysis.Rmd
@@
 ---
-title: "manuscript_analysis"
+title: "drivers response diversity"
 author: "Sophie Moreau"
-date: "2024-07-03"
+date: "2024-07-05"
 output: html_document
@@
 ---
-
-```{r}
-library(tidyverse)
-library(here)
-library(readr)
-library(viridis)
-
-```
-
-#Introduction
-This is the supplementary information of the Master thesis. This document
-covers the whole analysis performed to assess response diversity between fish communities
-in Swiss peri-alpine lakes. 
-This document has two parts: Data preparation (1) and Data analysis (2). 
-Function to model general additive models is provided separately (see function_GAM.R). 
-Code for figures used in the method figure is also provided separately (see supplementary_method_figure.R)
-
-```{r}
-# do we want to include zurich and constance separately?
-projet_lac_data <- read.csv("PLDB_final_short_depth_climate_cleantaxa_20052022.csv")
-
-str(projet_lac_data)
-projet_lac_data$LakeBasin <- as.factor(projet_lac_data$LakeBasin)
-
-projet_lac_data |> 
-  distinct(LakeBasin)
-```
-
-#1. Data preparation
-##Abundance data
-```{r}
-##read in the data: Simstrat model data and Projet Lac combined
-projet_lac_data <- read.csv("PLDB_final_short_depth_climate_cleantaxa_20052022.csv")
-
-str(projet_lac_data)
-
-# preparing additional column with abundance data of each species
-abundance_data <- projet_lac_data |> 
-  # exclude rows without taxa information
-  filter(!Taxa_latin_FINAL %in% "NO_FISH") |> 
-  # exclude samples obtained by electro-fishing
-  filter(!Protocol %in% "electro") |> 
-  #select columns relevant for species individual occurrences
- group_by(Lake, Basin3, Fishec_action, mean_temp_per_day, mean_last_7days, Weightg_raw, Weightg_soak, Depth_sample, Taxa_latin_FINAL, Protocol) |> 
-  # prepare column with abundance of each species
-  tally() |> 
-  # values_fill = 0 added to make all NAs into 0
-  pivot_wider(names_from = Taxa_latin_FINAL, values_from = n, values_fill = 0)
-
-# abundance of each species
-abundance_data_long <- abundance_data |> 
-  pivot_longer(cols = !(Lake : Protocol), names_to = "Species", values_to = "Abundance")
-
-# add presence-absence column
-abundance_data_long$Presence <- ifelse(abundance_data_long$Abundance >= 1, 1, 0)
-
-# View(abundance_data_long)
-
-max(abundance_data_long$Presence) #check if presence-absence only 0 and 1
-max(abundance_data_long$Abundance)
-
-#we only need information on species present in the lake, not all of them
-abundance_data_long_lakepresence <- abundance_data_long |> 
-  group_by(Lake, Species) |>
-  # . stands for data being piped in do function
-  do(LakePresence = max(.$Presence)) |> 
-  unnest(cols = LakePresence)
-
-# double-check 
-abundance_data_long_lakepresence |>
-  filter(LakePresence == "1") |> 
-  group_by(Lake, LakePresence) |>
-  count()
-
-# join the two dfs
-df_final <- abundance_data_long_lakepresence |> 
-  filter(LakePresence == "1") |> 
-  left_join(abundance_data_long )
-
-#double-check if all lakepresence = 1
-max(df_final$LakePresence)
-min(df_final$LakePresence)
-
-# save final df
-saveRDS(df_final, "df_final.rds")
-
-```
-
-##Taxonomy and requirements
-We pooled several species into ecomorphs to prevent loss of data-points. 
-We also excluded species that were caught less than 10 times in the whole Projet Lac.
-
-```{r}
-# read in prepared df
-df_final <- readRDS("df_final.rds")
-
-#######refining errors in species' identity that were noticed upon examination of the
-# data frame, following the table 16 in the Projet Lac "Synthesis report"
-
-# C. brienzii should not be present in lake Thun, add to Coregonus sp.
-
-df_final <- df_final |>
-  mutate(Species = ifelse(Species %in% c("Coregonus_brienzii") & Lake %in% c("Thun"),
-                          "Coregonus_sp", as.character(Species)))
-
-# Salvelinus umbla in Lake Thun should be Salvelinus sp.
-
-df_final <- df_final |>
-  mutate(Species = ifelse(Species %in% c("Salvelinus_umbla") & Lake %in% c("Thun"),
-                          "Salvelinus_sp", as.character(Species)))
-
-# Squalius squalus in Biel and Neuchatel to Squalius cephalus
-
-df_final <- df_final |>
-  mutate(Species = ifelse(Species %in% c("Squalius_squalus") & Lake %in% c("Biel", 
-                                                                           "Neuchatel"),
-                          "Squalius_cephalus", as.character(Species)))
-
-#######Pooling species to be able to include more samples
-# Profundal Cottus gobio are pooled
-
-df_final <- df_final |>
-  mutate(Species = ifelse(Species %in% c(
-    "Cottus_gobio_Profundal_Thun",
-    "Cottus_gobio_Profundal_Lucerne", "Cottus_gobio_Profundal_Walen"
-  ),
-  "Cottus_sp_Profundal", as.character(Species)
-  ))
-
-# Salvelinus is pooled into "profundal" and "limnetic" habitats
-
-# profundal
-
-df_final <- df_final |>
-  mutate(Species = ifelse(Species %in% c( "Salvelinus_sp_Profundal_dwarf_Thun", 
-                                          "Salvelinus_sp_Profundal_dwarf_VWS", 
-                                          "Salvelinus_sp_Profundal_extreme_Thun",
-    "Salvelinus_sp_Profundal_Walen_I", "Salvelinus_sp_Profundal_Walen_II"),
-  "Salvelinus_sp_Profundal", as.character(Species)
-  ))
-
-# limnetic 
-df_final <- df_final |>
-  mutate(Species = ifelse(Species %in% c("Salvelinus_sp_Limnetic_Thun", "Salvelinus_sp_Limnetic_VWS"),
-                          "Salvelinus_sp_Limnetic", as.character(Species)))
-
-# Coregonus species are pooled to new groups based on ecomorphs,
-# following De-Dayne et al. 2022
-
-# Albeli
-df_final <- df_final |>
-  mutate(Species = ifelse(Species %in% c("Coregonus_albellus", "Coregonus_candidus", "Coregonus_confusus",
-                                         "Coregonus_heglingus", "Coregonus_zugensis"),
-                          "Coregonus_sp_albeli", as.character(Species)))
-# Balchen
-df_final <- df_final |>
-  mutate(Species = ifelse(Species %in% c("Coregonus_alpinus", "Coregonus_arenicolus", "Coregonus_duplex",
-                        "Coregonus_helveticus", "Coregonus_palaea"),
-                          "Coregonus_sp_balchen", as.character(Species)))
-
-# Felchen
-
-df_final <- df_final |>
-  mutate(Species = ifelse(Species %in% c("Coregonus_brienzii", "Coregonus_fatioi", "Coregonus_intermundia",
-                                         "Coregonus_litoralis", "Coregonus_macrophthalmus", "Coregonus_zuerichensis"),
-                          "Coregonus_sp_felchen", as.character(Species)))
-# Large pelagic
-
-df_final <- df_final |>
-  mutate(Species = ifelse(Species %in% c("Coregonus_acrinasus", "Coregonus_wartmanni"),
-                          "Coregonus_sp_large_pelagic", as.character(Species)))
-
-# Benthic profundal
-df_final <- df_final |> 
-  mutate(Species = ifelse(Species == "Coregonus_profundus",
-                          "Coregonus_sp_benthic_profundal", as.character(Species)))
-# Pelagic profundal
-df_final <- df_final |> 
-  mutate(Species = ifelse(Species == "Coregonus_nobilis",
-                          "Coregonus_sp_pelagic_profundal", as.character(Species)))
-
-# "Coregonus_sarnensis": could not be assigned to Albeli or Felchen
-
-# check species list
-species <- df_final |> 
-  distinct(Species)
-
-# check how many species have less than 10 overall observations: 30
-df_final |> 
-  group_by(Species) |> 
-  summarize(tot_abu = sum(Presence)) |> 
-  filter(tot_abu <= 10)
-
-# exclusion of species with less than 10 overall observations
-df_models <- df_final |> 
-  group_by(Species) |> 
-  mutate(tot_abu = sum(Presence)) |> 
-  ungroup() |> 
-  filter(tot_abu >= 10)
-
-# save the df of the refined species, df will be used for the models
-saveRDS(df_models, "df_models.rds")
-
-df_models$Species <- as.factor(df_models$Species)
-levels(df_models$Species)
-
-species <- df_models |> 
-  group_by(Species) |> 
-  summarize(tot_obs = sum(Abundance))
-
-```
-##Data frames for GAMs
-We compiled four data frames for the four types of general additive models that 
-were run: Species with 1) binomial dataoccurring in one lake,
-2) abundance data occurring in one lake, 3) binomial data 
-occuring in multiple lakes, 4) abundance data and occuring in multiple lakes.
-
-```{r}
-###Overview
-#1. Separation of species into binomial and abundance data: sample with abundance > 1, 0 and 1 only -> binomial data
-#2. species that occur only in one lake: Lake not added as random effect
-
-#1. binomial vs. abundance data
-
-df_models$Species <- as.character(df_models$Species)
-
-non_binomial_species <- df_models |> 
-  filter(Abundance > 1) |> 
-  group_by(Species) |>
-  distinct(Species) |> 
-  pull(Species)
-
-binomial_species <- df_models |> 
-  filter(!Species %in% non_binomial_species) |> 
-  distinct(Species) |> 
-  pull(Species)
-
-#2. binomial vs. abundance species occuring only in one lake
-
-#1: GAM without lake as random effect, binomial
-
-bi_one_occurence <- df_models |> 
-  filter(Species %in% binomial_species) |>
-  group_by(Species) |>
-  summarize(n_lake = n_distinct(Lake)) |> 
-  filter(n_lake == 1) |>
-  pull(Species)
-
-# save df
-df_bi <- df_models |>
-  filter(Species %in% bi_one_occurence)
-
-saveRDS(df_bi, "data_frame_models/df_binomial_gam")
-
-#2 GAM abundance without lake as random effect
-abu_one_occurence <- df_models |> 
-  filter(Species %in% non_binomial_species) |> 
-  group_by(Species) |> 
-  summarize(n_lake = n_distinct(Lake)) |> 
-  filter(n_lake == 1) |>
-  pull(Species)
-
-df_abu <- df_models |> 
-  filter(Species %in% abu_one_occurence)
-
-saveRDS(df_abu, "data_frame_models/df_abundance_gam")
-
-#binomial vs. abundance species occuring in several lakes
-
-#3 gam with random intercept, binomial
-bi_multi_occurence <- df_models |> 
-  filter(Species %in% binomial_species) |> 
-  group_by(Species) |> 
-  summarize(Lakes = n_distinct(Lake)) |> 
-  filter(!Lakes == 1) |> 
-  pull(Species)
-
-df_binomial_re <- df_models |> 
-  filter(Species %in% bi_multi_occurence)
-
-saveRDS(df_binomial_re, "data_frame_models/df_binomial_re")
-
-#4 abundance GAM with random intercept
-abu_multi_occurence <- df_models |> 
-  filter(Species %in% non_binomial_species) |> 
-  group_by(Species) |> 
-  summarize(Lakes = n_distinct(Lake)) |> 
-  filter(!Lakes == 1) |> 
-  pull(Species)
-
-df_abundance_re <- df_models |> 
-  filter(Species %in% abu_multi_occurence)
-
-saveRDS(df_abundance_re, "data_frame_models/df_abundance_re")
-
-```
-
-#2. Data Analysis
-##General Additive Models
-Running the predictions for all GAMs with the smaller data frames to reduce running 
-time.
 
@@
 
-###Predictions of GAMs
-
-```{r}
-# read in the dataframes
-df_1 <- readRDS("data_frame_models/df_binomial_gam")
-df_2 <- readRDS("data_frame_models/df_abundance_gam")
-df_3 <- readRDS("data_frame_models/df_binomial_re")
-df_4 <- readRDS("data_frame_models/df_abundance_re")
-
-# use own function predictions() from function_GAM.R to predict species' models
-# across all lakes
-
-# predictions take VERY LONG! and two folders are needed to store the 
-# DHARMa and prediction output inside: 
-# total_models/gam_check
-# total_models/predictions
-
-#df1 and df2 are the fastest ones
-# predictions() can also be used directly with df_models -> takes longer
-# BECAUSE OF TYPO: RE RUN TEMP + DEPTH models from df1-3, df4 still needs to be run & stats overview too
-
-# DO NOT RUN NOW, RUN LATER!
-predictions_depth_temp(df_1)
-predictions_depth_temp(df_2)
-predictions_depth_temp(df_3)
-predictions_depth_temp(df_4)
-
-# statistics of the gamms
-total_models <- bind_rows(df_1, df_2, df_3, df_4)
-
-stats_gamms(total_models)
-
-#  compile data frame with all predictions from the saved species' model predictions
-
-df_predictions_all <- list.files(path = "total_models/predictions", pattern = ".rds", full.names = TRUE) |>
-  map_dfr(readRDS)
-
-# save total predictions as RDS
-saveRDS(df_predictions_all, "total_models/df_pred_all.rds")
-
-```
-
-###AIC comparisons in special cases
-
-```{r}
-
-AIC_testing <- function(data2, i){
-
-  data <- data2 |> 
-    filter(Species == i)
-
-  data$fProtocol <- as.factor(data$Protocol)
-  data$fLake <- as.factor(data$Lake)
-
-  model_0 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
-                   s(Depth_sample, k = 3)
-                            + s(fLake, bs = 're')
-                            +  s(fProtocol, bs = 're'), family = ziP())
-  print(glance(model_0))
-# new models
-
-model1 <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) + s(Depth_sample, k = 3)
-                            + s(fLake, bs = 're')
-                            +  s(fProtocol, bs = 're'), family = binomial)
-print(glance(model1))
-
-model2 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
-                            + s(fLake, bs = 're')
-                            +  s(fProtocol, bs = 're'), family = ziP())
-print(glance(model2))
-
-}
-
-AIC_testing(df_4, "Lepomis_gibbosus")
-
-# in Alosa agone: model_0 not running, smaller AIC for binomial depth + temp model -> will be modelled as binomial depth + temp
-# same for Coregonus_sp_large_pelagic
-# lepomis gibbosus also has the smallest AIC for binomial depth + temp model
-
-# also testing the other special cases:
- # "Alburnus_arborella", "Barbatula_sp_Lineage_I","Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta"
-
-AIC_testing(df_4, "Alburnus_arborella") 
-# not running with model0, Binomial with depth and temp best
-AIC_testing(df_4, "Barbatula_sp_Lineage_I")
-# binomial depth and temp has lowest AIC
-AIC_testing(df_4, "Cyprinus_carpio")
-# binomial depth and temp has lowest AIC 
-AIC_testing(df_4, "Phoxinus_csikii")
-#binomial depth and temp has lowest AIC 
-AIC_testing(df_4, "Salmo_trutta")
-# binomial depth and temp lowest AIC
-
-# 8 models are run with binomial temp and depth
-
-```
-####Fig. 2
-Overview of all model predictions
-
-```{r, fig.show='hide'}
-# Load total predictions
-model_predictions <- readRDS("total_models/df_pred_all.rds")
-
-# check the species list
-model_predictions$species <- as.factor(model_predictions$species)
-levels(model_predictions$species)
-
-# read in the category table and rename categories
-species_across_lakes <-  read_xlsx("species_category_across_lakes.xlsx") |> 
-  select(-notes) |> 
-    mutate(category = ifelse(category == "native", "non_endemic_native", category)) |>
-  mutate(category = ifelse(category == "non_native_region", "non_endemic_native_and_translocated", category)) |> 
-  mutate(category2 = factor("all"))
-
-species_across_lakes$category <- as.factor(species_across_lakes$category)
-levels(species_across_lakes$category)
-
-#merge predictions with categories
-predictions_categories <- merge(model_predictions, species_across_lakes)
-
-# EVTL NEW PLOT 2
-# change category names for facet_wrap
-category_names <- c(
-  `endemic` = "endemic",
-  `endemic_and_translocated` = "translocated endemic",
-  `non_endemic_native` = "non-endemic native",
-  `non_endemic_native_and_translocated` = "translocated non-endemic native",
-  `non_native` = "non-native"
-)
-
-# colors for the plot
-mycolors1 <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
-                "endemic_and_translocated" = "#F2AD63", "non_endemic_native_and_translocated" = "#F2AD63")
-
-# Rescale the predictions to 0-1 
-predictions_categories_rescaled <- predictions_categories |>
-  group_by(species) |>
-  mutate(fit_rescaled = (fit - min(fit)) / (max(fit) - min(fit)))
-
-plot_pred <- predictions_categories_rescaled |>
-  ggplot(aes(temp, fit_rescaled, group = species, color = category)) +
-  geom_line() +
-  # how can I get the right color?
-   # geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), alpha = 0.3) +
-  theme_bw(base_size = 16) +
-  ylab("Abundance") +
-  xlab("Temperature") +
-  scale_color_manual(values = mycolors1, guide = NULL) +
-  theme(panel.grid.major = element_blank(), 
-        panel.grid.minor = element_blank(), 
-        legend.position = "bottom",
-        legend.title = element_blank())
-
-plot_category_predictions <- plot_pred + facet_grid(~category, labeller = as_labeller(category_names))
-
-plot_category_predictions
-
-# trying to group by category and color
-predictions_categories |>
-  ggplot(aes(temp, fit)) +
-  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit), fill = factor(category)), alpha = 0.3) +
-  geom_line(aes(color = factor(category))) +
-  theme_bw() +
-  facet_wrap(~species + category, scale = "free_y", labeller = labeller(species = label_wrap_gen(width = 10))) +
-  theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
-  scale_color_manual(values = mycolors1, aesthetics = c("color", "fill")) +
-  xlab("Temperature") +
-  ylab("Abundance")
-
-# OLD FIGURE 2
-
-all_predictions <- predictions_categories |>
-  ggplot(aes(temp, fit)) +
-  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit), fill = factor(species)), alpha = 0.3) +
-  geom_line(aes(color = factor(species))) +
-  theme_bw() +
-  facet_wrap(~species, scale = "free_y", labeller = labeller(species = label_wrap_gen(width = 10))) +
-  theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
-  scale_color_viridis(discrete = TRUE, guide = NULL, option = "G", aesthetics = c("color", "fill")) +
-  xlab("Temperature") +
-  ylab("Abundance")
-
-all_predictions
-# save the plot as TIFF
-# tiff(paste("total_models/plots/all_predictions_models.tiff", sep = ""), units="in", width = 12, height=8, res=300)
-# plot(all_predictions)
-# 
-# # Closing the graphical device
-# dev.off()
-
-```
-
-##Sensitivity testing
-NOT CHANGED 
-Testing if depth-driven temperature models can adequately fit abundance data compared to
-a temperature and depth model. 
-Models were compared in the eleven most abundant species.
-
-```{r}
-# Read in abundance and depth data of all species included for modelling
-df_models <- readRDS("df_models.rds")
-
-# find 11 most abundant  species
-species_list <- df_models |> 
-  filter(tot_abu > 300) |> 
-  distinct(Species) |> 
-  pull(Species)
-
-# most abundant species
- # "Alburnus_alburnus", "Coregonus_sp_albeli", "Coregonus_sp","Gasterosteus_aculeatus",
- # "Gobio_gobio", "Leuciscus_leuciscus", "Perca_fluviatilis", "Rutilus_rutilus", 
- # "Sander_lucioperca", "Scardinius_erythrophthalmus","Gymnocephalus_cernua"
-
-# filter df into the most abundant species and select the columns of interest
-
-df_sensitivity <- df_models |> 
-  filter(tot_abu > 300) |> 
-  select(Lake, Species, Protocol, Depth_sample, Abundance, Presence, mean_last_7days) |> 
-  #there are some NAs in depth -> drop them
-  drop_na(Depth_sample, mean_last_7days)
-
-df_sensitivity$fLake <- as.factor(df_sensitivity$Lake)
-df_sensitivity$fProtocol <- as.factor(df_sensitivity$Protocol)
-
-# Loop through the eleven species and fit temp and temp-depth model
-
-species_list <- df_sensitivity |> 
-  distinct(Species) |> 
-  pull(Species)
-
-# loop through the species, model both GAMs, check DHARMa and predictions
-for (i in species_list){
-
-  data <- df_sensitivity |> 
-    filter(Species == i)
-
-  #Model fit for temp and depth
-  model1 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(Depth_sample, k = 3)
-                + s(fLake, bs = 're')
-                +  s(fProtocol, bs = 're'), family = ziP())
-  # Model fit for temp
-  model2 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3)
-                + s(fLake, bs = 're')
-                +  s(fProtocol, bs = 're'), family = ziP())
-  # validate model predictions for temp and depth model with dharma:
-  # prepare residuals
-  simulationOutput <- simulateResiduals(fittedModel = model1, plot = F)
-  # tiff_filename <- paste("total_models/gam_check/temp_depth_check_", i, ".tiff", sep = "")
-  # tiff(tiff_filename, width = 800, height = 600)
-  print(plot(simulationOutput))
-  # dev.off()
-  # Plotting standardized residuals against predictors
-  # tiff_file_2 <- paste("total_models/gam_check/temp_depth_predictor_", i, ".tiff", sep = "")
-  # tiff(tiff_file_2, width = 800, height = 600)
-  print(plotResiduals(model1, data$mean_last_7days, xlab = "temp", main=NULL))
-  # dev.off()
-
-  # for predictions with random effects in the GAMs, we select a random value
-  # of the level of random effects and exclude the random effects in the prediction
-
-  # lakes where the species is present
-  unique_lakes <- distinct(data, Lake) |>
-    pull()
-
-  # select random lake for prediction
-  random_lake <- sample(unique_lakes, 1)
-
-  # for temp-depth model only predict for depth where the species is present
-  presence_depth <- data |> 
-    filter(Presence == 1) |>
-    distinct(Depth_sample) |> 
-    pull(Depth_sample)
-
-  # take randomly 5 depth values for prediction
-  random_depth <- sort(sample(presence_depth, 5))
-
-  # Convert sorted values to factors with original order
-  factor_random_depth <- factor(random_depth, levels = random_depth, ordered = TRUE)
-
-  # grid for each of the models
-
-  # temp-depth model predicted with random depths where the species is present
- grid1 <- expand.grid(mean_last_7days = seq(
-    from = min(data$mean_last_7days, na.rm = TRUE),
-    to = max(data$mean_last_7days, na.rm = TRUE), length = 200),
-    fProtocol = factor("VERT"), fLake = factor(random_lake), 
-    Depth_sample = factor_random_depth)
-
-  grid2 <- expand.grid(mean_last_7days = seq(
-    from = min(data$mean_last_7days, na.rm = TRUE),
-    to = max(data$mean_last_7days, na.rm = TRUE), length = 1000),
-    fProtocol = factor("VERT"), fLake = factor(random_lake))
-
-  # print relevant stats
-  print(paste(i))
-  print(glance(model1))
-  print(glance(model2))
-
-  # manually predict both models
-  model_prediction1 <- predict.gam(model1, newdata = grid1,
-                                  exclude = c("s(fProtocol)", "s(fLake)"), #exclude the re
-                                  term = "s(mean_last_7days)", type = "response",             
-                                  se.fit = TRUE)
-
-  model_bind1 <- cbind(grid1, as.data.frame(model_prediction1)) |>
-    mutate(model = factor("temp_and_depth"))
-
-# predict model2
-  model_prediction2 <- predict.gam(model2, newdata = grid2,
-                                   exclude = c("s(fProtocol)", "s(fLake)"),
-                                   type = "response", se.fit = TRUE)
-
-  model_bind2 <- cbind(grid2, as.data.frame(model_prediction2)) |>
-    mutate(model = factor("temp"))
-
-  df_pred <- bind_rows(model_bind1, model_bind2) |>
-    rename(temp = mean_last_7days) |>
-    select(-fProtocol, -fLake) |>
-    mutate(Depth_sample = factor(ifelse(is.na(Depth_sample), "temp", as.character(Depth_sample))))
-
-  # Extract numeric depth levels
-  numeric_levels <- levels(df_pred$Depth_sample)[!levels(df_pred$Depth_sample) %in% "temp"]
-
-  # Sort numeric depth levels
-  sorted_numeric_levels <- sort(as.numeric(numeric_levels))
-
-  # Combine sorted numeric levels with "temp" level
-  sorted_levels <- c(as.character(sorted_numeric_levels), "temp")
-
-  # Reorder levels
-  df_pred$Depth_sample <- factor(df_pred$Depth_sample, levels = sorted_levels)
-
-  # Check the levels to confirm they are ordered from lowest to highest
-  levels(df_pred$Depth_sample)
-
-  # rescale predictions of both models
-  df_pred_rescaled <- df_pred |>
-    group_by(model) |>
-    mutate(fit_rescaled = (fit - min(fit)) / (max(fit) - min(fit))) |>
-    mutate(se_fit_rescaled = (se.fit - min(se.fit)) / (max(se.fit) - min(se.fit))) |> 
-    mutate(species = factor(i))
-
-  # save predictions of both models in a rds
-   saveRDS(df_pred_rescaled, paste0("total_models/sensitivity/rds/sensitivity_predictions_", i, ".rds"))
-
-}
-
-# prepare total predictions of all models
-sensitivity_predictions <- list.files(path = "total_models/sensitivity/rds", pattern = ".rds", full.names = TRUE) |>
-  map_dfr(readRDS)
-
-# save total predictions as RDS
-saveRDS(sensitivity_predictions, "total_models/sensitivity_predictions.rds")
-
-```
-
-###Fig. S1
-
-```{r}
-# read in the sensitivity testing data
-sensitivity_predictions <- readRDS("total_models/sensitivity_predictions.rds")
-
-levels(sensitivity_predictions$species)
-
-# Plot both model predictions together per species
- figure_s1 <- sensitivity_predictions |>
-    ggplot(aes(temp, fit_rescaled, color = model, fill = model)) +
-    geom_line() +
-    geom_ribbon(aes(ymin = (fit_rescaled - se.fit), ymax = (fit_rescaled + se.fit)), color = NA, alpha = 0.1) +
-    theme_bw() +
-    # labs(title = paste("Sensitivity Test", i)) +
-    scale_color_manual(values = c("#08306B", "#D73027"), aesthetics = c("color", "fill")) +
-    facet_wrap(~species, scale = "free") +
-    ylab("Abundance") +
-    xlab("Temperature") +
-   theme(panel.grid.major = element_blank(), 
-        panel.grid.minor = element_blank(), 
-        legend.position = "right",
-        legend.title = element_blank())
-
-  figure_s1
-  # save as TIFF
-# tiff(paste("total_models/plots/sensitivity_overview.tiff", sep = ""), units="in", width=11, height=7, res=300)
-# 
-# plot(figure_s1)
-# 
-# dev.off()
-
-```
-##Derivatives calculation
-
-We calculated derivatives for each species per lake.
-
-```{r}
-
-# Derivatives() will ignore any random effect smooths it encounters in object.
-
-# Table of pecies that were recorded with electro, fishbase or eawag, following Table 16 in Projet Lac Synthesis Report. 
-# Plus the minimum and maximum temp in each lake -> needed for the new data for 
-# calculating derivatives
-
-species_lake <- read_xlsx("species_occurrences_lakes.xlsx") 
-str(species_lake)
-
-species_lake$fLake <- as.factor(species_lake$Lake)
-species_lake$fProtocol <- as.factor(species_lake$Protocol)
-species_lake$Species <- as.factor(species_lake$Species)
-
-#Loading all four dfs to calculate derivatives
-
-df_binomial_gam <- readRDS("data_frame_models/df_binomial_gam")
-df_binomial_gam$fLake <- as.factor(df_binomial_gam$Lake)
-df_binomial_gam$fProtocol <- as.factor(df_binomial_gam$Protocol)
-
-df_abundance_gam <- readRDS("data_frame_models/df_abundance_gam")
-df_abundance_gam$fProtocol <- as.factor(df_abundance_gam$Protocol)
-df_abundance_gam$fLake <- as.factor(df_abundance_gam$Lake)
-
-df_binomial_re <- readRDS("data_frame_models/df_binomial_re")
-df_binomial_re$fLake <- as.factor(df_binomial_re$Lake)
-df_binomial_re$fProtocol <- as.factor(df_binomial_re$Protocol)
-
-df_abundance_re <- readRDS("data_frame_models/df_abundance_re")
-df_abundance_re$fLake <- as.factor(df_abundance_re$Lake)
-df_abundance_re$fProtocol <- as.factor(df_abundance_re$Protocol)
-
-# separating the dfs to separate model types
-depth_binomial_gam <- df_binomial_gam |> 
-  filter(tot_abu > 20)
-
-temp_binomial_gam <- df_binomial_gam |> 
-  filter(tot_abu < 20)
-
-depth_abundance_gam <- df_abundance_gam |> 
-  filter(tot_abu > 20)
-
-temp_abundance_gam <- df_abundance_gam |> 
-  filter(tot_abu < 20)
-
-depth_binomial_re <- df_binomial_re |> 
-  filter(tot_abu > 20)
-
-temp_binomial_re <- df_binomial_re |> 
-  filter(tot_abu < 20)
-
-depth_abundance_re <- df_abundance_re |> 
-  filter(tot_abu > 20)
-
-temp_abundance_re <- df_abundance_re |> 
-  filter(tot_abu < 20)
-
-```
-
-###Depth and temp models
-```{r}
-# DEPTH AND TEMPERATURE
-# UNCLEAR HOW THIS CAN BE DONE
-# Model 1: Binomial data without random intercepts
-
-species_list <- depth_binomial_gam |>
-  distinct(Species) |>
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-i <- "Salmo_sp_Blackspot"
-j <- "Poschiavo"
-# loop through the list of species
-for (i in species_list) {
-  data <- depth_binomial_gam |>
-    filter(Species == i)
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
-                           s(Depth_sample, k = 3) + s(fProtocol, bs = 're'),
-                         family = binomial)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  # loop through all lakes where the species is present, lake list is based on the information from
-  # Projet Lac
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    # using minimum and maximum temp in the lake for derivatives
-     newdata <- tibble(mean_last_7days = seq(from = min(data_lake$temp, na.rm = TRUE),
-                                             to = max(data_lake$temp, na.rm = TRUE)
-                                             ,length = 200), Depth_sample =
-                         mean(data$Depth_sample, na.rm = TRUE),
-                       fProtocol = factor("VERT"))
-
-     # calculate derivatives
-     # derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-     #   mutate(fLake = factor(j)) |>
-     #   mutate(species = factor(i))
-     #   rename(temp = data)
-     #
-
-       test_derivatives <- derivative_samples(gam_output[[i]]
-                                              , focal = "mean_last_7days", data = newdata)
-
-       test_derivatives |>
-         filter(.draw <= 200) |>
-         ggplot(aes(x = mean_last_7days, y = .derivative, group = .draw)) +
-         geom_line(alpha = 0.5)
-
-       derivatives <- test_derivatives |> 
-         filter(.draw == 500) |> 
-         mutate(fLake = factor(j)) |>
-         mutate(species = factor(i))
-
-       derivatives |> 
-         ggplot(aes(mean_last_7days, .derivative)) +
-         geom_line()
-
-     saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-  }
-}
-
-# needs to be discussed
-
-# nothing changed starting from here
-############################################################################3##3
-# Model 2.1: Abundance data and no random intercept for lake -> but binomial
-# Coregonus_sp_benthic_profundal did not run with ZIP and was thus
-# run with binomial.
-
-species_list <- depth_abundance_gam |> 
-  # binomial one
-  filter(Species == "Coregonus_sp_benthic_profundal") |>
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_gam |> 
-    filter(Species == i)
-
-  # binomial family instead of ZIP
-  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) +
-                           s(fProtocol, bs = 're'), family = binomial())
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-  }
-}
-
-# Model 2.2: Abundance data and no random intercept for lake -> ZIP
-# all other species in this group run with ZIP
-
-species_list <- df_abundance_gam |>
-  filter(!Species == "Coregonus_sp_benthic_profundal") |>
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_gam |> 
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'),
-                         family = ziP())
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-  }
-}
-
-# Model 3: Binomial data and random intercept for lake
-
-species_list <- df_binomial_re |> 
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_binomial_re |> 
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = "re")
-                         + s(fProtocol, bs = 're'), family = binomial)
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  unique_lakes <- distinct(data, Lake) |> 
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    # take all lakes where the species is present and take random one for 
-    # the data included in derivatives()
-    random_lake <- sample(unique_lakes, 1)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), lenqh = 200),
-      fProtocol = factor("VERT"), fLake = factor(random_lake))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-
-  }
-}
-
-# Model 4.1: Abundance data and random intercept for lake, simplified models
-# multiple species in this group were simplified and run with binomial
-
-species_list <- df_abundance_re |>
-  # simplified models
-  filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
-                                "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta",
-                        "Lepomis_gibbosus")) |>
-  distinct(Species) |>
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_re |>
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
-                         +  s(fProtocol, bs = 're'), family = binomial)
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  unique_lakes <- distinct(data, Lake) |>
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    random_lake <- sample(unique_lakes, 1)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fLake = factor(random_lake), fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-
-  }
-}
-
-# Model 4.1: Abundance data and random intercept for lake, complex models
-
-species_list <- df_abundance_re |> 
-  # exclusion of simplified models
-  filter(!Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
-                        "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta", 
-                        "Lepomis_gibbosus")) |>
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_re |> 
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
-                         +  s(fProtocol, bs = 're'), family = ziP())
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  unique_lakes <- distinct(data, Lake) |> 
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    random_lake <- sample(unique_lakes, 1)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fLake = factor(random_lake), fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-
-  }
-}
-
-# save all the derivatives in one file
-
-df_deriv <- list.files(path = "total_models/derivatives", pattern = ".rds", full.names = TRUE) |> 
-  map_dfr(readRDS)
-
-saveRDS(df_deriv, "total_models/df_deriv_all.rds")
-
-# save total derivatives as RDS (no exclusions)
-# saveRDS(df_deriv, "total_models/df_deriv_all_no_excl.rds")
-
-```
-
-###Temp models
-```{r}
-
-# Now four different loops are run to calculate derivatives of each species 
-# in each lake
-
-# Model 1: Binomial data without random intercepts
-
-species_list <- df_binomial_gam |>
-  distinct(Species) |>
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-# loop through the list of species
-for (i in species_list) {
-  data <- df_binomial_gam |>
-    filter(Species == i)
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'), family = binomial)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  # loop through all lakes where the species is present, lake list is based on the information from
-  # Projet Lac
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    # using minimum and maximum temp in the lake for derivatives
-     newdata <- tibble(mean_last_7days = seq(from = min(data_lake$temp, na.rm = TRUE),
-                                             to = max(data_lake$temp, na.rm = TRUE), length = 200),
-                       fProtocol = factor("VERT"))
-
-     # calculate derivatives
-     derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-       mutate(fLake = factor(j)) |>
-       mutate(species = factor(i)) |>
-       rename(temp = data)
-
-     saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-  }
-}
-
-# Model 2.1: Abundance data and no random intercept for lake -> but binomial
-# Coregonus_sp_benthic_profundal did not run with ZIP and was thus
-# run with binomial.
-
-species_list <- df_abundance_gam |> 
-  # binomial one
-  filter(Species == "Coregonus_sp_benthic_profundal") |>
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_gam |> 
-    filter(Species == i)
-
-  # binomial family instead of ZIP
-  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) +
-                           s(fProtocol, bs = 're'), family = binomial())
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-  }
-}
-
-# Model 2.2: Abundance data and no random intercept for lake -> ZIP
-# all other species in this group run with ZIP
-
-species_list <- df_abundance_gam |>
-  filter(!Species == "Coregonus_sp_benthic_profundal") |>
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_gam |> 
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'),
-                         family = ziP())
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-  }
-}
-
-# Model 3: Binomial data and random intercept for lake
-
-species_list <- df_binomial_re |> 
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_binomial_re |> 
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = "re")
-                         + s(fProtocol, bs = 're'), family = binomial)
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  unique_lakes <- distinct(data, Lake) |> 
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    # take all lakes where the species is present and take random one for 
-    # the data included in derivatives()
-    random_lake <- sample(unique_lakes, 1)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), lenqh = 200),
-      fProtocol = factor("VERT"), fLake = factor(random_lake))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-
-  }
-}
-
-# Model 4.1: Abundance data and random intercept for lake, simplified models
-# multiple species in this group were simplified and run with binomial
-
-species_list <- df_abundance_re |>
-  # simplified models
-  filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
-                                "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta",
-                        "Lepomis_gibbosus")) |>
-  distinct(Species) |>
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_re |>
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
-                         +  s(fProtocol, bs = 're'), family = binomial)
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  unique_lakes <- distinct(data, Lake) |>
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    random_lake <- sample(unique_lakes, 1)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fLake = factor(random_lake), fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-
-  }
-}
-
-# Model 4.1: Abundance data and random intercept for lake, complex models
-
-species_list <- df_abundance_re |> 
-  # exclusion of simplified models
-  filter(!Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
-                        "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta", 
-                        "Lepomis_gibbosus")) |>
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_re |> 
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
-                         +  s(fProtocol, bs = 're'), family = ziP())
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  unique_lakes <- distinct(data, Lake) |> 
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    random_lake <- sample(unique_lakes, 1)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fLake = factor(random_lake), fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-
-  }
-}
-
-# save all the derivatives in one file
-
-df_deriv <- list.files(path = "total_models/derivatives", pattern = ".rds", full.names = TRUE) |> 
-  map_dfr(readRDS)
-
-saveRDS(df_deriv, "total_models/df_deriv_all.rds")
-
-# save total derivatives as RDS (no exclusions)
-# saveRDS(df_deriv, "total_models/df_deriv_all_no_excl.rds")
-
-```
-
-###Fig. S2 
-Overview of derivatives in each lake community
-
-```{r}
-# Read in derivative data
-all_models_derivatives <- readRDS("total_models/df_deriv_all.rds")
-all_derivatives <- as_tibble(all_models_derivatives)
-
-str(all_derivatives)
-
-# Highlighting species with interesting derivatives: 
-species_group <- all_derivatives |> 
-  filter(species %in% c("Phoxinus_csikii", "Barbatula_sp_Lineage_II", "Cottus_sp_Po_profundal")) |> 
-  distinct(species) |> 
-  pull(species)
-
-# Plot S2: derivatives included in each lake community
-overview_derivatives <- all_derivatives |>
-  arrange(fLake) |> 
-  ggplot(aes(temp, derivative, color = species)) +
-  geom_line() +
-  facet_wrap(~fLake) +
-  theme_bw(base_size = 16) +
-  scale_color_viridis(discrete = TRUE, breaks = species_group, option = "G") +
-  xlab("Temperature") +
-  ylab("Derivatives") +
-  theme(panel.grid.major = element_blank(), 
-        panel.grid.minor = element_blank(), 
-        legend.position = "right",
-        legend.title = element_blank()) +
-  geom_hline(yintercept = 0, color = "#7A7A7A")
-
-overview_derivatives
-
-# save as TIFF
-# tiff(paste("total_models/plots/overview_derivatives_lake.tiff", sep = ""), units="in", width=11, height=7, res=300)
-# 
-# plot(overview_derivatives)
-# 
-# dev.off()
-```
-
-###Fig. S3
-
-```{r}
-# Data of the derivatives before simplifying multiple models
-# without changes the 5 species were not run with simplified models. 
-all_models_derivatives_no_excl <- readRDS("total_models/df_deriv_all_no_excl.rds")
-df_deriv_no_excl <- as_tibble(all_models_derivatives_no_excl)
-
-# Plot Figure S3a
-#calculating maximum and mean derivatives for each species per lake
-
-df_quantiles <- df_deriv_no_excl |> 
-  group_by(species, fLake) |> 
-  mutate(max_derivative = max(derivative)) |> 
-  mutate(mean_derivative = mean(derivative)) |> 
-  ungroup()
-
-# calculate quantiles of derivatives
-transform(quantile(df_quantiles$derivative,
-                   c(0,0.01,0.05,0.1,0.25,0.5,0.75,0.85,0.9,0.95,0.99,1)))
-
-# add column with the category of quantile each derivative is in
-df_quantiles$groups <- cut(df_quantiles$derivative,              
-                     breaks = c(-1.579943e+06, -2.342866e+00, -8.822087e-01,
-                                -4.543748e-01, 3.454218e-04, 2.485142e-01,
-                                5.527792e-01, 2.911987e+00, 2.018455e+01, 1.280898e+02,
-                                1.231212e+07, 1.406101e+07),
-                     labels = c("0-1%", "1-5%", "5-10%", "10-25%", "25-50%", "50-75%", "75-85%", "85-90%", "90-95%",
-                                "95-99%", "99-100%"))
-
-# Plot S3a: histogram of species' mean derivatives per lake
-plot_percentiles <- df_quantiles |> 
-  # Salmo trutta derivative is too large to be plotted
-  filter(species != "Salmo_trutta") |> 
-  ggplot(aes(x = mean_derivative, fill = factor(species))) + 
-  geom_histogram(aes(y = after_stat(count / sum(count))), binwidth = 1.5) +
-  scale_y_continuous(labels = scales::percent) +
-  scale_fill_manual(values = c("Cyprinus_carpio" = "#80CDC1",
-                                "Alburnus_arborella" = "#8C510A",
-                                "Barbatula_sp_Lineage_I"="#DFC27D",
-                               "Phoxinus_csikii" = "#01665E", "Salmo_trutta" = "#F46D43"))
-
-plot_percentiles
-
-# Plot S3a add lines for the percentiles
-plot_percentiles_1 <- plot_percentiles +
-  geom_vline(xintercept = 19.9, color = "red") +
-  annotate("text", x= 18, y = 0.12, label="90th percentile", angle=90) +
-  geom_vline(xintercept = 2.911987e+00, color = "red") +
-  annotate("text", x= 1.5, y = 0.12, label="85th percentile", angle=90) +
-  theme_classic(base_size = 13) +
-  xlab("Mean species derivative per lake") +
-  ylab("Percentage") +
- guides(fill = guide_legend(
-    title = "Species")) +
-  ylim(0, 0.2)
-
-plot_percentiles_1
-
-# save as TIFF
-# tiff(paste("total_models/plots/plot_deriv_percentiles_1.tiff", sep = ""), units="in", width=12, height=6, res=300)
-# 
-# plot(plot_percentiles_1)
-# 
-# dev.off()
-
-# above 90% = 2.018455e+01
-df_quantiles |> 
-  filter(derivative > 2.018455e+01) |> 
-  distinct(species)
-
-# 5 species above 90% -> those will be simplified to binomial 
-# 1 Alburnus_arborella    
-# 2 Barbatula_sp_Lineage_I
-# 3 Cyprinus_carpio       
-# 4 Phoxinus_csikii       
-# 5 Salmo_trutta
-
-# Plotting Figure S3b
-
-# just keeping mean and maximum derivative values for each lake and species
-data_quantiles <- df_quantiles |> 
-  distinct(mean_derivative, max_derivative, fLake, species)
-
-# Breaking mean derivatives in the same percentiles as before
-data_quantiles$percentiles <- cut(data_quantiles$mean_derivative,              
-                     breaks = c(-1.579943e+06, -2.342866e+00, -8.822087e-01,
-                                -4.543748e-01, 3.454218e-04, 2.485142e-01,
-                                5.527792e-01, 2.911987e+00, 2.018455e+01, 1.280898e+02,
-                                1.231212e+07, 1.406101e+07),
-                     labels = c("0-1%", "1-5%", "5-10%", "10-25%", "25-50%", "50-75%", "75-85%", "85-90%", "90-95%",
-                                "95-99%", "99-100%"))
-
-# Plot S3b tile plot of mean derivatives per species in each lake
-mean_deriv_plot <- data_quantiles |> 
-  ggplot(aes(fLake, y = fct_reorder(species, mean_derivative), fill= percentiles)) + 
-  geom_tile() +
-  scale_fill_manual(breaks = levels(data_quantiles$percentiles),
-                    values = rev(brewer.pal(11, "BrBG")))
-
-derivative_percentiles_2 <- mean_deriv_plot +
-  guides(fill = guide_legend(title = "percentiles (mean derivative)", reverse = TRUE)) +
-  xlab("") +
-  ylab("") +
-  theme_classic(base_size = 16)
-
-derivative_percentiles_2
-
-# save plot as TIFF
-tiff(paste("total_models/plots/plot_deriv_percentiles_2.tiff", sep = ""), units="in", width=19, height=12, res=300)
-
-plot(derivative_percentiles_2)
-
-dev.off()
-
-```
-
-##Calculating thermal response diversity
-
-```{r}
-# Function provided by Ross et al. "How to measure response diversity" used to calculate
-# response diversity
-
-# read in function
-source(here("functions.R"))
-
-# load derivatives data
-all_models_derivatives <- readRDS("total_models/df_deriv_all.rds")
-
-all_deriv <- as_tibble(all_models_derivatives)
-
-str(all_deriv)
-
-levels(all_deriv$species)
-
-#Calculating response diversity for each lake community
-
-# Lepomis is excluded from the analysis
-all_deriv <- all_deriv |>
-  filter(!species %in% c("Lepomis_gibbosus"))
-
-all_deriv$fLake <- as.character(all_deriv $fLake)
-
-lakes_list <- all_deriv |> 
-  distinct(fLake) |> 
-  pull(fLake)
-
-str(lakes_list)
-
-lakes_list <- sort(lakes_list)
-
-species_overview <- tibble()
-
-all_deriv$species <- as.character(all_deriv$species)
-all_deriv$species <- as.factor(all_deriv$species)
-
-str(all_deriv )
-
-# Loop to get response diversity measures for each lake 
-# loop through list of lakes and calculate response diversity in the lake communities
-for (i in lakes_list){
-
-  data <- all_deriv |>
-    select(temp, fLake, derivative, species) |> 
-    filter(fLake == i)
-
-  df_resp_div <- data |>
-    pivot_wider(
-      names_from = species,
-      values_from = derivative)
-
-  # apply the function from Ross et al. 
-  # Dissimilarity
-  df_resp_div$rdiv <- apply(df_resp_div[,-(1:2), drop = FALSE], 1, resp_div, sign_sens = F)
-  # Divergence
-  df_resp_div$sign <- apply(df_resp_div[,-(1:2), drop = FALSE], 1, resp_div, sign_sens = T)
-  # Median Dissimilarity
-  df_resp_div$Med <- median(df_resp_div$rdiv)
-
-  # save response diversity metrics per lake
-  saveRDS(df_resp_div, paste0("total_models/lakes_all_models/df_resp_div_", i, ".rds"))
-
-}
-
-# save the response diversity per lake into one file
-resp_div_no_excl <- list.files(path = "total_models/lakes_all_models", pattern = ".rds", full.names = TRUE) |>
-  map_dfr(readRDS) |>
-  relocate(rdiv, Med, sign, .after = temp)
-
-saveRDS(resp_div_no_excl,"total_models/resp_div_all.rds")
-```
-
 ##Drivers of thermal response diversity

### 2024-07-05 data_preparation.Rmd
data preparation

--- 2024-07-03 manuscript_analysis.Rmd
@@
 ---
-title: "manuscript_analysis"
+title: "data preparation"
 author: "Sophie Moreau"
-date: "2024-07-03"
+date: "2024-07-05"
 output: html_document
@@
 ```
-
-#2. Data Analysis
-##General Additive Models
-Running the predictions for all GAMs with the smaller data frames to reduce running 
-time.
-
-```{r}
-# packages relevant for data analysis
-library(tidyverse)
-library(here)
-library(readr)
-library(readxl)
-library(gratia)
-library(mgcv)
-library(forcats)
-library(ggpubr)
-library(RColorBrewer)
-library(broom)
-library(grid)
-library(DHARMa)
-library(mgcViz)
-library(gridExtra)
-library(grid)
-
-# function for the GAMs (old)
-# source(here("function_GAM.R"))
-
-# new function for temp and depth models 
-source(here("function_temp_depth.R"))
-```
-
-###Predictions of GAMs
-
-```{r}
-# read in the dataframes
-df_1 <- readRDS("data_frame_models/df_binomial_gam")
-df_2 <- readRDS("data_frame_models/df_abundance_gam")
-df_3 <- readRDS("data_frame_models/df_binomial_re")
-df_4 <- readRDS("data_frame_models/df_abundance_re")
-
-# use own function predictions() from function_GAM.R to predict species' models
-# across all lakes
-
-# predictions take VERY LONG! and two folders are needed to store the 
-# DHARMa and prediction output inside: 
-# total_models/gam_check
-# total_models/predictions
-
-#df1 and df2 are the fastest ones
-# predictions() can also be used directly with df_models -> takes longer
-# BECAUSE OF TYPO: RE RUN TEMP + DEPTH models from df1-3, df4 still needs to be run & stats overview too
-
-# DO NOT RUN NOW, RUN LATER!
-predictions_depth_temp(df_1)
-predictions_depth_temp(df_2)
-predictions_depth_temp(df_3)
-predictions_depth_temp(df_4)
-
-# statistics of the gamms
-total_models <- bind_rows(df_1, df_2, df_3, df_4)
-
-stats_gamms(total_models)
-
-#  compile data frame with all predictions from the saved species' model predictions
-
-df_predictions_all <- list.files(path = "total_models/predictions", pattern = ".rds", full.names = TRUE) |>
-  map_dfr(readRDS)
-
-# save total predictions as RDS
-saveRDS(df_predictions_all, "total_models/df_pred_all.rds")
-
-```
-
-###AIC comparisons in special cases
-
-```{r}
-
-AIC_testing <- function(data2, i){
-
-  data <- data2 |> 
-    filter(Species == i)
-
-  data$fProtocol <- as.factor(data$Protocol)
-  data$fLake <- as.factor(data$Lake)
-
-  model_0 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
-                   s(Depth_sample, k = 3)
-                            + s(fLake, bs = 're')
-                            +  s(fProtocol, bs = 're'), family = ziP())
-  print(glance(model_0))
-# new models
-
-model1 <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) + s(Depth_sample, k = 3)
-                            + s(fLake, bs = 're')
-                            +  s(fProtocol, bs = 're'), family = binomial)
-print(glance(model1))
-
-model2 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
-                            + s(fLake, bs = 're')
-                            +  s(fProtocol, bs = 're'), family = ziP())
-print(glance(model2))
-
-}
-
-AIC_testing(df_4, "Lepomis_gibbosus")
-
-# in Alosa agone: model_0 not running, smaller AIC for binomial depth + temp model -> will be modelled as binomial depth + temp
-# same for Coregonus_sp_large_pelagic
-# lepomis gibbosus also has the smallest AIC for binomial depth + temp model
-
-# also testing the other special cases:
- # "Alburnus_arborella", "Barbatula_sp_Lineage_I","Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta"
-
-AIC_testing(df_4, "Alburnus_arborella") 
-# not running with model0, Binomial with depth and temp best
-AIC_testing(df_4, "Barbatula_sp_Lineage_I")
-# binomial depth and temp has lowest AIC
-AIC_testing(df_4, "Cyprinus_carpio")
-# binomial depth and temp has lowest AIC 
-AIC_testing(df_4, "Phoxinus_csikii")
-#binomial depth and temp has lowest AIC 
-AIC_testing(df_4, "Salmo_trutta")
-# binomial depth and temp lowest AIC
-
-# 8 models are run with binomial temp and depth
-
-```
-####Fig. 2
-Overview of all model predictions
-
-```{r, fig.show='hide'}
-# Load total predictions
-model_predictions <- readRDS("total_models/df_pred_all.rds")
-
-# check the species list
-model_predictions$species <- as.factor(model_predictions$species)
-levels(model_predictions$species)
-
-# read in the category table and rename categories
-species_across_lakes <-  read_xlsx("species_category_across_lakes.xlsx") |> 
-  select(-notes) |> 
-    mutate(category = ifelse(category == "native", "non_endemic_native", category)) |>
-  mutate(category = ifelse(category == "non_native_region", "non_endemic_native_and_translocated", category)) |> 
-  mutate(category2 = factor("all"))
-
-species_across_lakes$category <- as.factor(species_across_lakes$category)
-levels(species_across_lakes$category)
-
-#merge predictions with categories
-predictions_categories <- merge(model_predictions, species_across_lakes)
-
-# EVTL NEW PLOT 2
-# change category names for facet_wrap
-category_names <- c(
-  `endemic` = "endemic",
-  `endemic_and_translocated` = "translocated endemic",
-  `non_endemic_native` = "non-endemic native",
-  `non_endemic_native_and_translocated` = "translocated non-endemic native",
-  `non_native` = "non-native"
-)
-
-# colors for the plot
-mycolors1 <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
-                "endemic_and_translocated" = "#F2AD63", "non_endemic_native_and_translocated" = "#F2AD63")
-
-# Rescale the predictions to 0-1 
-predictions_categories_rescaled <- predictions_categories |>
-  group_by(species) |>
-  mutate(fit_rescaled = (fit - min(fit)) / (max(fit) - min(fit)))
-
-plot_pred <- predictions_categories_rescaled |>
-  ggplot(aes(temp, fit_rescaled, group = species, color = category)) +
-  geom_line() +
-  # how can I get the right color?
-   # geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), alpha = 0.3) +
-  theme_bw(base_size = 16) +
-  ylab("Abundance") +
-  xlab("Temperature") +
-  scale_color_manual(values = mycolors1, guide = NULL) +
-  theme(panel.grid.major = element_blank(), 
-        panel.grid.minor = element_blank(), 
-        legend.position = "bottom",
-        legend.title = element_blank())
-
-plot_category_predictions <- plot_pred + facet_grid(~category, labeller = as_labeller(category_names))
-
-plot_category_predictions
-
-# trying to group by category and color
-predictions_categories |>
-  ggplot(aes(temp, fit)) +
-  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit), fill = factor(category)), alpha = 0.3) +
-  geom_line(aes(color = factor(category))) +
-  theme_bw() +
-  facet_wrap(~species + category, scale = "free_y", labeller = labeller(species = label_wrap_gen(width = 10))) +
-  theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
-  scale_color_manual(values = mycolors1, aesthetics = c("color", "fill")) +
-  xlab("Temperature") +
-  ylab("Abundance")
-
-# OLD FIGURE 2
-
-all_predictions <- predictions_categories |>
-  ggplot(aes(temp, fit)) +
-  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit), fill = factor(species)), alpha = 0.3) +
-  geom_line(aes(color = factor(species))) +
-  theme_bw() +
-  facet_wrap(~species, scale = "free_y", labeller = labeller(species = label_wrap_gen(width = 10))) +
-  theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
-  scale_color_viridis(discrete = TRUE, guide = NULL, option = "G", aesthetics = c("color", "fill")) +
-  xlab("Temperature") +
-  ylab("Abundance")
-
-all_predictions
-# save the plot as TIFF
-# tiff(paste("total_models/plots/all_predictions_models.tiff", sep = ""), units="in", width = 12, height=8, res=300)
-# plot(all_predictions)
-# 
-# # Closing the graphical device
-# dev.off()
-
-```
-
-##Sensitivity testing
-NOT CHANGED 
-Testing if depth-driven temperature models can adequately fit abundance data compared to
-a temperature and depth model. 
-Models were compared in the eleven most abundant species.
-
-```{r}
-# Read in abundance and depth data of all species included for modelling
-df_models <- readRDS("df_models.rds")
-
-# find 11 most abundant  species
-species_list <- df_models |> 
-  filter(tot_abu > 300) |> 
-  distinct(Species) |> 
-  pull(Species)
-
-# most abundant species
- # "Alburnus_alburnus", "Coregonus_sp_albeli", "Coregonus_sp","Gasterosteus_aculeatus",
- # "Gobio_gobio", "Leuciscus_leuciscus", "Perca_fluviatilis", "Rutilus_rutilus", 
- # "Sander_lucioperca", "Scardinius_erythrophthalmus","Gymnocephalus_cernua"
-
-# filter df into the most abundant species and select the columns of interest
-
-df_sensitivity <- df_models |> 
-  filter(tot_abu > 300) |> 
-  select(Lake, Species, Protocol, Depth_sample, Abundance, Presence, mean_last_7days) |> 
-  #there are some NAs in depth -> drop them
-  drop_na(Depth_sample, mean_last_7days)
-
-df_sensitivity$fLake <- as.factor(df_sensitivity$Lake)
-df_sensitivity$fProtocol <- as.factor(df_sensitivity$Protocol)
-
-# Loop through the eleven species and fit temp and temp-depth model
-
-species_list <- df_sensitivity |> 
-  distinct(Species) |> 
-  pull(Species)
-
-# loop through the species, model both GAMs, check DHARMa and predictions
-for (i in species_list){
-
-  data <- df_sensitivity |> 
-    filter(Species == i)
-
-  #Model fit for temp and depth
-  model1 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(Depth_sample, k = 3)
-                + s(fLake, bs = 're')
-                +  s(fProtocol, bs = 're'), family = ziP())
-  # Model fit for temp
-  model2 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3)
-                + s(fLake, bs = 're')
-                +  s(fProtocol, bs = 're'), family = ziP())
-  # validate model predictions for temp and depth model with dharma:
-  # prepare residuals
-  simulationOutput <- simulateResiduals(fittedModel = model1, plot = F)
-  # tiff_filename <- paste("total_models/gam_check/temp_depth_check_", i, ".tiff", sep = "")
-  # tiff(tiff_filename, width = 800, height = 600)
-  print(plot(simulationOutput))
-  # dev.off()
-  # Plotting standardized residuals against predictors
-  # tiff_file_2 <- paste("total_models/gam_check/temp_depth_predictor_", i, ".tiff", sep = "")
-  # tiff(tiff_file_2, width = 800, height = 600)
-  print(plotResiduals(model1, data$mean_last_7days, xlab = "temp", main=NULL))
-  # dev.off()
-
-  # for predictions with random effects in the GAMs, we select a random value
-  # of the level of random effects and exclude the random effects in the prediction
-
-  # lakes where the species is present
-  unique_lakes <- distinct(data, Lake) |>
-    pull()
-
-  # select random lake for prediction
-  random_lake <- sample(unique_lakes, 1)
-
-  # for temp-depth model only predict for depth where the species is present
-  presence_depth <- data |> 
-    filter(Presence == 1) |>
-    distinct(Depth_sample) |> 
-    pull(Depth_sample)
-
-  # take randomly 5 depth values for prediction
-  random_depth <- sort(sample(presence_depth, 5))
-
-  # Convert sorted values to factors with original order
-  factor_random_depth <- factor(random_depth, levels = random_depth, ordered = TRUE)
-
-  # grid for each of the models
-
-  # temp-depth model predicted with random depths where the species is present
- grid1 <- expand.grid(mean_last_7days = seq(
-    from = min(data$mean_last_7days, na.rm = TRUE),
-    to = max(data$mean_last_7days, na.rm = TRUE), length = 200),
-    fProtocol = factor("VERT"), fLake = factor(random_lake), 
-    Depth_sample = factor_random_depth)
-
-  grid2 <- expand.grid(mean_last_7days = seq(
-    from = min(data$mean_last_7days, na.rm = TRUE),
-    to = max(data$mean_last_7days, na.rm = TRUE), length = 1000),
-    fProtocol = factor("VERT"), fLake = factor(random_lake))
-
-  # print relevant stats
-  print(paste(i))
-  print(glance(model1))
-  print(glance(model2))
-
-  # manually predict both models
-  model_prediction1 <- predict.gam(model1, newdata = grid1,
-                                  exclude = c("s(fProtocol)", "s(fLake)"), #exclude the re
-                                  term = "s(mean_last_7days)", type = "response",             
-                                  se.fit = TRUE)
-
-  model_bind1 <- cbind(grid1, as.data.frame(model_prediction1)) |>
-    mutate(model = factor("temp_and_depth"))
-
-# predict model2
-  model_prediction2 <- predict.gam(model2, newdata = grid2,
-                                   exclude = c("s(fProtocol)", "s(fLake)"),
-                                   type = "response", se.fit = TRUE)
-
-  model_bind2 <- cbind(grid2, as.data.frame(model_prediction2)) |>
-    mutate(model = factor("temp"))
-
-  df_pred <- bind_rows(model_bind1, model_bind2) |>
-    rename(temp = mean_last_7days) |>
-    select(-fProtocol, -fLake) |>
-    mutate(Depth_sample = factor(ifelse(is.na(Depth_sample), "temp", as.character(Depth_sample))))
-
-  # Extract numeric depth levels
-  numeric_levels <- levels(df_pred$Depth_sample)[!levels(df_pred$Depth_sample) %in% "temp"]
-
-  # Sort numeric depth levels
-  sorted_numeric_levels <- sort(as.numeric(numeric_levels))
-
-  # Combine sorted numeric levels with "temp" level
-  sorted_levels <- c(as.character(sorted_numeric_levels), "temp")
-
-  # Reorder levels
-  df_pred$Depth_sample <- factor(df_pred$Depth_sample, levels = sorted_levels)
-
-  # Check the levels to confirm they are ordered from lowest to highest
-  levels(df_pred$Depth_sample)
-
-  # rescale predictions of both models
-  df_pred_rescaled <- df_pred |>
-    group_by(model) |>
-    mutate(fit_rescaled = (fit - min(fit)) / (max(fit) - min(fit))) |>
-    mutate(se_fit_rescaled = (se.fit - min(se.fit)) / (max(se.fit) - min(se.fit))) |> 
-    mutate(species = factor(i))
-
-  # save predictions of both models in a rds
-   saveRDS(df_pred_rescaled, paste0("total_models/sensitivity/rds/sensitivity_predictions_", i, ".rds"))
-
-}
-
-# prepare total predictions of all models
-sensitivity_predictions <- list.files(path = "total_models/sensitivity/rds", pattern = ".rds", full.names = TRUE) |>
-  map_dfr(readRDS)
-
-# save total predictions as RDS
-saveRDS(sensitivity_predictions, "total_models/sensitivity_predictions.rds")
-
-```
-
-###Fig. S1
-
-```{r}
-# read in the sensitivity testing data
-sensitivity_predictions <- readRDS("total_models/sensitivity_predictions.rds")
-
-levels(sensitivity_predictions$species)
-
-# Plot both model predictions together per species
- figure_s1 <- sensitivity_predictions |>
-    ggplot(aes(temp, fit_rescaled, color = model, fill = model)) +
-    geom_line() +
-    geom_ribbon(aes(ymin = (fit_rescaled - se.fit), ymax = (fit_rescaled + se.fit)), color = NA, alpha = 0.1) +
-    theme_bw() +
-    # labs(title = paste("Sensitivity Test", i)) +
-    scale_color_manual(values = c("#08306B", "#D73027"), aesthetics = c("color", "fill")) +
-    facet_wrap(~species, scale = "free") +
-    ylab("Abundance") +
-    xlab("Temperature") +
-   theme(panel.grid.major = element_blank(), 
-        panel.grid.minor = element_blank(), 
-        legend.position = "right",
-        legend.title = element_blank())
-
-  figure_s1
-  # save as TIFF
-# tiff(paste("total_models/plots/sensitivity_overview.tiff", sep = ""), units="in", width=11, height=7, res=300)
-# 
-# plot(figure_s1)
-# 
-# dev.off()
-
-```
-##Derivatives calculation
-
-We calculated derivatives for each species per lake.
-
-```{r}
-
-# Derivatives() will ignore any random effect smooths it encounters in object.
-
-# Table of pecies that were recorded with electro, fishbase or eawag, following Table 16 in Projet Lac Synthesis Report. 
-# Plus the minimum and maximum temp in each lake -> needed for the new data for 
-# calculating derivatives
-
-species_lake <- read_xlsx("species_occurrences_lakes.xlsx") 
-str(species_lake)
-
-species_lake$fLake <- as.factor(species_lake$Lake)
-species_lake$fProtocol <- as.factor(species_lake$Protocol)
-species_lake$Species <- as.factor(species_lake$Species)
-
-#Loading all four dfs to calculate derivatives
-
-df_binomial_gam <- readRDS("data_frame_models/df_binomial_gam")
-df_binomial_gam$fLake <- as.factor(df_binomial_gam$Lake)
-df_binomial_gam$fProtocol <- as.factor(df_binomial_gam$Protocol)
-
-df_abundance_gam <- readRDS("data_frame_models/df_abundance_gam")
-df_abundance_gam$fProtocol <- as.factor(df_abundance_gam$Protocol)
-df_abundance_gam$fLake <- as.factor(df_abundance_gam$Lake)
-
-df_binomial_re <- readRDS("data_frame_models/df_binomial_re")
-df_binomial_re$fLake <- as.factor(df_binomial_re$Lake)
-df_binomial_re$fProtocol <- as.factor(df_binomial_re$Protocol)
-
-df_abundance_re <- readRDS("data_frame_models/df_abundance_re")
-df_abundance_re$fLake <- as.factor(df_abundance_re$Lake)
-df_abundance_re$fProtocol <- as.factor(df_abundance_re$Protocol)
-
-# separating the dfs to separate model types
-depth_binomial_gam <- df_binomial_gam |> 
-  filter(tot_abu > 20)
-
-temp_binomial_gam <- df_binomial_gam |> 
-  filter(tot_abu < 20)
-
-depth_abundance_gam <- df_abundance_gam |> 
-  filter(tot_abu > 20)
-
-temp_abundance_gam <- df_abundance_gam |> 
-  filter(tot_abu < 20)
-
-depth_binomial_re <- df_binomial_re |> 
-  filter(tot_abu > 20)
-
-temp_binomial_re <- df_binomial_re |> 
-  filter(tot_abu < 20)
-
-depth_abundance_re <- df_abundance_re |> 
-  filter(tot_abu > 20)
-
-temp_abundance_re <- df_abundance_re |> 
-  filter(tot_abu < 20)
-
-```
-
-###Depth and temp models
-```{r}
-# DEPTH AND TEMPERATURE
-# UNCLEAR HOW THIS CAN BE DONE
-# Model 1: Binomial data without random intercepts
-
-species_list <- depth_binomial_gam |>
-  distinct(Species) |>
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-i <- "Salmo_sp_Blackspot"
-j <- "Poschiavo"
-# loop through the list of species
-for (i in species_list) {
-  data <- depth_binomial_gam |>
-    filter(Species == i)
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
-                           s(Depth_sample, k = 3) + s(fProtocol, bs = 're'),
-                         family = binomial)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  # loop through all lakes where the species is present, lake list is based on the information from
-  # Projet Lac
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    # using minimum and maximum temp in the lake for derivatives
-     newdata <- tibble(mean_last_7days = seq(from = min(data_lake$temp, na.rm = TRUE),
-                                             to = max(data_lake$temp, na.rm = TRUE)
-                                             ,length = 200), Depth_sample =
-                         mean(data$Depth_sample, na.rm = TRUE),
-                       fProtocol = factor("VERT"))
-
-     # calculate derivatives
-     # derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-     #   mutate(fLake = factor(j)) |>
-     #   mutate(species = factor(i))
-     #   rename(temp = data)
-     #
-
-       test_derivatives <- derivative_samples(gam_output[[i]]
-                                              , focal = "mean_last_7days", data = newdata)
-
-       test_derivatives |>
-         filter(.draw <= 200) |>
-         ggplot(aes(x = mean_last_7days, y = .derivative, group = .draw)) +
-         geom_line(alpha = 0.5)
-
-       derivatives <- test_derivatives |> 
-         filter(.draw == 500) |> 
-         mutate(fLake = factor(j)) |>
-         mutate(species = factor(i))
-
-       derivatives |> 
-         ggplot(aes(mean_last_7days, .derivative)) +
-         geom_line()
-
-     saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-  }
-}
-
-# needs to be discussed
-
-# nothing changed starting from here
-############################################################################3##3
-# Model 2.1: Abundance data and no random intercept for lake -> but binomial
-# Coregonus_sp_benthic_profundal did not run with ZIP and was thus
-# run with binomial.
-
-species_list <- depth_abundance_gam |> 
-  # binomial one
-  filter(Species == "Coregonus_sp_benthic_profundal") |>
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_gam |> 
-    filter(Species == i)
-
-  # binomial family instead of ZIP
-  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) +
-                           s(fProtocol, bs = 're'), family = binomial())
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-  }
-}
-
-# Model 2.2: Abundance data and no random intercept for lake -> ZIP
-# all other species in this group run with ZIP
-
-species_list <- df_abundance_gam |>
-  filter(!Species == "Coregonus_sp_benthic_profundal") |>
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_gam |> 
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'),
-                         family = ziP())
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-  }
-}
-
-# Model 3: Binomial data and random intercept for lake
-
-species_list <- df_binomial_re |> 
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_binomial_re |> 
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = "re")
-                         + s(fProtocol, bs = 're'), family = binomial)
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  unique_lakes <- distinct(data, Lake) |> 
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    # take all lakes where the species is present and take random one for 
-    # the data included in derivatives()
-    random_lake <- sample(unique_lakes, 1)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), lenqh = 200),
-      fProtocol = factor("VERT"), fLake = factor(random_lake))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-
-  }
-}
-
-# Model 4.1: Abundance data and random intercept for lake, simplified models
-# multiple species in this group were simplified and run with binomial
-
-species_list <- df_abundance_re |>
-  # simplified models
-  filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
-                                "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta",
-                        "Lepomis_gibbosus")) |>
-  distinct(Species) |>
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_re |>
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
-                         +  s(fProtocol, bs = 're'), family = binomial)
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  unique_lakes <- distinct(data, Lake) |>
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    random_lake <- sample(unique_lakes, 1)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fLake = factor(random_lake), fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-
-  }
-}
-
-# Model 4.1: Abundance data and random intercept for lake, complex models
-
-species_list <- df_abundance_re |> 
-  # exclusion of simplified models
-  filter(!Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
-                        "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta", 
-                        "Lepomis_gibbosus")) |>
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_re |> 
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
-                         +  s(fProtocol, bs = 're'), family = ziP())
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  unique_lakes <- distinct(data, Lake) |> 
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    random_lake <- sample(unique_lakes, 1)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fLake = factor(random_lake), fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-
-  }
-}
-
-# save all the derivatives in one file
-
-df_deriv <- list.files(path = "total_models/derivatives", pattern = ".rds", full.names = TRUE) |> 
-  map_dfr(readRDS)
-
-saveRDS(df_deriv, "total_models/df_deriv_all.rds")
-
-# save total derivatives as RDS (no exclusions)
-# saveRDS(df_deriv, "total_models/df_deriv_all_no_excl.rds")
-
-```
-
-###Temp models
-```{r}
-
-# Now four different loops are run to calculate derivatives of each species 
-# in each lake
-
-# Model 1: Binomial data without random intercepts
-
-species_list <- df_binomial_gam |>
-  distinct(Species) |>
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-# loop through the list of species
-for (i in species_list) {
-  data <- df_binomial_gam |>
-    filter(Species == i)
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'), family = binomial)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  # loop through all lakes where the species is present, lake list is based on the information from
-  # Projet Lac
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    # using minimum and maximum temp in the lake for derivatives
-     newdata <- tibble(mean_last_7days = seq(from = min(data_lake$temp, na.rm = TRUE),
-                                             to = max(data_lake$temp, na.rm = TRUE), length = 200),
-                       fProtocol = factor("VERT"))
-
-     # calculate derivatives
-     derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-       mutate(fLake = factor(j)) |>
-       mutate(species = factor(i)) |>
-       rename(temp = data)
-
-     saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-  }
-}
-
-# Model 2.1: Abundance data and no random intercept for lake -> but binomial
-# Coregonus_sp_benthic_profundal did not run with ZIP and was thus
-# run with binomial.
-
-species_list <- df_abundance_gam |> 
-  # binomial one
-  filter(Species == "Coregonus_sp_benthic_profundal") |>
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_gam |> 
-    filter(Species == i)
-
-  # binomial family instead of ZIP
-  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) +
-                           s(fProtocol, bs = 're'), family = binomial())
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-  }
-}
-
-# Model 2.2: Abundance data and no random intercept for lake -> ZIP
-# all other species in this group run with ZIP
-
-species_list <- df_abundance_gam |>
-  filter(!Species == "Coregonus_sp_benthic_profundal") |>
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_gam |> 
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'),
-                         family = ziP())
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-  }
-}
-
-# Model 3: Binomial data and random intercept for lake
-
-species_list <- df_binomial_re |> 
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_binomial_re |> 
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = "re")
-                         + s(fProtocol, bs = 're'), family = binomial)
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  unique_lakes <- distinct(data, Lake) |> 
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    # take all lakes where the species is present and take random one for 
-    # the data included in derivatives()
-    random_lake <- sample(unique_lakes, 1)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), lenqh = 200),
-      fProtocol = factor("VERT"), fLake = factor(random_lake))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-
-  }
-}
-
-# Model 4.1: Abundance data and random intercept for lake, simplified models
-# multiple species in this group were simplified and run with binomial
-
-species_list <- df_abundance_re |>
-  # simplified models
-  filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
-                                "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta",
-                        "Lepomis_gibbosus")) |>
-  distinct(Species) |>
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_re |>
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
-                         +  s(fProtocol, bs = 're'), family = binomial)
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  unique_lakes <- distinct(data, Lake) |>
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    random_lake <- sample(unique_lakes, 1)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fLake = factor(random_lake), fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-
-  }
-}
-
-# Model 4.1: Abundance data and random intercept for lake, complex models
-
-species_list <- df_abundance_re |> 
-  # exclusion of simplified models
-  filter(!Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
-                        "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta", 
-                        "Lepomis_gibbosus")) |>
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_re |> 
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
-                         +  s(fProtocol, bs = 're'), family = ziP())
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  unique_lakes <- distinct(data, Lake) |> 
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    random_lake <- sample(unique_lakes, 1)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fLake = factor(random_lake), fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = data)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-
-  }
-}
-
-# save all the derivatives in one file
-
-df_deriv <- list.files(path = "total_models/derivatives", pattern = ".rds", full.names = TRUE) |> 
-  map_dfr(readRDS)
-
-saveRDS(df_deriv, "total_models/df_deriv_all.rds")
-
-# save total derivatives as RDS (no exclusions)
-# saveRDS(df_deriv, "total_models/df_deriv_all_no_excl.rds")
-
-```
-
-###Fig. S2 
-Overview of derivatives in each lake community
-
-```{r}
-# Read in derivative data
-all_models_derivatives <- readRDS("total_models/df_deriv_all.rds")
-all_derivatives <- as_tibble(all_models_derivatives)
-
-str(all_derivatives)
-
-# Highlighting species with interesting derivatives: 
-species_group <- all_derivatives |> 
-  filter(species %in% c("Phoxinus_csikii", "Barbatula_sp_Lineage_II", "Cottus_sp_Po_profundal")) |> 
-  distinct(species) |> 
-  pull(species)
-
-# Plot S2: derivatives included in each lake community
-overview_derivatives <- all_derivatives |>
-  arrange(fLake) |> 
-  ggplot(aes(temp, derivative, color = species)) +
-  geom_line() +
-  facet_wrap(~fLake) +
-  theme_bw(base_size = 16) +
-  scale_color_viridis(discrete = TRUE, breaks = species_group, option = "G") +
-  xlab("Temperature") +
-  ylab("Derivatives") +
-  theme(panel.grid.major = element_blank(), 
-        panel.grid.minor = element_blank(), 
-        legend.position = "right",
-        legend.title = element_blank()) +
-  geom_hline(yintercept = 0, color = "#7A7A7A")
-
-overview_derivatives
-
-# save as TIFF
-# tiff(paste("total_models/plots/overview_derivatives_lake.tiff", sep = ""), units="in", width=11, height=7, res=300)
-# 
-# plot(overview_derivatives)
-# 
-# dev.off()
-```
-
-###Fig. S3
-
-```{r}
-# Data of the derivatives before simplifying multiple models
-# without changes the 5 species were not run with simplified models. 
-all_models_derivatives_no_excl <- readRDS("total_models/df_deriv_all_no_excl.rds")
-df_deriv_no_excl <- as_tibble(all_models_derivatives_no_excl)
-
-# Plot Figure S3a
-#calculating maximum and mean derivatives for each species per lake
-
-df_quantiles <- df_deriv_no_excl |> 
-  group_by(species, fLake) |> 
-  mutate(max_derivative = max(derivative)) |> 
-  mutate(mean_derivative = mean(derivative)) |> 
-  ungroup()
-
-# calculate quantiles of derivatives
-transform(quantile(df_quantiles$derivative,
-                   c(0,0.01,0.05,0.1,0.25,0.5,0.75,0.85,0.9,0.95,0.99,1)))
-
-# add column with the category of quantile each derivative is in
-df_quantiles$groups <- cut(df_quantiles$derivative,              
-                     breaks = c(-1.579943e+06, -2.342866e+00, -8.822087e-01,
-                                -4.543748e-01, 3.454218e-04, 2.485142e-01,
-                                5.527792e-01, 2.911987e+00, 2.018455e+01, 1.280898e+02,
-                                1.231212e+07, 1.406101e+07),
-                     labels = c("0-1%", "1-5%", "5-10%", "10-25%", "25-50%", "50-75%", "75-85%", "85-90%", "90-95%",
-                                "95-99%", "99-100%"))
-
-# Plot S3a: histogram of species' mean derivatives per lake
-plot_percentiles <- df_quantiles |> 
-  # Salmo trutta derivative is too large to be plotted
-  filter(species != "Salmo_trutta") |> 
-  ggplot(aes(x = mean_derivative, fill = factor(species))) + 
-  geom_histogram(aes(y = after_stat(count / sum(count))), binwidth = 1.5) +
-  scale_y_continuous(labels = scales::percent) +
-  scale_fill_manual(values = c("Cyprinus_carpio" = "#80CDC1",
-                                "Alburnus_arborella" = "#8C510A",
-                                "Barbatula_sp_Lineage_I"="#DFC27D",
-                               "Phoxinus_csikii" = "#01665E", "Salmo_trutta" = "#F46D43"))
-
-plot_percentiles
-
-# Plot S3a add lines for the percentiles
-plot_percentiles_1 <- plot_percentiles +
-  geom_vline(xintercept = 19.9, color = "red") +
-  annotate("text", x= 18, y = 0.12, label="90th percentile", angle=90) +
-  geom_vline(xintercept = 2.911987e+00, color = "red") +
-  annotate("text", x= 1.5, y = 0.12, label="85th percentile", angle=90) +
-  theme_classic(base_size = 13) +
-  xlab("Mean species derivative per lake") +
-  ylab("Percentage") +
- guides(fill = guide_legend(
-    title = "Species")) +
-  ylim(0, 0.2)
-
-plot_percentiles_1
-
-# save as TIFF
-# tiff(paste("total_models/plots/plot_deriv_percentiles_1.tiff", sep = ""), units="in", width=12, height=6, res=300)
-# 
-# plot(plot_percentiles_1)
-# 
-# dev.off()
-
-# above 90% = 2.018455e+01
-df_quantiles |> 
-  filter(derivative > 2.018455e+01) |> 
-  distinct(species)
-
-# 5 species above 90% -> those will be simplified to binomial 
-# 1 Alburnus_arborella    
-# 2 Barbatula_sp_Lineage_I
-# 3 Cyprinus_carpio       
-# 4 Phoxinus_csikii       
-# 5 Salmo_trutta
-
-# Plotting Figure S3b
-
-# just keeping mean and maximum derivative values for each lake and species
-data_quantiles <- df_quantiles |> 
-  distinct(mean_derivative, max_derivative, fLake, species)
-
-# Breaking mean derivatives in the same percentiles as before
-data_quantiles$percentiles <- cut(data_quantiles$mean_derivative,              
-                     breaks = c(-1.579943e+06, -2.342866e+00, -8.822087e-01,
-                                -4.543748e-01, 3.454218e-04, 2.485142e-01,
-                                5.527792e-01, 2.911987e+00, 2.018455e+01, 1.280898e+02,
-                                1.231212e+07, 1.406101e+07),
-                     labels = c("0-1%", "1-5%", "5-10%", "10-25%", "25-50%", "50-75%", "75-85%", "85-90%", "90-95%",
-                                "95-99%", "99-100%"))
-
-# Plot S3b tile plot of mean derivatives per species in each lake
-mean_deriv_plot <- data_quantiles |> 
-  ggplot(aes(fLake, y = fct_reorder(species, mean_derivative), fill= percentiles)) + 
-  geom_tile() +
-  scale_fill_manual(breaks = levels(data_quantiles$percentiles),
-                    values = rev(brewer.pal(11, "BrBG")))
-
-derivative_percentiles_2 <- mean_deriv_plot +
-  guides(fill = guide_legend(title = "percentiles (mean derivative)", reverse = TRUE)) +
-  xlab("") +
-  ylab("") +
-  theme_classic(base_size = 16)
-
-derivative_percentiles_2
-
-# save plot as TIFF
-tiff(paste("total_models/plots/plot_deriv_percentiles_2.tiff", sep = ""), units="in", width=19, height=12, res=300)
-
-plot(derivative_percentiles_2)
-
-dev.off()
-
-```
-
-##Calculating thermal response diversity
-
-```{r}
-# Function provided by Ross et al. "How to measure response diversity" used to calculate
-# response diversity
-
-# read in function
-source(here("functions.R"))
-
-# load derivatives data
-all_models_derivatives <- readRDS("total_models/df_deriv_all.rds")
-
-all_deriv <- as_tibble(all_models_derivatives)
-
-str(all_deriv)
-
-levels(all_deriv$species)
-
-#Calculating response diversity for each lake community
-
-# Lepomis is excluded from the analysis
-all_deriv <- all_deriv |>
-  filter(!species %in% c("Lepomis_gibbosus"))
-
-all_deriv$fLake <- as.character(all_deriv $fLake)
-
-lakes_list <- all_deriv |> 
-  distinct(fLake) |> 
-  pull(fLake)
-
-str(lakes_list)
-
-lakes_list <- sort(lakes_list)
-
-species_overview <- tibble()
-
-all_deriv$species <- as.character(all_deriv$species)
-all_deriv$species <- as.factor(all_deriv$species)
-
-str(all_deriv )
-
-# Loop to get response diversity measures for each lake 
-# loop through list of lakes and calculate response diversity in the lake communities
-for (i in lakes_list){
-
-  data <- all_deriv |>
-    select(temp, fLake, derivative, species) |> 
-    filter(fLake == i)
-
-  df_resp_div <- data |>
-    pivot_wider(
-      names_from = species,
-      values_from = derivative)
-
-  # apply the function from Ross et al. 
-  # Dissimilarity
-  df_resp_div$rdiv <- apply(df_resp_div[,-(1:2), drop = FALSE], 1, resp_div, sign_sens = F)
-  # Divergence
-  df_resp_div$sign <- apply(df_resp_div[,-(1:2), drop = FALSE], 1, resp_div, sign_sens = T)
-  # Median Dissimilarity
-  df_resp_div$Med <- median(df_resp_div$rdiv)
-
-  # save response diversity metrics per lake
-  saveRDS(df_resp_div, paste0("total_models/lakes_all_models/df_resp_div_", i, ".rds"))
-
-}
-
-# save the response diversity per lake into one file
-resp_div_no_excl <- list.files(path = "total_models/lakes_all_models", pattern = ".rds", full.names = TRUE) |>
-  map_dfr(readRDS) |>
-  relocate(rdiv, Med, sign, .after = temp)
-
-saveRDS(resp_div_no_excl,"total_models/resp_div_all.rds")
-```
-
-##Drivers of thermal response diversity
-
-Preparing a df with the information on past eutrophication, species richness, and species categorization in each lake.
-```{r}
-# Df with properties of lakes
-lake_info <- read_xlsx("lake_info.xlsx")
-
-# select the columns we use
-lake_selection <- lake_info |> 
-  select(Lake, Phos_max, Max_depth, Lake_area)
-
-# information about categorization in each lake for the subgroups
-species_endemism <- read_excel("species_category_per_lake.xlsx") |> 
-  rename(endemism = detail_category)
-
-# exclude lepomis gibbosus and re-calculate species richness per lake
-species_endemism_short <- species_endemism |> 
-  filter(species != "Lepomis_gibbosus") |> 
-  select(-sum_species) |> 
-  group_by(fLake) |> 
-  mutate(sum_species = sum(num_species))
-
-species_endemism_short$endemism <- as.factor(species_endemism_short$endemism)
-
-str(species_endemism_short)
-
-# adding column with number of species per category in each lake
-df_species_endemism <- species_endemism_short |> 
-  select(-species, - num_species) |> 
-  add_count(fLake, endemism) |> 
-  rename(count = n) |> 
-  distinct(fLake, count, sum_species, endemism)
-
-str(df_species_endemism)
-
-df_species_endemism$count <- as.numeric(df_species_endemism$count)
-
-# making it wide 
-df_species_endemism_long <- df_species_endemism |> 
-  pivot_wider(names_from = endemism, values_from = count) |> 
-  mutate_if(is.numeric, ~replace(., is.na(.), 0)) |> 
-  rename(Lake = fLake)
-
-str(df_species_endemism_long)
-
-# merge endemism info and lake info
-
-endemism_lake <- merge(df_species_endemism_long, lake_selection)
-
-# Read in data on response diversity metrics in each lake
-resp_div_all <- readRDS("total_models/resp_div_all.rds") |> 
-  select(temp:fLake)
-
-# select response diversity columns we need
-rdiv <- resp_div_all |>  
-  select(temp, fLake, rdiv, sign) |>
-  drop_na()
-
-# calculate mean and maximum values and rename Lake
-df_rdiv <- rdiv |> 
-  group_by(fLake) |> 
-  mutate(mean_rdiv = mean(rdiv)) |> 
-  mutate(max_rdiv = max(rdiv)) |> 
-  mutate(mean_sign = mean(sign)) |> 
-  mutate(max_sign = max(sign)) |> 
-  distinct(fLake, mean_rdiv, max_rdiv, mean_sign, max_sign) |> 
-  rename(Lake = fLake)
-
-# combine lake properties, species subgroups and response diversity
-
-df_lm <- merge(df_rdiv, endemism_lake)
-
-str(df_lm)
-
-# phosphorus values as numeric
-df_lm$Phos_max <- as.numeric(df_lm$Phos_max)
-
-```
-
-###Species richness and eutrophication
-Multiple linear regression to calculate the effect of species richness and
-eutrophication on response diversity.
-
-###Fig. 3
-
-```{r}
-# Multiple linear regression for mean dissimilarity explained by species richness and past eutrophication: 
-# lm(mean_rdiv  ~ species richness + eutrophication)
-# Predict manually with new data eutrophication value fixed at mean and species richness min() to max()
-# do the same the other way around: mean_rdiv  ~  eutrophication + species richness, with species richness fixed at mean and eutrophication min() to max()species richness +
-#   prepare these two plots
-
-# Phosphorus data on Poschiavo is not available
-df_lm_excl <- df_lm |> 
-  filter(Lake != "Poschiavo")
-
-# first lm for species richness ~ maximum phosphorus
-
-lm1 <- lm(mean_rdiv ~ sum_species + Phos_max, data = df_lm_excl)
-
-# checking model assumptions
-summary(lm1)
-shapiro.test(resid(lm1))
-lmtest::bptest(lm1)
-plot(lm1)
-
-# model assumptions validated
-
-# manually predict data
-new_data <- tibble(sum_species = seq(from = min(df_lm_excl$sum_species), to = max(df_lm_excl$sum_species), length = 50), Phos_max = mean(df_lm_excl$Phos_max))
-
-# predict lm
-prediction_lm1 <- predict.lm(lm1, newdata = new_data, se.fit = TRUE, type = "response")
-
-# combine new data and predictions
-bind_model1 <- cbind(new_data, prediction_lm1) |> 
-  mutate(category ="color")
-
-# second lm for maximum phosphorus ~ species richness
-lm2 <- lm(mean_rdiv ~ Phos_max + sum_species, data = df_lm_excl)
-
-# checking model assumptions
-summary(lm2)
-shapiro.test(resid(lm2))
-lmtest::bptest(lm2)
-plot(lm2)
-
-# new data
-new_data2 <- tibble(Phos_max = seq(from = min(df_lm_excl$Phos_max), to = max(df_lm_excl$Phos_max),length = 50), sum_species = mean(df_lm_excl$sum_species))
-
-# predict model
-prediction_lm2 <- predict.lm(lm2, newdata = new_data2, se.fit = TRUE, type = "response")
-
-# combine new data and predictions
-bind_model2 <- cbind(new_data2, prediction_lm2) |> 
-  mutate(category ="color")
-
-# Plot Fig. 3
-"#78BE63"
-mycolor <- c("color" = "#78BE63")
-# mycolor <- c("color" = "#A0E66A")
-"98D47C"
-
-lm1_plot <- bind_model1 |> 
-  ggplot(aes(sum_species, fit,  color = category, fill = category)) +
-  geom_line(linewidth = 1.2) +
-  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
-  theme_classic(base_size = 20) +
-  ylab("Mean dissimilarity") +
-  xlab("Species richness") +
-  ylim(1,3) +
-  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
-
-lm2_plot <- bind_model2 |> 
-  ggplot(aes(Phos_max, fit, color = category, fill = category)) +
-  geom_line(linewidth = 1.2) +
-  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
-  theme_classic(base_size = 20) +
-  ylab("Mean dissimilarity") +
-  xlab("Maximum total phosphorus (µg/L)") +
-  ylim(1,3) +
-  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
-
-# combine both plots
-figure_3 <- ggarrange(lm1_plot, lm2_plot)
-
-figure_3
-# save as TIFF
-# tiff(paste("total_models/plots/lm_species_richness_eutroph.tiff", sep = ""), units="in", width=12, height=5, res=300)
-# 
-# plot(figure_3)
-# 
-# # Closing the graphical device
-# dev.off()
-
-```
-
-###Divergence figure 3
-
-```{r}
-# first lm for species richness ~ maximum phosphorus
-
-lm3 <- lm(mean_sign ~ sum_species + Phos_max, data = df_lm_excl)
-
-# checking model assumptions
-summary(lm3)
-shapiro.test(resid(lm3))
-lmtest::bptest(lm3)
-plot(lm3)
-
-# model assumptions validated
-
-# manually predict data
-new_data3 <- tibble(sum_species = seq(from = min(df_lm_excl$sum_species), to = max(df_lm_excl$sum_species), length = 50), Phos_max = mean(df_lm_excl$Phos_max))
-
-# predict lm
-prediction_lm3 <- predict.lm(lm3, newdata = new_data3, se.fit = TRUE, type = "response")
-
-# combine new data and predictions
-bind_model3 <- cbind(new_data3, prediction_lm3) |> 
-  mutate(category ="color")
-
-# second lm for maximum phosphorus ~ species richness
-lm4 <- lm(mean_sign ~ Phos_max + sum_species, data = df_lm_excl)
-
-# checking model assumptions
-summary(lm4)
-shapiro.test(resid(lm4))
-lmtest::bptest(lm4)
-plot(lm4)
-
-# new data
-new_data4 <- tibble(Phos_max = seq(from = min(df_lm_excl$Phos_max), to = max(df_lm_excl$Phos_max),length = 50), sum_species = mean(df_lm_excl$sum_species))
-
-# predict model
-prediction_lm4 <- predict.lm(lm4, newdata = new_data4, se.fit = TRUE, type = "response")
-
-# combine new data and predictions
-bind_model4 <- cbind(new_data4, prediction_lm4) |> 
-  mutate(category ="color")
-
-# Plot Fig. 3
-mycolor <- c("color" = "#88D44F")
-# mycolor <- c("color" = "#A0E66A")
-"98D47C"
-
-lm3_plot <- bind_model3 |> 
-  ggplot(aes(sum_species, fit,  color = category, fill = category)) +
-  geom_line(linewidth = 1.2) +
-  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
-  theme_classic(base_size = 20) +
-  ylab("Mean divergence") +
-  xlab("Species richness") +
-  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
-
-lm4_plot <- bind_model4 |> 
-  ggplot(aes(Phos_max, fit, color = category, fill = category)) +
-  geom_line(linewidth = 1.2) +
-  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
-  theme_classic(base_size = 20) +
-  ylab("Mean divergence") +
-  xlab("Maximum total phosphorus (µg/L)") +
-  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
-
-# combine both plots
-figure_divergence <- ggarrange(lm3_plot, lm4_plot)
-
-figure_divergence
-# save as TIFF
-# tiff(paste("total_models/plots/lm_divergence_multiple_linear.tiff", sep = ""), units="in", width=12, height=5, res=300)
-# 
-# plot(figure_divergence)
-# 
-# # Closing the graphical device
-# dev.off()
-
-```
-###Community subgroups
-####Fig. 4a
-We fitted simple linear models for the number of species in each category and the mean dissimilarity in each lake. 
-```{r}
-# Lms for all categories: endemic, translcated, non_native and non_endemic_native
-# with data on all lakes
-# Endemic
-lm_endemic <- lm(mean_rdiv ~ endemic, data = df_lm)
-
-# checking model assumptions
-summary(lm_endemic)
-shapiro.test(resid(lm_endemic))
-lmtest::bptest(lm_endemic)
-plot(lm_endemic)
-
-new_data_a <- tibble(endemic = seq(from = min(df_lm$endemic), to = max(df_lm$endemic),
-                                     length = 15))
-
-prediction_endemic <- predict.lm(lm_endemic, newdata = new_data_a, se.fit = TRUE, type = "response")
-
-df_lm_endemic <- cbind(new_data_a, prediction_endemic) |> 
-  mutate(category = factor("endemic")) |> 
-  rename(species_richness = endemic)
-
-# Non-native
-
-lm_nn <- lm(mean_rdiv ~ non_native, data = df_lm)
-
-# checking model assumptions
-summary(lm_nn)
-shapiro.test(resid(lm_nn))
-lmtest::bptest(lm_nn)
-plot(lm_nn)
-
-new_data_b <- tibble(non_native = seq(from = min(df_lm$non_native), to = max(df_lm$non_native),
-                                   length = 15))
-
-prediction_nn <- predict.lm(lm_nn, newdata = new_data_b, se.fit = TRUE, type = "response")
-
-df_lm_nn <- cbind(new_data_b, prediction_nn)  |> 
-  mutate(category = factor("non_native")) |> 
-  rename(species_richness = non_native)
-
-# Non-endemic native
-
-lm_nne <- lm(mean_rdiv ~ native, data = df_lm)
-
-# checking model assumptions
-summary(lm_nne)
-shapiro.test(resid(lm_nne))
-lmtest::bptest(lm_nne)
-plot(lm_nne)
-
-new_data_c <- tibble(native = seq(from = min(df_lm$native), to = max(df_lm$native),
-                                      length = 15))
-
-prediction_nne <- predict.lm(lm_nne, newdata = new_data_c, se.fit = TRUE, type = "response")
-
-df_lm_nne <- cbind(new_data_c, prediction_nne)  |> 
-  mutate(category = factor("non_endemic_native")) |> 
-  rename(species_richness = native)
-
-# Translocated (non-native to region) species
-
-lm_trans <- lm(mean_rdiv ~ non_native_region, data = df_lm)
-
-# checking model assumptions
-summary(lm_trans)
-shapiro.test(resid(lm_trans))
-lmtest::bptest(lm_trans)
-plot(lm_trans)
-
-new_data_d <- tibble(non_native_region = seq(from = min(df_lm$non_native_region), to = max(df_lm$non_native_region),
-                                              length = 15))
-
-prediction_trans <- predict.lm(lm_trans, newdata = new_data_d, se.fit = TRUE, type = "response")
-
-df_lm_trans <- cbind(new_data_d, prediction_trans)  |> 
-  mutate(category = factor("translocated")) |> 
-  rename(species_richness = non_native_region)
-
-# Combine all four lm predictions
-df_all_lms <- rbind(df_lm_endemic, df_lm_nn, df_lm_nne, df_lm_trans)
-
-mycolors <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
-               "translocated" = "#F2AD63")
-
-plot_all_lms <- df_all_lms|> 
-  ggplot(aes(species_richness, fit, color = category, fill = category, linetype = category)) +  
-  geom_line(linewidth = 1) +
-  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), color = NA, alpha = 0.3) +
-  theme_classic(base_size = 16) +
-  ylab("Mean dissimilarity") +
-  xlab("Species richness") +
-  scale_fill_manual(values = mycolors,
-                    aesthetics = c("color", "fill"),
-                    labels = c("endemic", "non-native (ns)", "non-endemic native", "translocated (ns)")) +
-   scale_linetype_manual(values = c("endemic" = "solid",
-                                   "non_native" = "dashed",
-                                   "non_endemic_native" = "solid",
-                                   "translocated" = "dashed"), labels = c("endemic", "non-native (ns)", "non-endemic native", "translocated (ns)"))
-
-plot_all_lms
-# save as TIFF
-# tiff(paste("total_models/plots/lm_all.tiff", sep = ""), units="in", width=11, height=7, res=300)
-# # plot(ggarrange(depth1, depth2, ncol = 2))
-# # plot science discussion
-# plot(plot_all_lms)
-# 
-# dev.off()
-```
-
-####Divergence Fig. 4a
-
-```{r}
-# Lms for all categories: endemic, translcated, non_native and non_endemic_native
-# with data on all lakes
-# Endemic
-deriv_lm1 <- lm(mean_sign ~ endemic, data = df_lm)
-
-# checking model assumptions
-summary(deriv_lm1)
-shapiro.test(resid(deriv_lm1))
-lmtest::bptest(deriv_lm1)
-plot(deriv_lm1)
-
-new_data_1 <- tibble(endemic = seq(from = min(df_lm$endemic), to = max(df_lm$endemic),
-                                     length = 15))
-
-prediction_endemic_1 <- predict.lm(deriv_lm1, newdata = new_data_1, se.fit = TRUE, type = "response")
-
-df_deriv_endemic <- cbind(new_data_1, prediction_endemic_1) |> 
-  mutate(category = factor("endemic")) |> 
-  rename(species_richness = endemic)
-
-# Non-native
-
-deriv_lm2 <- lm(mean_sign ~ non_native, data = df_lm)
-
-# checking model assumptions
-summary(deriv_lm2)
-shapiro.test(resid(deriv_lm2))
-lmtest::bptest(deriv_lm2)
-plot(deriv_lm2)
-
-new_data_2 <- tibble(non_native = seq(from = min(df_lm$non_native), to = max(df_lm$non_native),
-                                   length = 15))
-
-prediction_2 <- predict.lm(deriv_lm2, newdata = new_data_2, se.fit = TRUE, type = "response")
-
-df_deriv_nn <- cbind(new_data_2, prediction_2)  |> 
-  mutate(category = factor("non_native")) |> 
-  rename(species_richness = non_native)
-
-# Non-endemic native
-
-deriv_lm3 <- lm(mean_sign ~ native, data = df_lm)
-
-# checking model assumptions
-summary(deriv_lm3)
-shapiro.test(resid(deriv_lm3))
-lmtest::bptest(deriv_lm3)
-plot(deriv_lm3)
-
-new_data_3 <- tibble(native = seq(from = min(df_lm$native), to = max(df_lm$native),
-                                      length = 15))
-
-prediction_3 <- predict.lm(deriv_lm3, newdata = new_data_3, se.fit = TRUE, type = "response")
-
-df_deriv_nne <- cbind(new_data_3, prediction_3)  |> 
-  mutate(category = factor("non_endemic_native")) |> 
-  rename(species_richness = native)
-
-# Translocated (non-native to region) species
-
-deriv_lm4 <- lm(mean_sign ~ non_native_region, data = df_lm)
-
-# checking model assumptions
-summary(deriv_lm4)
-shapiro.test(resid(deriv_lm4))
-lmtest::bptest(deriv_lm4)
-plot(deriv_lm4)
-
-new_data_4 <- tibble(non_native_region = seq(from = min(df_lm$non_native_region), to = max(df_lm$non_native_region),
-                                              length = 15))
-
-prediction_4 <- predict.lm(deriv_lm4, newdata = new_data_4, se.fit = TRUE, type = "response")
-
-df_deriv_trans <- cbind(new_data_4, prediction_4)  |> 
-  mutate(category = factor("translocated")) |> 
-  rename(species_richness = non_native_region)
-
-# Combine all four lm predictions
-df_deriv_lms <- rbind(df_deriv_endemic, df_deriv_nn, df_deriv_nne, df_deriv_trans)
-
-mycolors <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
-               "translocated" = "#F2AD63")
-
-lm_divergence_all <- df_deriv_lms|> 
-  ggplot(aes(species_richness, fit, color = category, fill = category)) +
-  geom_line(linewidth = 1, linetype = "dashed") +
-  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), color = NA, alpha = 0.3) +
-  theme_classic(base_size = 16) +
-  ylab("Mean divergence") +
-  xlab("Species richness") +
-  scale_fill_manual(values = mycolors,
-                    aesthetics = c("color", "fill"), 
-                    labels = c("endemic", "non-native", "non-endemic native", "translocated"),
-                    name = NULL)
-
-lm_divergence_all
-
-# save as TIFF
-# tiff(paste("total_models/plots/lm_divergence_all.tiff", sep = ""), units="in", width=11, height=7, res=300)
-# # plot(ggarrange(depth1, depth2, ncol = 2))
-# # plot science discussion
-# plot(lm_divergence_all)
-# 
-# dev.off()
-```
-#### Fig. 4b 
-Predictions of all species shown inside the categories. 
-```{r, fig.show='hide'}
-
-# Plot species predictions inside categories across all lakes
-# 5 categories;
-# endemic: geographically constrained range
-# non-native: from NA or Asia 
-# non-endemic native: native to Switzerland but not endemic
-# non-native region: native to Switzerland and surroundings but has been translocated
-# to other Swiss lakes where the species was not native
-# endemic translocated: endemic species that were translocated to other lakes
-
-# read in table with the categories and rename the categories
-species_category <- read_excel("species_category_across_lakes.xlsx") |> 
-  select(-notes) |> 
-  mutate(category = ifelse(category == "native", "non_endemic_native", category)) |> 
-  mutate(category = ifelse(category == "non_native_region", "non_endemic_native_and_translocated", category)) |> 
-  mutate(category2 = factor("all"))
-
-species_category$category <- as.factor(species_category$category)
-levels(species_category$category)
-
-# merge predictions with categories
-model_pred_categories <- merge(model_predictions, species_category)
-
-# change category names for facet_wrap
-category_names <- c(
-  `endemic` = "endemic",
-  `endemic_and_translocated` = "translocated endemic",
-  `non_endemic_native` = "non-endemic native",
-  `non_endemic_native_and_translocated` = "translocated non-endemic native",
-  `non_native` = "non-native"
-)
-
-# colors for the plot
-mycolors1 <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
-                "endemic_and_translocated" = "#F2AD63", "non_endemic_native_and_translocated" = "#F2AD63")
-
-# Rescale the predictions to 0-1 
-model_pred_categories_rescaled <- model_pred_categories |> 
-  group_by(species) |> 
-  mutate(fit_rescaled = (fit - min(fit)) / (max(fit) - min(fit)))
-
-# Plot the rescaled predictions
-plot_pred <- model_pred_categories_rescaled |>
-  # exclude Lepomis gibbosus
-  filter(species != "Lepomis_gibbosus") |> 
-  ggplot(aes(temp, fit_rescaled, group = species, color = category)) +
-  geom_line(linewidth = 0.8) +
-  theme_bw(base_size = 16) +
-  ylab("Abundance") +
-  xlab("Temperature") +
-  scale_color_manual(values = mycolors1, guide = NULL) +
-  theme(panel.grid.major = element_blank(), 
-        panel.grid.minor = element_blank(), 
-        legend.position = "bottom",
-        legend.title = element_blank())
-
-plot_category_predictions <- plot_pred + facet_grid(~category, labeller = as_labeller(category_names))
-
-plot_category_predictions
-
-# save plot as TIFF
-# tiff(paste("total_models/plots/plot_category_predictions.tiff", sep = ""), units="in", width=15, height=4, res=300)
-# 
-# plot(plot_category_predictions)
-# 
-# dev.off()
-
-```

### 2024-07-05 models_derivatives_response_diversity.Rmd
lepomis run with zip

@@
 
-no_exclusion_predictions_all <- list.files(path = "total_models/predictions/temp", pattern = ".rds", full.names = TRUE) |>
+simplified_predictions_all <- list.files(path = "total_models/predictions/temp", pattern = ".rds", full.names = TRUE) |>
   map_dfr(readRDS)
@@
 # save total predictions as RDS
-saveRDS(no_exclusion_predictions_all, "total_models/df_pred_all.rds")
+saveRDS(simplified_predictions_all, "total_models/df_pred_all.rds")
 
@@
 
+```
+
+```{r}
+# check out lepomis gibbosus
+excl_model_predictions <- readRDS("total_models/df_pred_no_exclusion.rds")
+excl_model_predictions |> 
+  filter(species == "Lepomis_gibbosus") |> 
+    ggplot(aes(temp, fit)) +
+  # geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit), fill = factor(species)), alpha = 0.3) +
+  geom_line(aes(color = factor(species))) +
+  theme_bw() +
+  facet_wrap(~species, scale = "free_y", labeller = labeller(species = label_wrap_gen(width = 10))) +
+  theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
+  scale_color_viridis(discrete = TRUE, guide = NULL, option = "G", aesthetics = c("color", "fill")) +
+  xlab("Temperature") +
+  ylab("Abundance")
 ```

### 2024-07-05 models_derivatives_response_diversity.Rmd
tidier, added new fig 2b

--- 2024-07-05 models_derivatives_response_diversity.Rmd
@@
 ---
-#2. Data Analysis
-##General Additive Models
+#General Additive Models
 Running the predictions for all GAMs with the smaller data frames to reduce running 
@@
 
-# use own function predictions() from function_GAM.R to predict species' models
-# across all lakes
-
+# we have three different functions. One predictions() does not simply any models
+# predictions_temp() has multiple models that are simplified (see supplementary 
+# figures for the disucssion), predictions_depth_temp() models all species with 
+# abundance > 20 with both coviarates temperature and depth.
+
+# For downstream analysis for thermal response diversity models from predictions_temp()
+# are used.
+
+# temperature models with simplified models 
 # predictions take VERY LONG! and two folders are needed to store the 
@@
 # total_models/gam_check
-# total_models/predictions
-# with three subfolders: temp, temp_test, depth_temp (total_models/predictions/depth_temp)
+# needs a folder called total_models/predictions/temp
+
+predictions_temp(df_1)
+predictions_temp(df_2)
+predictions_temp(df_3)
+predictions_temp(df_4)
+
+simplified_predictions_all <- list.files(path = "total_models/predictions/temp", pattern = ".rds", full.names = TRUE) |>
+  map_dfr(readRDS)
+
+# save total predictions as RDS
+saveRDS(simplified_predictions_all, "total_models/df_pred_all.rds")
+
+# you can directly use the dataframe df_pred_all for downstream analysis
+
+```
+###Simplification and temp depth models
+```{r}
+# both other functions:
+
+# you can also directly use the dataframe df_pred_no_simplification for decisions about
+# which models should be simplified 
+# and
+# depth_temp_pred_all for the sensitivity testing of the temp + depth models
+
+# testing for which models need to be simplified
+# needs a folder: total_models/predictions/temp_test
 
@@
 # # save total predictions as RDS
-# saveRDS(no_exclusion_predictions_all, "total_models/df_pred_no_exclusion.rds")
-
-# temperature models with simplified models
-
-predictions_temp(df_1)
-predictions_temp(df_2)
-predictions_temp(df_3)
-predictions_temp(df_4)
-
-no_exclusion_predictions_all <- list.files(path = "total_models/predictions/temp", pattern = ".rds", full.names = TRUE) |>
-  map_dfr(readRDS)
-
-# save total predictions as RDS
-saveRDS(no_exclusion_predictions_all, "total_models/df_pred_all.rds")
+# saveRDS(no_exclusion_predictions_all, "total_models/df_pred_no_simplification.rds")
+
+# testing if modelling temp and depth improves the models
+# needs a folder: total_models/predictions/depth_temp
 
@@
 # # issue with cottus sp po
-# # statistics of the gamms
+
+# # statistics of the gamms with dept and temp
 # total_models <- bind_rows(df_1, df_2, df_3, df_4)
@@
 
-####Fig. 2
-Overview of all model predictions
+####Fig. 2b
+New figure 2b with categories across all lakes
 
@@
     mutate(category = ifelse(category == "native", "non_endemic_native", category)) |>
-  mutate(category = ifelse(category == "non_native_region", "non_endemic_native_and_translocated", category)) |> 
   mutate(category2 = factor("all"))
@@
 
-# EVTL NEW PLOT 2
 # change category names for facet_wrap
@@
   `endemic` = "endemic",
-  `endemic_and_translocated` = "translocated endemic",
   `non_endemic_native` = "non-endemic native",
-  `non_endemic_native_and_translocated` = "translocated non-endemic native",
   `non_native` = "non-native"
@@
 # colors for the plot
-mycolors1 <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
-                "endemic_and_translocated" = "#F2AD63", "non_endemic_native_and_translocated" = "#F2AD63")
+mycolors1 <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8")
 
@@
 
-plot_pred <- predictions_categories_rescaled |>
+plot_2b <- predictions_categories_rescaled |>
   ggplot(aes(temp, fit_rescaled, group = species, color = category)) +
-  geom_line() +
-  # how can I get the right color?
-   # geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), alpha = 0.3) +
+   geom_line(linewidth = 0.5) +
   theme_bw(base_size = 16) +
@@
 
-plot_category_predictions <- plot_pred + facet_grid(~category, labeller = as_labeller(category_names))
-
-plot_category_predictions
-
+plot_2b_final <- plot_2b + facet_grid(~category, labeller = as_labeller(category_names))
+
+plot_2b_final
+
+# save plot as TIFF
+# tiff(paste("total_models/plots/plot_2b.tiff", sep = ""), units="in", width=10, height=4, res=300)
+# 
+# plot(plot_2b_final)
+# 
+# dev.off()
+
+```
+
+####Fig. Supplementary predictions plot
+```{r}
+# also using predictions_categories from chunk above
 # trying to group by category and color
-figure_test_2 <- predictions_categories |>
+
+str(categories_supp)
+
+categories_supp <- predictions_categories |>
+  arrange(category) |> 
+  mutate(species = fct_reorder(species, as.numeric(category))) |> 
   ggplot(aes(temp, fit)) +
@@
   theme_bw() +
-  facet_wrap(~species + category, scale = "free_y", labeller = labeller(species = label_wrap_gen(width = 10))) +
+  facet_wrap(~species, scale = "free_y", labeller = labeller(species = label_wrap_gen(width = 10))) +
   theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
@@
 
-# tiff(paste("total_models/plots/draft_all_predictions.tiff", sep = ""), units="in", width = 12, height=8, res=300)
-# plot(figure_test_2)
+categories_supp
+
+# tiff(paste("total_models/plots/supplementary_predictions.tiff", sep = ""), units="in", width = 12, height=8, res=300)
+# plot(categories_supp)
 # 
@@
 # dev.off()
+
+```
+##Sensitivity testing
+Testing if depth-driven temperature models can adequately fit abundance data compared to a temperature and depth model.
+
+```{r}
+temp_only <- readRDS("total_models/df_pred_all.rds") |> 
+  mutate(category = "temp")
+temp_depth <- readRDS("total_models/depth_temp_pred_all.rds") |> 
+  mutate(category = "temp_depth")
+
+sensitivity
+
+```
+
+```{r}
 
@@
 # dev.off()
-
-```
-
-##Sensitivity testing
-NOT CHANGED 
-Testing if depth-driven temperature models can adequately fit abundance data compared to
-a temperature and depth model. 
-Models were compared in the eleven most abundant species.
+```
 
@@
 ```
-##Derivatives calculation
-
+#Derivatives calculation
+
+##Data preparation
 We calculated derivatives for each species per lake.
@@
 
-###Temp models
+##Derivatives
 ```{r}
@@
 
-# Model 4: before simplification
-
-species_list <- df_abundance_re |> 
-  # exclusion of simplified models
-  filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
-                        "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta",
-                        "Lepomis_gibbosus")) |>
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_re |> 
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
-                         +  s(fProtocol, bs = 're'), family = ziP())
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  unique_lakes <- distinct(data, Lake) |> 
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    random_lake <- sample(unique_lakes, 1)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fLake = factor(random_lake), fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = mean_last_7days)
-
-    saveRDS(derivatives, paste0("total_models/derivatives_no_simplification/derivatives_", i, "_",  j, ".rds"))
-
-  }
-}
-
 # Model 4.1: Abundance data and random intercept for lake, simplified models
@@
   filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
-                                "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta",
-                        "Lepomis_gibbosus")) |>
+                        "Cyprinus_carpio","Phoxinus_csikii", "Salmo_trutta")) |>
   distinct(Species) |>
@@
   filter(!Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
-                        "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta",
-                        "Lepomis_gibbosus")) |>
+                        "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta")) |>
   distinct(Species) |> 
@@
 
+```
+
+#Testing derivatives
+##Simplification of models
+```{r}
+# to get the derivatives before simplifying any model: 
+# copy all other derivatives to a new folder called total_models/derivatives_no_simplification and then run this code.
+
+# # Model 4: before simplification
+# species_list <- df_abundance_re |>
+#   # no simplification
+#   filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
+#                         "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta")) |>
+#   distinct(Species) |> 
+#   pull(Species)
+# 
+# species_list <- sort(species_list)
+# 
+# derivatives <- list()
+# gam_output <- list()
+# model_prediction <- list()
+# 
+# 
+# for (i in species_list) {
+#   
+#   data <- df_abundance_re |> 
+#     filter(Species == i)
+#   
+#   gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
+#                          +  s(fProtocol, bs = 're'), family = ziP())
+#   
+#   lake_data <- species_lake |>
+#     filter(Species == i)
+#   
+#   lake_list <- distinct(lake_data, Lake) |>
+#     pull()
+#   
+#   unique_lakes <- distinct(data, Lake) |> 
+#     pull()
+#   
+#   for (j in lake_list){
+#     
+#     data_lake <- species_lake |>
+#       filter(Species == i) |>
+#       filter(Lake == j)
+#    
+#     random_lake <- sample(unique_lakes, 1)
+# 
+#     newdata <- tibble(mean_last_7days = seq(
+#       from = min(data_lake$temp, na.rm = TRUE),
+#       to = max(data_lake$temp, na.rm = TRUE), length = 200),
+#       fLake = factor(random_lake), fProtocol = factor("VERT"))
+#     
+#     derivatives <- derivatives(gam_output[[i]], data = newdata) |>
+#       mutate(fLake = factor(j)) |>
+#       mutate(species = factor(i)) |>
+#       rename(temp = mean_last_7days)
+#     
+#     saveRDS(derivatives, paste0("total_models/derivatives_no_simplification/derivatives_", i, "_",  j, ".rds"))
+# 
+#   }
+# }
+
 # save total derivatives as RDS (no exclusions)
@@
 
-###Fig. S2 
+##Fig. S3ab
+
+```{r}
+# cyprinus carpio and lepomis should not be simplified
+
+# Data of the derivatives before simplifying multiple models
+# without changes the 5 species were not run with simplified models. 
+all_models_derivatives_no_excl <- readRDS("total_models/df_deriv_all_no_excl.rds")
+df_deriv_no_excl <- as_tibble(all_models_derivatives_no_excl)
+
+# Plot Figure S3a
+#calculating maximum and mean derivatives for each species per lake
+
+df_quantiles <- df_deriv_no_excl |> 
+  group_by(species, fLake) |> 
+  mutate(max_derivative = max(.derivative)) |> 
+  mutate(mean_derivative = mean(.derivative)) |> 
+  ungroup()
+
+# calculate quantiles of derivatives
+transform(quantile(df_quantiles$.derivative,
+                   c(0,0.01,0.05,0.1,0.25,0.5,0.75,0.85,0.9,0.95,0.99,1)))
+
+# add column with the category of quantile each derivative is in
+df_quantiles$groups <- cut(df_quantiles$.derivative,              
+                     breaks = c(-1.579943e+06, -2.706252e+00, -1.269492e+00,
+                                -3.348207e-01, 1.844665e-03, 2.262231e-01,
+                                9.864616e-01, 2.018455e+01, 3.013806e+01, 1.648215e+06,
+                                1.303406e+07, 1.406101e+07),
+                     labels = c("0-1%", "1-5%", "5-10%", "10-25%", "25-50%", "50-75%", "75-85%", "85-90%", "90-95%",
+                                "95-99%", "99-100%"))
+
+# Plot S3a: histogram of species' mean derivatives per lake
+plot_percentiles <- df_quantiles |> 
+  # Salmo trutta derivative is too large to be plotted
+  filter(species != "Salmo_trutta") |> 
+  ggplot(aes(x = mean_derivative, fill = factor(species))) + 
+  geom_histogram(aes(y = after_stat(count / sum(count))), binwidth = 1.5) +
+  scale_y_continuous(labels = scales::percent) +
+  scale_fill_manual(values = c("Cyprinus_carpio" = "#80CDC1",
+                                "Alburnus_arborella" = "#8C510A",
+                                "Barbatula_sp_Lineage_I"="#DFC27D",
+                               "Phoxinus_csikii" = "#01665E",  "Salmo_trutta" = "#F46D43"))
+
+plot_percentiles
+
+# Plot S3a add lines for the percentiles
+plot_percentiles_1 <- plot_percentiles +
+  geom_vline(xintercept = 3.013806e+01, color = "red") +
+  annotate("text", x= 26, y = 0.12, label="90th percentile", angle=90) +
+  geom_vline(xintercept = 2.018455e+01, color = "red") +
+  annotate("text", x= 16, y = 0.12, label="85th percentile", angle=90) +
+  theme_classic(base_size = 13) +
+  xlab("Mean species derivative per lake") +
+  ylab("Percentage") +
+ guides(fill = guide_legend(
+    title = "Species")) +
+  ylim(0, 0.2)
+
+plot_percentiles_1
+
+# save as TIFF
+# tiff(paste("total_models/plots/plot_deriv_percentiles_1.tiff", sep = ""), units="in", width=12, height=6, res=300)
+# 
+# plot(plot_percentiles_1)
+# 
+# dev.off()
+
+# above the 85th percentile
+
+df_quantiles |> 
+  filter(.derivative > 2.018455e+01) |> 
+  distinct(species)
+
+# 1 Alburnus_arborella    
+# 2 Barbatula_sp_Lineage_I
+# 3 Cyprinus_carpio       
+# 4 Phoxinus_csikii       
+# 5 Salmo_trutta
+
+# NEW: upon inclusion of Lepomis gibbosus
+# above 90% = 3.013806e+01
+df_quantiles |> 
+  filter(.derivative > 3.013806e+01) |> 
+  distinct(species)
+
+# 1 Alburnus_arborella    
+# 2 Barbatula_sp_Lineage_I
+# 3 Phoxinus_csikii       
+# 4 Salmo_trutta
+
+# 5 species above 90% -> those will be simplified to binomial 
+# 1 Alburnus_arborella    
+# 2 Barbatula_sp_Lineage_I
+# 3 Cyprinus_carpio       
+# 4 Phoxinus_csikii       
+# 5 Salmo_trutta
+
+# Plotting Figure S3b
+
+# just keeping mean and maximum derivative values for each lake and species
+data_quantiles <- df_quantiles |> 
+  distinct(mean_derivative, max_derivative, fLake, species)
+
+# Breaking mean derivatives in the same percentiles as before
+
+data_quantiles$percentiles <- cut(data_quantiles$mean_derivative,              
+                     breaks = c(-1.579943e+06, -2.706252e+00, -1.269492e+00,
+                                -3.348207e-01, 1.844665e-03, 2.262231e-01,
+                                9.864616e-01, 2.018455e+01, 3.013806e+01, 1.648215e+06,
+                                1.303406e+07, 1.406101e+07),
+                     labels = c("0-1%", "1-5%", "5-10%", "10-25%", "25-50%", "50-75%", "75-85%", "85-90%", "90-95%",
+                                "95-99%", "99-100%"))
+
+# Plot S3b tile plot of mean derivatives per species in each lake
+mean_deriv_plot <- data_quantiles |> 
+  ggplot(aes(fLake, y = fct_reorder(species, mean_derivative), fill= percentiles)) + 
+  geom_tile() +
+  scale_fill_manual(breaks = levels(data_quantiles$percentiles),
+                    values = rev(brewer.pal(11, "BrBG")))
+
+derivative_percentiles_2 <- mean_deriv_plot +
+  guides(fill = guide_legend(title = "percentiles (mean derivative)", reverse = TRUE)) +
+  xlab("") +
+  ylab("") +
+  theme_classic(base_size = 16)
+
+derivative_percentiles_2
+
+# save plot as TIFF
+# tiff(paste("total_models/plots/plot_deriv_percentiles_2.tiff", sep = ""), units="in", width=19, height=12, res=300)
+# 
+# plot(derivative_percentiles_2)
+# 
+# dev.off()
+
+```
+
+##Fig. S2: Overview derivatives
 Overview of derivatives in each lake community
@@
 
-###Fig. S3
-
-```{r}
-# cyprinus carpio and lepomis should not be simplified
-
-# Data of the derivatives before simplifying multiple models
-# without changes the 5 species were not run with simplified models. 
-all_models_derivatives_no_excl <- readRDS("total_models/df_deriv_all_no_excl.rds")
-df_deriv_no_excl <- as_tibble(all_models_derivatives_no_excl)
-
-# Plot Figure S3a
-#calculating maximum and mean derivatives for each species per lake
-
-df_quantiles <- df_deriv_no_excl |> 
-  group_by(species, fLake) |> 
-  mutate(max_derivative = max(.derivative)) |> 
-  mutate(mean_derivative = mean(.derivative)) |> 
-  ungroup()
-
-# calculate quantiles of derivatives
-transform(quantile(df_quantiles$.derivative,
-                   c(0,0.01,0.05,0.1,0.25,0.5,0.75,0.85,0.9,0.95,0.99,1)))
-
-# add column with the category of quantile each derivative is in
-df_quantiles$groups <- cut(df_quantiles$.derivative,              
-                     breaks = c(-1.579943e+06, -2.706252e+00, -1.269492e+00,
-                                -3.348207e-01, 1.844665e-03, 2.262231e-01,
-                                9.864616e-01, 2.018455e+01, 3.013806e+01, 1.648215e+06,
-                                1.303406e+07, 1.406101e+07),
-                     labels = c("0-1%", "1-5%", "5-10%", "10-25%", "25-50%", "50-75%", "75-85%", "85-90%", "90-95%",
-                                "95-99%", "99-100%"))
-
-# Plot S3a: histogram of species' mean derivatives per lake
-plot_percentiles <- df_quantiles |> 
-  # Salmo trutta derivative is too large to be plotted
-  filter(species != "Salmo_trutta") |> 
-  ggplot(aes(x = mean_derivative, fill = factor(species))) + 
-  geom_histogram(aes(y = after_stat(count / sum(count))), binwidth = 1.5) +
-  scale_y_continuous(labels = scales::percent) +
-  scale_fill_manual(values = c("Cyprinus_carpio" = "#80CDC1",
-                                "Alburnus_arborella" = "#8C510A",
-                                "Barbatula_sp_Lineage_I"="#DFC27D",
-                               "Phoxinus_csikii" = "#01665E",  "Salmo_trutta" = "#F46D43"))
-
-plot_percentiles
-
-# Plot S3a add lines for the percentiles
-plot_percentiles_1 <- plot_percentiles +
-  geom_vline(xintercept = 3.013806e+01, color = "red") +
-  annotate("text", x= 26, y = 0.12, label="90th percentile", angle=90) +
-  geom_vline(xintercept = 2.018455e+01, color = "red") +
-  annotate("text", x= 16, y = 0.12, label="85th percentile", angle=90) +
-  theme_classic(base_size = 13) +
-  xlab("Mean species derivative per lake") +
-  ylab("Percentage") +
- guides(fill = guide_legend(
-    title = "Species")) +
-  ylim(0, 0.2)
-
-plot_percentiles_1
-
-# save as TIFF
-# tiff(paste("total_models/plots/plot_deriv_percentiles_1.tiff", sep = ""), units="in", width=12, height=6, res=300)
-# 
-# plot(plot_percentiles_1)
-# 
-# dev.off()
-
-# above 90% = 2.018455e+01
-df_quantiles |> 
-  filter(.derivative > 2.018455e+01) |> 
-  distinct(species)
-
-# 5 species above 90% -> those will be simplified to binomial 
-# 1 Alburnus_arborella    
-# 2 Barbatula_sp_Lineage_I
-# 3 Cyprinus_carpio       
-# 4 Phoxinus_csikii       
-# 5 Salmo_trutta
-
-# Plotting Figure S3b
-
-# just keeping mean and maximum derivative values for each lake and species
-data_quantiles <- df_quantiles |> 
-  distinct(mean_derivative, max_derivative, fLake, species)
-
-# Breaking mean derivatives in the same percentiles as before
-
-data_quantiles$percentiles <- cut(data_quantiles$mean_derivative,              
-                     breaks = c(-1.579943e+06, -2.706252e+00, -1.269492e+00,
-                                -3.348207e-01, 1.844665e-03, 2.262231e-01,
-                                9.864616e-01, 2.018455e+01, 3.013806e+01, 1.648215e+06,
-                                1.303406e+07, 1.406101e+07),
-                     labels = c("0-1%", "1-5%", "5-10%", "10-25%", "25-50%", "50-75%", "75-85%", "85-90%", "90-95%",
-                                "95-99%", "99-100%"))
-
-# Plot S3b tile plot of mean derivatives per species in each lake
-mean_deriv_plot <- data_quantiles |> 
-  ggplot(aes(fLake, y = fct_reorder(species, mean_derivative), fill= percentiles)) + 
-  geom_tile() +
-  scale_fill_manual(breaks = levels(data_quantiles$percentiles),
-                    values = rev(brewer.pal(11, "BrBG")))
-
-derivative_percentiles_2 <- mean_deriv_plot +
-  guides(fill = guide_legend(title = "percentiles (mean derivative)", reverse = TRUE)) +
-  xlab("") +
-  ylab("") +
-  theme_classic(base_size = 16)
-
-derivative_percentiles_2
-
-# save plot as TIFF
-tiff(paste("total_models/plots/plot_deriv_percentiles_2.tiff", sep = ""), units="in", width=19, height=12, res=300)
-
-plot(derivative_percentiles_2)
-
-dev.off()
-
-```
-
-##Calculating thermal response diversity
+#Calculating thermal response diversity
 

### 2024-07-05 drivers_response_diversity.Rmd
tidied

@@
 library(grid)
-
-# function for the GAMs (old)
-# source(here("function_GAM.R"))
-
-# new function for temp and depth models 
-source(here("function_temp_depth.R"))
+library(viridis)
 ```
@@
   xlab("Species richness") +
-  ylim(1,3) +
+  # ylim(1,3) +
   scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
@@
   xlab("Maximum total phosphorus (µg/L)") +
-  ylim(1,3) +
+  # ylim(1,3) +
   scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)

### 2024-07-05 models_derivatives_response_diversity.Rmd
new sensitivity figure

@@
 ##Sensitivity testing
+###NEW Fig. S1
 Testing if depth-driven temperature models can adequately fit abundance data compared to a temperature and depth model.
@@
 temp_only <- readRDS("total_models/df_pred_all.rds") |> 
-  mutate(category = "temp")
+  mutate(model = "temp")
+
 temp_depth <- readRDS("total_models/depth_temp_pred_all.rds") |> 
-  mutate(category = "temp_depth")
-
-sensitivity
-
-```
-
-```{r}
-
-# OLD FIGURE 2
-
-all_predictions <- predictions_categories |>
-  ggplot(aes(temp, fit)) +
-  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit), fill = factor(species)), alpha = 0.3) +
-  geom_line(aes(color = factor(species))) +
-  theme_bw() +
-  facet_wrap(~species, scale = "free_y", labeller = labeller(species = label_wrap_gen(width = 10))) +
-  theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
-  scale_color_viridis(discrete = TRUE, guide = NULL, option = "G", aesthetics = c("color", "fill")) +
-  xlab("Temperature") +
-  ylab("Abundance")
-
-all_predictions
-# save the plot as TIFF
-# tiff(paste("total_models/plots/all_predictions_models.tiff", sep = ""), units="in", width = 12, height=8, res=300)
-# plot(all_predictions)
-# 
-# # Closing the graphical device
-# dev.off()
-```
-
-```{r}
-# Read in abundance and depth data of all species included for modelling
-df_models <- readRDS("df_models.rds")
-
-# find 11 most abundant  species
-species_list <- df_models |> 
-  filter(tot_abu > 300) |> 
-  distinct(Species) |> 
-  pull(Species)
-
-# most abundant species
- # "Alburnus_alburnus", "Coregonus_sp_albeli", "Coregonus_sp","Gasterosteus_aculeatus",
- # "Gobio_gobio", "Leuciscus_leuciscus", "Perca_fluviatilis", "Rutilus_rutilus", 
- # "Sander_lucioperca", "Scardinius_erythrophthalmus","Gymnocephalus_cernua"
-
-# filter df into the most abundant species and select the columns of interest
-
-df_sensitivity <- df_models |> 
-  filter(tot_abu > 300) |> 
-  select(Lake, Species, Protocol, Depth_sample, Abundance, Presence, mean_last_7days) |> 
-  #there are some NAs in depth -> drop them
-  drop_na(Depth_sample, mean_last_7days)
-
-df_sensitivity$fLake <- as.factor(df_sensitivity$Lake)
-df_sensitivity$fProtocol <- as.factor(df_sensitivity$Protocol)
-
-# Loop through the eleven species and fit temp and temp-depth model
-
-species_list <- df_sensitivity |> 
-  distinct(Species) |> 
-  pull(Species)
-
-# loop through the species, model both GAMs, check DHARMa and predictions
-for (i in species_list){
-
-  data <- df_sensitivity |> 
-    filter(Species == i)
-
-  #Model fit for temp and depth
-  model1 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(Depth_sample, k = 3)
-                + s(fLake, bs = 're')
-                +  s(fProtocol, bs = 're'), family = ziP())
-  # Model fit for temp
-  model2 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3)
-                + s(fLake, bs = 're')
-                +  s(fProtocol, bs = 're'), family = ziP())
-  # validate model predictions for temp and depth model with dharma:
-  # prepare residuals
-  simulationOutput <- simulateResiduals(fittedModel = model1, plot = F)
-  # tiff_filename <- paste("total_models/gam_check/temp_depth_check_", i, ".tiff", sep = "")
-  # tiff(tiff_filename, width = 800, height = 600)
-  print(plot(simulationOutput))
-  # dev.off()
-  # Plotting standardized residuals against predictors
-  # tiff_file_2 <- paste("total_models/gam_check/temp_depth_predictor_", i, ".tiff", sep = "")
-  # tiff(tiff_file_2, width = 800, height = 600)
-  print(plotResiduals(model1, data$mean_last_7days, xlab = "temp", main=NULL))
-  # dev.off()
-
-  # for predictions with random effects in the GAMs, we select a random value
-  # of the level of random effects and exclude the random effects in the prediction
-
-  # lakes where the species is present
-  unique_lakes <- distinct(data, Lake) |>
-    pull()
-
-  # select random lake for prediction
-  random_lake <- sample(unique_lakes, 1)
-
-  # for temp-depth model only predict for depth where the species is present
-  presence_depth <- data |> 
-    filter(Presence == 1) |>
-    distinct(Depth_sample) |> 
-    pull(Depth_sample)
-
-  # take randomly 5 depth values for prediction
-  random_depth <- sort(sample(presence_depth, 5))
-
-  # Convert sorted values to factors with original order
-  factor_random_depth <- factor(random_depth, levels = random_depth, ordered = TRUE)
-
-  # grid for each of the models
-
-  # temp-depth model predicted with random depths where the species is present
- grid1 <- expand.grid(mean_last_7days = seq(
-    from = min(data$mean_last_7days, na.rm = TRUE),
-    to = max(data$mean_last_7days, na.rm = TRUE), length = 200),
-    fProtocol = factor("VERT"), fLake = factor(random_lake), 
-    Depth_sample = factor_random_depth)
-
-  grid2 <- expand.grid(mean_last_7days = seq(
-    from = min(data$mean_last_7days, na.rm = TRUE),
-    to = max(data$mean_last_7days, na.rm = TRUE), length = 1000),
-    fProtocol = factor("VERT"), fLake = factor(random_lake))
-
-  # print relevant stats
-  print(paste(i))
-  print(glance(model1))
-  print(glance(model2))
-
-  # manually predict both models
-  model_prediction1 <- predict.gam(model1, newdata = grid1,
-                                  exclude = c("s(fProtocol)", "s(fLake)"), #exclude the re
-                                  term = "s(mean_last_7days)", type = "response",             
-                                  se.fit = TRUE)
-
-  model_bind1 <- cbind(grid1, as.data.frame(model_prediction1)) |>
-    mutate(model = factor("temp_and_depth"))
-
-# predict model2
-  model_prediction2 <- predict.gam(model2, newdata = grid2,
-                                   exclude = c("s(fProtocol)", "s(fLake)"),
-                                   type = "response", se.fit = TRUE)
-
-  model_bind2 <- cbind(grid2, as.data.frame(model_prediction2)) |>
-    mutate(model = factor("temp"))
-
-  df_pred <- bind_rows(model_bind1, model_bind2) |>
-    rename(temp = mean_last_7days) |>
-    select(-fProtocol, -fLake) |>
-    mutate(Depth_sample = factor(ifelse(is.na(Depth_sample), "temp", as.character(Depth_sample))))
-
-  # Extract numeric depth levels
-  numeric_levels <- levels(df_pred$Depth_sample)[!levels(df_pred$Depth_sample) %in% "temp"]
-
-  # Sort numeric depth levels
-  sorted_numeric_levels <- sort(as.numeric(numeric_levels))
-
-  # Combine sorted numeric levels with "temp" level
-  sorted_levels <- c(as.character(sorted_numeric_levels), "temp")
-
-  # Reorder levels
-  df_pred$Depth_sample <- factor(df_pred$Depth_sample, levels = sorted_levels)
-
-  # Check the levels to confirm they are ordered from lowest to highest
-  levels(df_pred$Depth_sample)
-
-  # rescale predictions of both models
-  df_pred_rescaled <- df_pred |>
-    group_by(model) |>
-    mutate(fit_rescaled = (fit - min(fit)) / (max(fit) - min(fit))) |>
-    mutate(se_fit_rescaled = (se.fit - min(se.fit)) / (max(se.fit) - min(se.fit))) |> 
-    mutate(species = factor(i))
-
-  # save predictions of both models in a rds
-   saveRDS(df_pred_rescaled, paste0("total_models/sensitivity/rds/sensitivity_predictions_", i, ".rds"))
-
-}
-
-# prepare total predictions of all models
-sensitivity_predictions <- list.files(path = "total_models/sensitivity/rds", pattern = ".rds", full.names = TRUE) |>
-  map_dfr(readRDS)
-
-# save total predictions as RDS
-saveRDS(sensitivity_predictions, "total_models/sensitivity_predictions.rds")
-
-```
-
-###Fig. S1
-
-```{r}
-# read in the sensitivity testing data
-sensitivity_predictions <- readRDS("total_models/sensitivity_predictions.rds")
-
-levels(sensitivity_predictions$species)
-
-# Plot both model predictions together per species
- figure_s1 <- sensitivity_predictions |>
-    ggplot(aes(temp, fit_rescaled, color = model, fill = model)) +
+  mutate(model = "temp_depth")
+
+sensitivity_predictions <- bind_rows(temp_only, temp_depth)
+
+supplementary_sensitivity <- sensitivity_predictions |>
+    ggplot(aes(temp, fit, color = model, fill = model)) +
     geom_line() +
-    geom_ribbon(aes(ymin = (fit_rescaled - se.fit), ymax = (fit_rescaled + se.fit)), color = NA, alpha = 0.1) +
+    # geom_ribbon(aes(ymin = (fit_rescaled - se.fit), ymax = (fit_rescaled + se.fit)), color = NA, alpha = 0.1) +
     theme_bw() +
     # labs(title = paste("Sensitivity Test", i)) +
-    scale_color_manual(values = c("#08306B", "#D73027"), aesthetics = c("color", "fill")) +
-    facet_wrap(~species, scale = "free") +
+    scale_color_manual(values = c("#D73027", "#08306B"), aesthetics = c("color", "fill")) +
+     facet_wrap(~species, scale = "free", labeller = labeller(species = label_wrap_gen(width = 10))) +
+    theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
     ylab("Abundance") +
@@
 
-  figure_s1
-  # save as TIFF
-# tiff(paste("total_models/plots/sensitivity_overview.tiff", sep = ""), units="in", width=11, height=7, res=300)
-# 
-# plot(figure_s1)
-# 
+# # save the plot as TIFF
+# tiff(paste("total_models/plots/supplementary_sensitivity_predictions.tiff", sep = ""), units="in", width = 12, height=8, res=300)
+# plot(supplementary_sensitivity)
+# 
+# # Closing the graphical device
 # dev.off()
@@
 ```
+####AIC comparison 
+```{r}
+# to be added
+```
+
 #Derivatives calculation

### 2024-07-05 function_GAM.R
lepomis is not simplified

@@ -259,7 +259,7 @@ predictions_temp <- function(df){
       
       # simplified models
       if (i %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
-                   "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta", "Lepomis_gibbosus")){
+                   "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta")){
         
         
         gam_output<- gam(data = data, Presence ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')

### 2024-07-05 drivers_response_diversity.Rmd
(ns) deleted from figure

@@
                                    "non_endemic_native" = "solid",
-                                   "translocated" = "dashed"), labels = c("endemic", "non-native (ns)", "non-endemic native", "translocated (ns)"))
+                                   "translocated" = "dashed"), labels = c("endemic", "non-native", "non-endemic native", "translocated"))
 

### 2024-07-08 drivers_response_diversity.Rmd
new plots 3 and 4 for both dissimilarity and divergence

@@
 
+# pltting the first linear regression
+lm1_plot <- bind_model1 |> 
+  ggplot(aes(sum_species, fit,  color = category, fill = category)) +
+  geom_line(linewidth = 1.2) +
+  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
+  theme_classic(base_size = 20) +
+  ylab("Mean dissimilarity") +
+  xlab("Species richness") +
+  # ylim(1,3) +
+  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
+
+# Lms for all categories: endemic, translcated, non_native and non_endemic_native
+# with data on all lakes
+# Endemic
+lm_endemic <- lm(mean_rdiv ~ endemic, data = df_lm)
+
+# checking model assumptions
+summary(lm_endemic)
+shapiro.test(resid(lm_endemic))
+lmtest::bptest(lm_endemic)
+plot(lm_endemic)
+
+new_data_a <- tibble(endemic = seq(from = min(df_lm$endemic), to = max(df_lm$endemic),
+                                     length = 15))
+
+prediction_endemic <- predict.lm(lm_endemic, newdata = new_data_a, se.fit = TRUE, type = "response")
+
+df_lm_endemic <- cbind(new_data_a, prediction_endemic) |> 
+  mutate(category = factor("endemic")) |> 
+  rename(species_richness = endemic)
+
+# Non-native
+
+lm_nn <- lm(mean_rdiv ~ non_native, data = df_lm)
+
+# checking model assumptions
+summary(lm_nn)
+shapiro.test(resid(lm_nn))
+lmtest::bptest(lm_nn)
+plot(lm_nn)
+
+new_data_b <- tibble(non_native = seq(from = min(df_lm$non_native), to = max(df_lm$non_native),
+                                   length = 15))
+
+prediction_nn <- predict.lm(lm_nn, newdata = new_data_b, se.fit = TRUE, type = "response")
+
+df_lm_nn <- cbind(new_data_b, prediction_nn)  |> 
+  mutate(category = factor("non_native")) |> 
+  rename(species_richness = non_native)
+
+# Non-endemic native
+
+lm_nne <- lm(mean_rdiv ~ native, data = df_lm)
+
+# checking model assumptions
+summary(lm_nne)
+shapiro.test(resid(lm_nne))
+lmtest::bptest(lm_nne)
+plot(lm_nne)
+
+new_data_c <- tibble(native = seq(from = min(df_lm$native), to = max(df_lm$native),
+                                      length = 15))
+
+prediction_nne <- predict.lm(lm_nne, newdata = new_data_c, se.fit = TRUE, type = "response")
+
+df_lm_nne <- cbind(new_data_c, prediction_nne)  |> 
+  mutate(category = factor("non_endemic_native")) |> 
+  rename(species_richness = native)
+
+# Translocated (non-native to region) species
+
+lm_trans <- lm(mean_rdiv ~ non_native_region, data = df_lm)
+
+# checking model assumptions
+summary(lm_trans)
+shapiro.test(resid(lm_trans))
+lmtest::bptest(lm_trans)
+plot(lm_trans)
+
+new_data_d <- tibble(non_native_region = seq(from = min(df_lm$non_native_region), to = max(df_lm$non_native_region),
+                                              length = 15))
+
+prediction_trans <- predict.lm(lm_trans, newdata = new_data_d, se.fit = TRUE, type = "response")
+
+df_lm_trans <- cbind(new_data_d, prediction_trans)  |> 
+  mutate(category = factor("translocated")) |> 
+  rename(species_richness = non_native_region)
+
+# Combine all four lm predictions
+df_all_lms <- rbind(df_lm_endemic, df_lm_nn, df_lm_nne, df_lm_trans)
+
+mycolors <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
+               "translocated" = "#F2AD63")
+
+plot_all_lms <- df_all_lms|> 
+  ggplot(aes(species_richness, fit, color = category, fill = category, linetype = category)) +  
+  geom_line(linewidth = 1) +
+  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), color = NA, alpha = 0.3) +
+  theme_classic(base_size = 16) +
+  ylab("Mean dissimilarity") +
+  xlab("Species richness") +
+  scale_fill_manual(values = mycolors,
+                    aesthetics = c("color", "fill"),
+                    labels = c("endemic", "non-native (ns)", "non-endemic native", "translocated (ns)")) +
+   scale_linetype_manual(values = c("endemic" = "solid",
+                                   "non_native" = "dashed",
+                                   "non_endemic_native" = "solid",
+                                   "translocated" = "dashed"), labels = c("endemic", "non-native", "non-endemic native", "translocated"))
+
+plot_all_lms
+# save as TIFF
+# tiff(paste("total_models/plots/lm_all.tiff", sep = ""), units="in", width=11, height=7, res=300)
+# # plot(ggarrange(depth1, depth2, ncol = 2))
+# # plot science discussion
+# plot(plot_all_lms)
+# 
+# dev.off()
+
+```
+
+###Figure 4
+```{r}
+```
+
+###Figure 3 Divergence
+
+```{r}
+
+# first lm for species richness ~ maximum phosphorus
+
+lm3 <- lm(mean_sign ~ sum_species + Phos_max, data = df_lm_excl)
+
+# checking model assumptions
+summary(lm3)
+shapiro.test(resid(lm3))
+lmtest::bptest(lm3)
+plot(lm3)
+
+# model assumptions validated
+
+# manually predict data
+new_data3 <- tibble(sum_species = seq(from = min(df_lm_excl$sum_species), to = max(df_lm_excl$sum_species), length = 50), Phos_max = mean(df_lm_excl$Phos_max))
+
+# predict lm
+prediction_lm3 <- predict.lm(lm3, newdata = new_data3, se.fit = TRUE, type = "response")
+
+# combine new data and predictions
+bind_model3 <- cbind(new_data3, prediction_lm3) |> 
+  mutate(category ="color")
+
+lm3_plot <- bind_model3 |> 
+  ggplot(aes(sum_species, fit,  color = category, fill = category)) +
+  geom_line(linewidth = 1.2) +
+  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
+  theme_classic(base_size = 20) +
+  ylab("Mean divergence") +
+  xlab("Species richness") +
+  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
+
+# Lms for all categories: endemic, translcated, non_native and non_endemic_native
+# with data on all lakes
+# Endemic
+deriv_lm1 <- lm(mean_sign ~ endemic, data = df_lm)
+
+# checking model assumptions
+summary(deriv_lm1)
+shapiro.test(resid(deriv_lm1))
+lmtest::bptest(deriv_lm1)
+plot(deriv_lm1)
+
+new_data_1 <- tibble(endemic = seq(from = min(df_lm$endemic), to = max(df_lm$endemic),
+                                     length = 15))
+
+prediction_endemic_1 <- predict.lm(deriv_lm1, newdata = new_data_1, se.fit = TRUE, type = "response")
+
+df_deriv_endemic <- cbind(new_data_1, prediction_endemic_1) |> 
+  mutate(category = factor("endemic")) |> 
+  rename(species_richness = endemic)
+
+# Non-native
+
+deriv_lm2 <- lm(mean_sign ~ non_native, data = df_lm)
+
+# checking model assumptions
+summary(deriv_lm2)
+shapiro.test(resid(deriv_lm2))
+lmtest::bptest(deriv_lm2)
+plot(deriv_lm2)
+
+new_data_2 <- tibble(non_native = seq(from = min(df_lm$non_native), to = max(df_lm$non_native),
+                                   length = 15))
+
+prediction_2 <- predict.lm(deriv_lm2, newdata = new_data_2, se.fit = TRUE, type = "response")
+
+df_deriv_nn <- cbind(new_data_2, prediction_2)  |> 
+  mutate(category = factor("non_native")) |> 
+  rename(species_richness = non_native)
+
+# Non-endemic native
+
+deriv_lm3 <- lm(mean_sign ~ native, data = df_lm)
+
+# checking model assumptions
+summary(deriv_lm3)
+shapiro.test(resid(deriv_lm3))
+lmtest::bptest(deriv_lm3)
+plot(deriv_lm3)
+
+new_data_3 <- tibble(native = seq(from = min(df_lm$native), to = max(df_lm$native),
+                                      length = 15))
+
+prediction_3 <- predict.lm(deriv_lm3, newdata = new_data_3, se.fit = TRUE, type = "response")
+
+df_deriv_nne <- cbind(new_data_3, prediction_3)  |> 
+  mutate(category = factor("non_endemic_native")) |> 
+  rename(species_richness = native)
+
+# Translocated (non-native to region) species
+
+deriv_lm4 <- lm(mean_sign ~ non_native_region, data = df_lm)
+
+# checking model assumptions
+summary(deriv_lm4)
+shapiro.test(resid(deriv_lm4))
+lmtest::bptest(deriv_lm4)
+plot(deriv_lm4)
+
+new_data_4 <- tibble(non_native_region = seq(from = min(df_lm$non_native_region), to = max(df_lm$non_native_region),
+                                              length = 15))
+
+prediction_4 <- predict.lm(deriv_lm4, newdata = new_data_4, se.fit = TRUE, type = "response")
+
+df_deriv_trans <- cbind(new_data_4, prediction_4)  |> 
+  mutate(category = factor("translocated")) |> 
+  rename(species_richness = non_native_region)
+
+# Combine all four lm predictions
+df_deriv_lms <- rbind(df_deriv_endemic, df_deriv_nn, df_deriv_nne, df_deriv_trans)
+
+mycolors <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
+               "translocated" = "#F2AD63")
+
+lm_divergence_all <- df_deriv_lms|> 
+  ggplot(aes(species_richness, fit, color = category, fill = category)) +
+  geom_line(linewidth = 1, linetype = "dashed") +
+  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), color = NA, alpha = 0.3) +
+  theme_classic(base_size = 16) +
+  ylab("Mean divergence") +
+  xlab("Species richness") +
+  scale_fill_manual(values = mycolors,
+                    aesthetics = c("color", "fill"), 
+                    labels = c("endemic", "non-native", "non-endemic native", "translocated"),
+                    name = NULL)
+
+lm_divergence_all
+
+# save as TIFF
+# tiff(paste("total_models/plots/lm_divergence_all.tiff", sep = ""), units="in", width=11, height=7, res=300)
+# # plot(ggarrange(depth1, depth2, ncol = 2))
+# # plot science discussion
+# plot(lm_divergence_all)
+# 
+# dev.off()
+```
+
+```{r}
 # second lm for maximum phosphorus ~ species richness
@@
 
-###Divergence figure 3
+###Divergence figure 4
 
 ```{r}
-# first lm for species richness ~ maximum phosphorus
-
-lm3 <- lm(mean_sign ~ sum_species + Phos_max, data = df_lm_excl)
-
-# checking model assumptions
-summary(lm3)
-shapiro.test(resid(lm3))
-lmtest::bptest(lm3)
-plot(lm3)
-
-# model assumptions validated
-
-# manually predict data
-new_data3 <- tibble(sum_species = seq(from = min(df_lm_excl$sum_species), to = max(df_lm_excl$sum_species), length = 50), Phos_max = mean(df_lm_excl$Phos_max))
-
-# predict lm
-prediction_lm3 <- predict.lm(lm3, newdata = new_data3, se.fit = TRUE, type = "response")
-
-# combine new data and predictions
-bind_model3 <- cbind(new_data3, prediction_lm3) |> 
-  mutate(category ="color")
 

### 2024-07-08 drivers_response_diversity.Rmd
new plots 3 and 4

@@
 
-##Drivers of thermal response diversity
+#Drivers of thermal response diversity
 
@@
 
-###Species richness and eutrophication
-Multiple linear regression to calculate the effect of species richness and
-eutrophication on response diversity.
-
-###Fig. 3
+##Fig. 3
+Dissimilarity, species richness and categories
 
@@
 
+mycolor = "black"
+
 # pltting the first linear regression
@@
 
+lm1_plot
 # Lms for all categories: endemic, translcated, non_native and non_endemic_native
@@
                     aesthetics = c("color", "fill"),
-                    labels = c("endemic", "non-native (ns)", "non-endemic native", "translocated (ns)")) +
+                    labels = c("endemic", "non-native", "non-endemic native", "translocated")) +
    scale_linetype_manual(values = c("endemic" = "solid",
@@
 plot_all_lms
+
 # save as TIFF
@@
 
-###Figure 4
-```{r}
-```
-
-###Figure 3 Divergence
+## Figure 3 Divergence
 
@@
 
+## Figure 4
 ```{r}
@@
 "#78BE63"
-mycolor <- c("color" = "#78BE63")
+mycolor <- c("color" = "black")
 # mycolor <- c("color" = "#A0E66A")
 "98D47C"
-
-lm1_plot <- bind_model1 |> 
-  ggplot(aes(sum_species, fit,  color = category, fill = category)) +
-  geom_line(linewidth = 1.2) +
-  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
-  theme_classic(base_size = 20) +
-  ylab("Mean dissimilarity") +
-  xlab("Species richness") +
-  # ylim(1,3) +
-  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
 
@@
 
-# combine both plots
-figure_3 <- ggarrange(lm1_plot, lm2_plot)
-
-figure_3
+lm2_plot
 # save as TIFF
@@
 
-###Divergence figure 4
+## Figure 4 Divergence
 
@@
 ```
-###Community subgroups
+### OLD CODE Community subgroups
 ####Fig. 4a

### 2024-07-08 models_derivatives_response_diversity.Rmd
re-testing the addition of lepomis gibbosus

@@
 
-# # Model 4: before simplification
-# species_list <- df_abundance_re |>
-#   # no simplification
-#   filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
-#                         "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta")) |>
-#   distinct(Species) |> 
-#   pull(Species)
-# 
-# species_list <- sort(species_list)
-# 
-# derivatives <- list()
-# gam_output <- list()
-# model_prediction <- list()
-# 
-# 
-# for (i in species_list) {
-#   
-#   data <- df_abundance_re |> 
-#     filter(Species == i)
-#   
-#   gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
-#                          +  s(fProtocol, bs = 're'), family = ziP())
-#   
-#   lake_data <- species_lake |>
-#     filter(Species == i)
-#   
-#   lake_list <- distinct(lake_data, Lake) |>
-#     pull()
-#   
-#   unique_lakes <- distinct(data, Lake) |> 
-#     pull()
-#   
-#   for (j in lake_list){
-#     
-#     data_lake <- species_lake |>
-#       filter(Species == i) |>
-#       filter(Lake == j)
-#    
-#     random_lake <- sample(unique_lakes, 1)
-# 
-#     newdata <- tibble(mean_last_7days = seq(
-#       from = min(data_lake$temp, na.rm = TRUE),
-#       to = max(data_lake$temp, na.rm = TRUE), length = 200),
-#       fLake = factor(random_lake), fProtocol = factor("VERT"))
-#     
-#     derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-#       mutate(fLake = factor(j)) |>
-#       mutate(species = factor(i)) |>
-#       rename(temp = mean_last_7days)
-#     
-#     saveRDS(derivatives, paste0("total_models/derivatives_no_simplification/derivatives_", i, "_",  j, ".rds"))
-# 
-#   }
-# }
+# Model 4: before simplification
+species_list <- df_abundance_re |>
+# #   # no simplification
+filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
+                     "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta")) |>
+distinct(Species) |> 
+pull(Species)
+
+species_list <- sort(species_list)
+
+derivatives <- list()
+gam_output <- list()
+model_prediction <- list()
+
+for (i in species_list) {
+ data <- df_abundance_re |> 
+     filter(Species == i)
+
+ gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're') +  s(fProtocol, bs = 're'), family = ziP())
+
+ lake_data <- species_lake |>
+   filter(Species == i)
+
+ lake_list <- distinct(lake_data, Lake) |>
+   pull()
+
+ unique_lakes <- distinct(data, Lake) |> 
+     pull()
+
+  for (j in lake_list){
+    data_lake <- species_lake |>
+      filter(Species == i) |>
+     filter(Lake == j)
+
+     random_lake <- sample(unique_lakes, 1)
+
+     newdata <- tibble(mean_last_7days = seq(
+       from = min(data_lake$temp, na.rm = TRUE),
+       to = max(data_lake$temp, na.rm = TRUE), length = 200),
+       fLake = factor(random_lake), fProtocol = factor("VERT"))
+
+     derivatives <- derivatives(gam_output[[i]], data = newdata) |>
+       mutate(fLake = factor(j)) |>
+       mutate(species = factor(i)) |>
+       rename(temp = mean_last_7days)
+
+     saveRDS(derivatives, paste0("total_models/derivatives_no_simplification/derivatives_", i, "_",  j, ".rds"))
+
+   }
+ }
 
@@
 #
-# df_deriv_no_excl <- list.files(path = "total_models/derivatives_no_simplification", pattern = ".rds", full.names = TRUE) |>
-#   map_dfr(readRDS)
+df_deriv_no_excl <- list.files(path = "total_models/derivatives_no_simplification", pattern = ".rds", full.names = TRUE) |>
+  map_dfr(readRDS)
 #
-# saveRDS(df_deriv_no_excl, "total_models/df_deriv_all_no_excl.rds")
+saveRDS(df_deriv_no_excl, "total_models/df_deriv_all_no_excl.rds")
 

### 2024-07-08 models_derivatives_response_diversity.Rmd
re-examination of max derivatives

@@
 
-# Plot S3a: histogram of species' mean derivatives per lake
+#  derivatives per lake
 plot_percentiles <- df_quantiles |> 
@@
                                 "Barbatula_sp_Lineage_I"="#DFC27D",
-                               "Phoxinus_csikii" = "#01665E",  "Salmo_trutta" = "#F46D43"))
+                               "Phoxinus_csikii" = "#01665E",  "Salmo_trutta" = "#F46D43", "Cottus_sp_Po_profundal" = "red"))
 
@@
 plot_percentiles_1
+
+# LOOKING AT MAX DERIVATIVE
+# Plot S3a: histogram of species' mean derivatives per lake
+df_quantiles |> 
+  # Salmo trutta derivative is too large to be plotted
+  filter(species != "Salmo_trutta") |> 
+  ggplot(aes(x = max_derivative, fill = factor(species))) + 
+  geom_histogram(aes(y = after_stat(count / sum(count))), binwidth = 1.5) +
+  scale_y_continuous(labels = scales::percent) +
+  scale_fill_manual(values = c("Cyprinus_carpio" = "#80CDC1",
+                                "Alburnus_arborella" = "#8C510A",
+                                "Barbatula_sp_Lineage_I"="#DFC27D",
+                               "Phoxinus_csikii" = "#01665E",  "Salmo_trutta" = "#F46D43", "Cottus_sp_Po_profundal" = "red"))
 
@@
 derivative_percentiles_2
+
+# max derivatives per species in each lake
+max_deriv_plot <- data_quantiles |> 
+  ggplot(aes(fLake, y = fct_reorder(species, max_derivative), fill= percentiles)) + 
+  geom_tile() +
+  scale_fill_manual(breaks = levels(data_quantiles$percentiles),
+                    values = rev(brewer.pal(11, "BrBG")))
+
+max_deriv_plot +
+  guides(fill = guide_legend(title = "percentiles (max derivative)", reverse = TRUE)) +
+  xlab("") +
+  ylab("") +
+  theme_classic(base_size = 16)
 

### 2024-07-08 function_GAM.R
new function for stats overview

@@ -415,6 +415,385 @@ predictions_temp <- function(df){
   }
 }
 
+predictions_temp <- function(df){
+  require(broom)
+  require(tidyverse)
+  require(DHARMa)
+  require(mgcv)
+  require(gratia)
+  require(mgcViz)
+
+  
+  species_list <- df |> 
+    distinct(Species) |> 
+    pull(Species)
+  
+  species_list <- sort(species_list)
+  
+  gam_output <- list()
+  model_prediction <- list()
+  derivatives <- list()
+  grid <- list()
+  pred_df <- list()
+  tiff_filename <- list()
+  tiff_file_2 <- list()
+  temp_data <- list()
+  summary <- list()
+  
+  df$fLake <- as.factor(df$Lake)
+  
+  df$fProtocol <- as.factor(df$Protocol)
+  
+  for (i in species_list) {
+    data <- df |> 
+      filter(Species == i) |> 
+      mutate(n_lake = n_distinct(Lake))
+    
+    unique_lakes <- unique(data$fLake)
+    
+    random_lake <- sample(levels(unique_lakes), 1)
+    
+    # species occurring in one lake -> net-type as random effect
+    if(max(data$n_lake) == 1) {
+      
+      # newdata for all species that fall into this category -> occuring
+      # in one lake only
+      grid <- expand.grid(mean_last_7days = seq(
+        from = min(data$mean_last_7days, na.rm = TRUE),
+        to = max(data$mean_last_7days, na.rm = TRUE), by = 0.02),
+        fProtocol = factor("VERT"))
+      
+      # special case, because model not running with ZIP
+      if (i == "Coregonus_sp_benthic_profundal")  { 
+        
+        gam_output <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) +
+                            s(fProtocol, bs = 're'), family = binomial)
+        # prepare residuals
+        simulationOutput <- simulateResiduals(fittedModel = gam_output, plot = F)
+        tiff_filename <- paste("total_models/gam_check/gam_check_", i, ".tiff", sep = "")
+        tiff(tiff_filename, width = 800, height = 600)
+        print(plot(simulationOutput))
+        dev.off()
+        
+        # Plotting standardized residuals against predictors
+        tiff_file_2 <- paste("total_models/gam_check/predictor_", i, ".tiff", sep = "")
+        tiff(tiff_file_2, width = 800, height = 600)
[…]

### 2024-07-08 function_temp_depth.R
deleted code needed to test the function

@@ -953,11 +953,6 @@ predictions_depth_temp <- function(df){
 }
 
 
-subs <- total_models |> 
-  filter(Species == "Cottus_sp_Po")
-df <- subs
-
-i <- "Cottus_sp_Po"
 stats_gamms <- function(df){
   require(broom)
   require(tidyverse)
@@ -984,6 +979,7 @@ stats_gamms <- function(df){
   temp_data <- list()
   summary <- list()
   model_summary <- list()
+  total_summary <- list()
   
   df$fLake <- as.factor(df$Lake)
   

### 2024-09-26 03_drivers_response_diversity_Fig3_Fig4.Rmd
renamed script
preparing the code for figure 4 a and c

--- 2024-07-08 drivers_response_diversity.Rmd
@@
 lake_selection <- lake_info |> 
-  select(Lake, Phos_max, Max_depth, Lake_area)
+  select(Lake, Phos_max, Max_depth, Lake_area, Species_loss, Phos_max_Vonlanthen)
 
@@
 
-# exclude lepomis gibbosus and re-calculate species richness per lake
+# re-calculate species richness per lake
 species_endemism_short <- species_endemism |> 
-  filter(species != "Lepomis_gibbosus") |> 
+  # filter(species != "Lepomis_gibbosus") |> 
   select(-sum_species) |> 
@@
 df_lm$Phos_max <- as.numeric(df_lm$Phos_max)
+df_lm$Phos_max_Vonlanthen <- as.numeric(df_lm$Phos_max_Vonlanthen)
 
@@
 ```
+## Figure 4a
+```{r}
+df_lm_4 <- df_lm |> 
+  filter(!Lake %in% c("Poschiavo", "Joux", "Maggiore", "Lugano"))
 
-## Figure 4
+lm_species_loss <- lm(Species_loss ~ Phos_max_Vonlanthen, data = df_lm_4)
+
+# checking model assumptions
+summary(lm_species_loss)
+shapiro.test(resid(lm_species_loss))
+lmtest::bptest(lm_species_loss)
+plot(lm_species_loss)
+
+# new data
+new_data_species_loss <- tibble(Phos_max_Vonlanthen = seq(from = min(df_lm_4$Phos_max_Vonlanthen), to = max(df_lm_4$Vonlanthen), length = 15))
+
+# predict model
+prediction_species_loss <- predict.lm(lm_species_loss, newdata = new_data_species_loss, se.fit = TRUE, type = "response")
+
+# combine new data and predictions
+bind_model_species_loss <- cbind(new_data_species_loss, prediction_species_loss) |> 
+  mutate(category ="color")
+
+# Plot Fig. 3
+"#78BE63"
+mycolor <- c("color" = "black")
+# mycolor <- c("color" = "#A0E66A")
+"98D47C"
+
+bind_model_species_loss|> 
+  ggplot(aes(Phos_max, fit, color = category, fill = category)) +
+  geom_line(linewidth = 1.2) +
+  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
+  theme_classic(base_size = 20) +
+  ylab("Species loss") +
+  xlab("Maximum total phosphorus (µg/L)") +
+  # ylim(1,3) +
+  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
+
+```
+## Figure 4b
 ```{r}
@@
 # dev.off()
+
+```
+
+##Figure 4c
+```{r}
+
+lm_test <- lm(mean_rdiv ~ Species_loss, data = df_lm_4)
+
+# checking model assumptions
+summary(lm_test)
+# shapiro.test(resid(lm_species_loss))
+# lmtest::bptest(lm_species_loss)
+# plot(lm_species_loss)
+
+# new data
+new_data_test <- tibble(Species_loss = seq(from = min(df_lm_4$Species_loss), to = max(df_lm_4$Species_loss), length = 15))
+
+# predict model
+prediction_test <- predict.lm(lm_test, newdata = new_data_test, se.fit = TRUE, type = "response")
+
+# combine new data and predictions
+bind_model_test <- cbind(new_data_test, prediction_test) |> 
+  mutate(category ="color")
+
+# Plot Fig. 3
+"#78BE63"
+mycolor <- c("color" = "black")
+# mycolor <- c("color" = "#A0E66A")
+"98D47C"
+
+bind_model_test |> 
+  ggplot(aes(Species_loss, fit, color = category, fill = category)) +
+  geom_line(linewidth = 1.2) +
+  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
+  theme_classic(base_size = 20) +
+  ylab("Mean dissimilarity") +
+  xlab("Species loss") +
+  # ylim(1,3) +
+  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
 
@@
 ```
-### OLD CODE Community subgroups
+# OLD CODE Community subgroups
 ####Fig. 4a

### 2024-09-26 01_data_preparation.Rmd
deleted introduction text

--- 2024-07-05 data_preparation.Rmd
@@
 #Introduction
-This is the supplementary information of the Master thesis. This document
-covers the whole analysis performed to assess response diversity between fish communities
-in Swiss peri-alpine lakes. 
-This document has two parts: Data preparation (1) and Data analysis (2). 
-Function to model general additive models is provided separately (see function_GAM.R). 
-Code for figures used in the method figure is also provided separately (see supplementary_method_figure.R)
+Data preparation for analysis.
 
 ```{r}
-# do we want to include zurich and constance separately?
+# read in data
 projet_lac_data <- read.csv("PLDB_final_short_depth_climate_cleantaxa_20052022.csv")

### 2024-09-26 03_drivers_response_diversity_Fig3_Fig4.Rmd
figures 4a-c ready. needed for divergence too

@@
 # Read in data on response diversity metrics in each lake
-resp_div_all <- readRDS("total_models/resp_div_all.rds") |> 
+resp_div_all <- readRDS("total_models/resp_div_all_MA.rds") |> 
   select(temp:fLake)
@@
 
-##Fig. 3
+##Fig. 3a
 Dissimilarity, species richness and categories
@@
 lm1_plot
+
+```
+
+##Figure 3b
+```{r}
 # Lms for all categories: endemic, translcated, non_native and non_endemic_native
@@
 ```
+## Figure 4a Log
+```{r}
+
+# LOG
+df_lm_4 <- df_lm |> 
+  filter(!Lake %in% c("Poschiavo", "Joux", "Maggiore", "Lugano"))
+
+# add log 
+df_lm_4 <- df_lm_4 |> 
+  mutate(P_max_log = log(Phos_max_Vonlanthen))
+
+lm_species_loss_log <- lm(Species_loss ~ P_max_log, data = df_lm_4)
+
+# checking model assumptions
+summary(lm_species_loss_log)
+shapiro.test(resid(lm_species_loss_log))
+lmtest::bptest(lm_species_loss_log)
+plot(lm_species_loss_log)
+
+# new data
+new_data_species_loss_log <- tibble(P_max_log = seq(from = min(df_lm_4$P_max_log), to = max(df_lm_4$P_max_log), length = 15))
+
+# predict model
+prediction_species_loss_log <- predict.lm(lm_species_loss_log, newdata = new_data_species_loss, se.fit = TRUE, type = "response")
+
+# combine new data and predictions
+bind_model_species_loss_log <- cbind(new_data_species_loss_log, prediction_species_loss_log) |> 
+  mutate(category ="color")
+
+# Plot Fig. 3
+"#78BE63"
+mycolor <- c("color" = "black")
+# mycolor <- c("color" = "#A0E66A")
+"98D47C"
+
+bind_model_species_loss_log|> 
+  ggplot(aes(P_max_log, fit, color = category, fill = category)) +
+  geom_line(linewidth = 1.2) +
+  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
+  theme_classic(base_size = 20) +
+  ylab("Species loss") +
+  xlab("Maximum total phosphorus (µg/L)") +
+  # ylim(1,3) +
+  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
+
+```
 ## Figure 4a
 ```{r}
+
 df_lm_4 <- df_lm |> 
   filter(!Lake %in% c("Poschiavo", "Joux", "Maggiore", "Lugano"))
+
+# add log 
+df_lm_4 <- df_lm_4 |> 
+  mutate(P_max_log = log(Phos_max_Vonlanthen))
 
@@
 # new data
-new_data_species_loss <- tibble(Phos_max_Vonlanthen = seq(from = min(df_lm_4$Phos_max_Vonlanthen), to = max(df_lm_4$Vonlanthen), length = 15))
+new_data_species_loss <- tibble(Phos_max_Vonlanthen = seq(from = min(df_lm_4$Phos_max_Vonlanthen), to = max(df_lm_4$Phos_max_Vonlanthen), length = 15))
 
@@
 bind_model_species_loss|> 
-  ggplot(aes(Phos_max, fit, color = category, fill = category)) +
+  ggplot(aes(Phos_max_Vonlanthen, fit, color = category, fill = category)) +
   geom_line(linewidth = 1.2) +
   geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
+  geom_point(data = df_lm_4, aes(x = Phos_max_Vonlanthen, y = Species_loss), inherit.aes = FALSE, color = "black", size = 3) +
   theme_classic(base_size = 20) +
@@
 ```
+
 ## Figure 4b
@@
 summary(lm_test)
-# shapiro.test(resid(lm_species_loss))
-# lmtest::bptest(lm_species_loss)
-# plot(lm_species_loss)
+shapiro.test(resid(lm_test))
+lmtest::bptest(lm_test)
+plot(lm_test)
 

### 2025-04-07 02_models_1.Rmd
added creating folders, runs models with all species and only with temp

--- 2024-07-08 models_derivatives_response_diversity.Rmd
@@
 ---
-title: "Models, derivatives, response diversity"
+title: "02_models_1"
 author: "Sophie Moreau"
-date: "2024-07-05"
+date: "2024-09-27"
 output: html_document
@@
 ```
-
-###Predictions of GAMs
-
-```{r}
-# read in the dataframes
-df_1 <- readRDS("data_frame_models/df_binomial_gam")
-df_2 <- readRDS("data_frame_models/df_abundance_gam")
-df_3 <- readRDS("data_frame_models/df_binomial_re")
-df_4 <- readRDS("data_frame_models/df_abundance_re")
-
-# we have three different functions. One predictions() does not simply any models
-# predictions_temp() has multiple models that are simplified (see supplementary 
-# figures for the disucssion), predictions_depth_temp() models all species with 
-# abundance > 20 with both coviarates temperature and depth.
-
-# For downstream analysis for thermal response diversity models from predictions_temp()
-# are used.
-
-# temperature models with simplified models 
-# predictions take VERY LONG! and two folders are needed to store the 
-# DHARMa and prediction output inside: 
-# total_models/gam_check
-# needs a folder called total_models/predictions/temp
-
-predictions_temp(df_1)
-predictions_temp(df_2)
-predictions_temp(df_3)
-predictions_temp(df_4)
-
-simplified_predictions_all <- list.files(path = "total_models/predictions/temp", pattern = ".rds", full.names = TRUE) |>
-  map_dfr(readRDS)
-
-# save total predictions as RDS
-saveRDS(simplified_predictions_all, "total_models/df_pred_all.rds")
-
-# you can directly use the dataframe df_pred_all for downstream analysis
-
-```
-###Simplification and temp depth models
-```{r}
-# both other functions:
-
-# you can also directly use the dataframe df_pred_no_simplification for decisions about
-# which models should be simplified 
-# and
-# depth_temp_pred_all for the sensitivity testing of the temp + depth models
-
-# testing for which models need to be simplified
-# needs a folder: total_models/predictions/temp_test
-
-# all temperature models, no simplificaiton
-# predictions(df_1)
-# predictions(df_2)
-# predictions(df_3)
-# predictions(df_4)
-# 
-# no_exclusion_predictions_all <- list.files(path = "total_models/predictions/temp_test", pattern = ".rds", full.names = TRUE) |>
-#   map_dfr(readRDS)
-# 
-# # save total predictions as RDS
-# saveRDS(no_exclusion_predictions_all, "total_models/df_pred_no_simplification.rds")
-
-# testing if modelling temp and depth improves the models
-# needs a folder: total_models/predictions/depth_temp
-
-# sensitivity
-# predictions_depth_temp(df_1)
-# predictions_depth_temp(df_2)
-# predictions_depth_temp(df_3)
-# predictions_depth_temp(df_4)
-# 
-# # issue with cottus sp po
-
-# # statistics of the gamms with dept and temp
-# total_models <- bind_rows(df_1, df_2, df_3, df_4)
-# 
-# stats_gamms(total_models)
-# 
-# #  compile data frame with all predictions from the saved species' model predictions
-# 
-# depth_temp_predictions_all <- list.files(path = "total_models/predictions/depth_temp", pattern = ".rds", full.names = TRUE) |>
-#   map_dfr(readRDS)
-# 
-# # save total predictions as RDS
-# saveRDS(depth_temp_predictions_all, "total_models/depth_temp_pred_all.rds")
-
-```
-
-####Fig. 2b
-New figure 2b with categories across all lakes
-
-```{r, fig.show='hide'}
-
-# Load total predictions
-model_predictions <- readRDS("total_models/df_pred_all.rds")
-
-# check the species list
-model_predictions$species <- as.factor(model_predictions$species)
-levels(model_predictions$species)
-
-# read in the category table and rename categories
-species_across_lakes <-  read_xlsx("species_category_across_lakes.xlsx") |> 
-  select(-notes) |> 
-    mutate(category = ifelse(category == "native", "non_endemic_native", category)) |>
-  mutate(category2 = factor("all"))
-
-species_across_lakes$category <- as.factor(species_across_lakes$category)
-levels(species_across_lakes$category)
-
-#merge predictions with categories
-predictions_categories <- merge(model_predictions, species_across_lakes)
-
-# change category names for facet_wrap
-category_names <- c(
-  `endemic` = "endemic",
-  `non_endemic_native` = "non-endemic native",
-  `non_native` = "non-native"
-)
-
-# colors for the plot
-mycolors1 <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8")
-
-# Rescale the predictions to 0-1 
-predictions_categories_rescaled <- predictions_categories |>
-  group_by(species) |>
-  mutate(fit_rescaled = (fit - min(fit)) / (max(fit) - min(fit)))
-
-plot_2b <- predictions_categories_rescaled |>
-  ggplot(aes(temp, fit_rescaled, group = species, color = category)) +
-   geom_line(linewidth = 0.5) +
-  theme_bw(base_size = 16) +
-  ylab("Abundance") +
-  xlab("Temperature") +
-  scale_color_manual(values = mycolors1, guide = NULL) +
-  theme(panel.grid.major = element_blank(), 
-        panel.grid.minor = element_blank(), 
-        legend.position = "bottom",
-        legend.title = element_blank())
-
-plot_2b_final <- plot_2b + facet_grid(~category, labeller = as_labeller(category_names))
-
-plot_2b_final
-
-# save plot as TIFF
-# tiff(paste("total_models/plots/plot_2b.tiff", sep = ""), units="in", width=10, height=4, res=300)
-# 
-# plot(plot_2b_final)
-# 
-# dev.off()
-
-```
-
-####Fig. Supplementary predictions plot
-```{r}
-# also using predictions_categories from chunk above
-# trying to group by category and color
-
-str(categories_supp)
-
-categories_supp <- predictions_categories |>
-  arrange(category) |> 
-  mutate(species = fct_reorder(species, as.numeric(category))) |> 
-  ggplot(aes(temp, fit)) +
-  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit), fill = factor(category)), alpha = 0.3) +
-  geom_line(aes(color = factor(category))) +
-  theme_bw() +
-  facet_wrap(~species, scale = "free_y", labeller = labeller(species = label_wrap_gen(width = 10))) +
-  theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
-  scale_color_manual(values = mycolors1, aesthetics = c("color", "fill")) +
-  xlab("Temperature") +
-  ylab("Abundance")
-
-categories_supp
-
-# tiff(paste("total_models/plots/supplementary_predictions.tiff", sep = ""), units="in", width = 12, height=8, res=300)
-# plot(categories_supp)
-# 
-# # Closing the graphical device
-# dev.off()
-
-```
-##Sensitivity testing
-###NEW Fig. S1
-Testing if depth-driven temperature models can adequately fit abundance data compared to a temperature and depth model.
-
-```{r}
-temp_only <- readRDS("total_models/df_pred_all.rds") |> 
-  mutate(model = "temp")
-
-temp_depth <- readRDS("total_models/depth_temp_pred_all.rds") |> 
-  mutate(model = "temp_depth")
-
-sensitivity_predictions <- bind_rows(temp_only, temp_depth)
-
-supplementary_sensitivity <- sensitivity_predictions |>
-    ggplot(aes(temp, fit, color = model, fill = model)) +
-    geom_line() +
-    # geom_ribbon(aes(ymin = (fit_rescaled - se.fit), ymax = (fit_rescaled + se.fit)), color = NA, alpha = 0.1) +
-    theme_bw() +
-    # labs(title = paste("Sensitivity Test", i)) +
-    scale_color_manual(values = c("#D73027", "#08306B"), aesthetics = c("color", "fill")) +
-     facet_wrap(~species, scale = "free", labeller = labeller(species = label_wrap_gen(width = 10))) +
-    theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
-    ylab("Abundance") +
-    xlab("Temperature") +
-   theme(panel.grid.major = element_blank(), 
-        panel.grid.minor = element_blank(), 
-        legend.position = "right",
-        legend.title = element_blank())
-
-# # save the plot as TIFF
-# tiff(paste("total_models/plots/supplementary_sensitivity_predictions.tiff", sep = ""), units="in", width = 12, height=8, res=300)
-# plot(supplementary_sensitivity)
-# 
-# # Closing the graphical device
-# dev.off()
-
-```
-####AIC comparison 
-```{r}
-# to be added
-```
-
-#Derivatives calculation
-
-##Data preparation
-We calculated derivatives for each species per lake.
-
-```{r}
-
-# Derivatives() will ignore any random effect smooths it encounters in object.
-
-# Table of pecies that were recorded with electro, fishbase or eawag, following Table 16 in Projet Lac Synthesis Report. 
-# Plus the minimum and maximum temp in each lake -> needed for the new data for 
-# calculating derivatives
-
-species_lake <- read_xlsx("species_occurrences_lakes.xlsx") 
-str(species_lake)
-
-species_lake$fLake <- as.factor(species_lake$Lake)
-species_lake$fProtocol <- as.factor(species_lake$Protocol)
-species_lake$Species <- as.factor(species_lake$Species)
-
-#Loading all four dfs to calculate derivatives
-
-df_binomial_gam <- readRDS("data_frame_models/df_binomial_gam")
-df_binomial_gam$fLake <- as.factor(df_binomial_gam$Lake)
-df_binomial_gam$fProtocol <- as.factor(df_binomial_gam$Protocol)
-
-df_abundance_gam <- readRDS("data_frame_models/df_abundance_gam")
-df_abundance_gam$fProtocol <- as.factor(df_abundance_gam$Protocol)
-df_abundance_gam$fLake <- as.factor(df_abundance_gam$Lake)
-
-df_binomial_re <- readRDS("data_frame_models/df_binomial_re")
-df_binomial_re$fLake <- as.factor(df_binomial_re$Lake)
-df_binomial_re$fProtocol <- as.factor(df_binomial_re$Protocol)
-
-df_abundance_re <- readRDS("data_frame_models/df_abundance_re")
-df_abundance_re$fLake <- as.factor(df_abundance_re$Lake)
-df_abundance_re$fProtocol <- as.factor(df_abundance_re$Protocol)
-
-```
-
-##Derivatives
-```{r}
-
-# Now four different loops are run to calculate derivatives of each species 
-# in each lake
-
-# Model 1: Binomial data without random intercepts
-
-species_list <- df_binomial_gam |>
-  distinct(Species) |>
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-# loop through the list of species
-for (i in species_list) {
-  data <- df_binomial_gam |>
-    filter(Species == i)
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'), family = binomial)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  # loop through all lakes where the species is present, lake list is based on the information from
-  # Projet Lac
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    # using minimum and maximum temp in the lake for derivatives
-     newdata <- tibble(mean_last_7days = seq(from = min(data_lake$temp, na.rm = TRUE),
-                                             to = max(data_lake$temp, na.rm = TRUE), length = 200),
-                       fProtocol = factor("VERT"))
-
-     # calculate derivatives
-     derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-       mutate(fLake = factor(j)) |>
-       mutate(species = factor(i)) |>
-       rename(temp = mean_last_7days)
-
-     saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-  }
-}
-
-# Model 2.1: Abundance data and no random intercept for lake -> but binomial
-# Coregonus_sp_benthic_profundal did not run with ZIP and was thus
-# run with binomial.
-
-species_list <- df_abundance_gam |> 
-  # binomial one
-  filter(Species == "Coregonus_sp_benthic_profundal") |>
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_gam |> 
-    filter(Species == i)
-
-  # binomial family instead of ZIP
-  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) +
-                           s(fProtocol, bs = 're'), family = binomial())
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |> 
-      rename(temp = mean_last_7days)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-  }
-}
-
-# Model 2.2: Abundance data and no random intercept for lake -> ZIP
-# all other species in this group run with ZIP
-
-species_list <- df_abundance_gam |>
-  filter(!Species == "Coregonus_sp_benthic_profundal") |>
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_gam |> 
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'),
-                         family = ziP())
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = mean_last_7days)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-  }
-}
-
-# Model 3: Binomial data and random intercept for lake
-
-species_list <- df_binomial_re |> 
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_binomial_re |> 
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = "re")
-                         + s(fProtocol, bs = 're'), family = binomial)
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  unique_lakes <- distinct(data, Lake) |> 
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    # take all lakes where the species is present and take random one for 
-    # the data included in derivatives()
-    random_lake <- sample(unique_lakes, 1)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fProtocol = factor("VERT"), fLake = factor(random_lake))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = mean_last_7days)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-
-  }
-}
-
-# Model 4.1: Abundance data and random intercept for lake, simplified models
-# multiple species in this group were simplified and run with binomial
-
-species_list <- df_abundance_re |>
-  # simplified models
-  filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
-                        "Cyprinus_carpio","Phoxinus_csikii", "Salmo_trutta")) |>
-  distinct(Species) |>
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_re |>
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
-                         +  s(fProtocol, bs = 're'), family = binomial)
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  unique_lakes <- distinct(data, Lake) |>
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    random_lake <- sample(unique_lakes, 1)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fLake = factor(random_lake), fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = mean_last_7days)
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-
-  }
-}
-
-# Model 4.1: Abundance data and random intercept for lake, complex models
-
-species_list <- df_abundance_re |> 
-  # exclusion of simplified models
-  filter(!Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
-                        "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta")) |>
-  distinct(Species) |> 
-  pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
-
-  data <- df_abundance_re |> 
-    filter(Species == i)
-
-  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
-                         +  s(fProtocol, bs = 're'), family = ziP())
-
-  lake_data <- species_lake |>
-    filter(Species == i)
-
-  lake_list <- distinct(lake_data, Lake) |>
-    pull()
-
-  unique_lakes <- distinct(data, Lake) |> 
-    pull()
-
-  for (j in lake_list){
-
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-      filter(Lake == j)
-
-    random_lake <- sample(unique_lakes, 1)
-
-    newdata <- tibble(mean_last_7days = seq(
-      from = min(data_lake$temp, na.rm = TRUE),
-      to = max(data_lake$temp, na.rm = TRUE), length = 200),
-      fLake = factor(random_lake), fProtocol = factor("VERT"))
-
-    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-      mutate(fLake = factor(j)) |>
-      mutate(species = factor(i)) |>
-      rename(temp = mean_last_7days)
-
-    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
-
-  }
-}
-
-# save all the derivatives in one file
-# 
-df_deriv <- list.files(path = "total_models/derivatives", pattern = ".rds", full.names = TRUE) |>
-  map_dfr(readRDS)
-
-saveRDS(df_deriv, "total_models/df_deriv_all.rds")
-
-```
-
-#Testing derivatives
-##Simplification of models
-```{r}
-# to get the derivatives before simplifying any model: 
-# copy all other derivatives to a new folder called total_models/derivatives_no_simplification and then run this code.
-
-# Model 4: before simplification
-species_list <- df_abundance_re |>
-# #   # no simplification
-filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
-                     "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta")) |>
-distinct(Species) |> 
-pull(Species)
-
-species_list <- sort(species_list)
-
-derivatives <- list()
-gam_output <- list()
-model_prediction <- list()
-
-for (i in species_list) {
- data <- df_abundance_re |> 
-     filter(Species == i)
-
- gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're') +  s(fProtocol, bs = 're'), family = ziP())
-
- lake_data <- species_lake |>
-   filter(Species == i)
-
- lake_list <- distinct(lake_data, Lake) |>
-   pull()
-
- unique_lakes <- distinct(data, Lake) |> 
-     pull()
-
-  for (j in lake_list){
-    data_lake <- species_lake |>
-      filter(Species == i) |>
-     filter(Lake == j)
-
-     random_lake <- sample(unique_lakes, 1)
-
-     newdata <- tibble(mean_last_7days = seq(
-       from = min(data_lake$temp, na.rm = TRUE),
-       to = max(data_lake$temp, na.rm = TRUE), length = 200),
-       fLake = factor(random_lake), fProtocol = factor("VERT"))
-
-     derivatives <- derivatives(gam_output[[i]], data = newdata) |>
-       mutate(fLake = factor(j)) |>
-       mutate(species = factor(i)) |>
-       rename(temp = mean_last_7days)
-
-     saveRDS(derivatives, paste0("total_models/derivatives_no_simplification/derivatives_", i, "_",  j, ".rds"))
-
-   }
- }
-
-# save total derivatives as RDS (no exclusions)
-#
-df_deriv_no_excl <- list.files(path = "total_models/derivatives_no_simplification", pattern = ".rds", full.names = TRUE) |>
-  map_dfr(readRDS)
-#
-saveRDS(df_deriv_no_excl, "total_models/df_deriv_all_no_excl.rds")
-
-```
-
-##Fig. S3ab
-
-```{r}
-# cyprinus carpio and lepomis should not be simplified
-
-# Data of the derivatives before simplifying multiple models
-# without changes the 5 species were not run with simplified models. 
-all_models_derivatives_no_excl <- readRDS("total_models/df_deriv_all_no_excl.rds")
-df_deriv_no_excl <- as_tibble(all_models_derivatives_no_excl)
-
-# Plot Figure S3a
-#calculating maximum and mean derivatives for each species per lake
-
-df_quantiles <- df_deriv_no_excl |> 
-  group_by(species, fLake) |> 
-  mutate(max_derivative = max(.derivative)) |> 
-  mutate(mean_derivative = mean(.derivative)) |> 
-  ungroup()
-
-# calculate quantiles of derivatives
-transform(quantile(df_quantiles$.derivative,
-                   c(0,0.01,0.05,0.1,0.25,0.5,0.75,0.85,0.9,0.95,0.99,1)))
-
-# add column with the category of quantile each derivative is in
-df_quantiles$groups <- cut(df_quantiles$.derivative,              
-                     breaks = c(-1.579943e+06, -2.706252e+00, -1.269492e+00,
-                                -3.348207e-01, 1.844665e-03, 2.262231e-01,
-                                9.864616e-01, 2.018455e+01, 3.013806e+01, 1.648215e+06,
-                                1.303406e+07, 1.406101e+07),
-                     labels = c("0-1%", "1-5%", "5-10%", "10-25%", "25-50%", "50-75%", "75-85%", "85-90%", "90-95%",
-                                "95-99%", "99-100%"))
-
-# Plot S3a: histogram of species' mean derivatives per lake
-plot_percentiles <- df_quantiles |> 
-  # Salmo trutta derivative is too large to be plotted
-  filter(species != "Salmo_trutta") |> 
-  ggplot(aes(x = mean_derivative, fill = factor(species))) + 
-  geom_histogram(aes(y = after_stat(count / sum(count))), binwidth = 1.5) +
-  scale_y_continuous(labels = scales::percent) +
-  scale_fill_manual(values = c("Cyprinus_carpio" = "#80CDC1",
-                                "Alburnus_arborella" = "#8C510A",
-                                "Barbatula_sp_Lineage_I"="#DFC27D",
-                               "Phoxinus_csikii" = "#01665E",  "Salmo_trutta" = "#F46D43"))
-
-plot_percentiles
-
-# Plot S3a add lines for the percentiles
-plot_percentiles_1 <- plot_percentiles +
-  geom_vline(xintercept = 3.013806e+01, color = "red") +
-  annotate("text", x= 26, y = 0.12, label="90th percentile", angle=90) +
-  geom_vline(xintercept = 2.018455e+01, color = "red") +
-  annotate("text", x= 16, y = 0.12, label="85th percentile", angle=90) +
-  theme_classic(base_size = 13) +
-  xlab("Mean species derivative per lake") +
-  ylab("Percentage") +
- guides(fill = guide_legend(
-    title = "Species")) +
-  ylim(0, 0.2)
-
-plot_percentiles_1
-
-# save as TIFF
-# tiff(paste("total_models/plots/plot_deriv_percentiles_1.tiff", sep = ""), units="in", width=12, height=6, res=300)
-# 
-# plot(plot_percentiles_1)
-# 
-# dev.off()
-
-# above the 85th percentile
-
-df_quantiles |> 
-  filter(.derivative > 2.018455e+01) |> 
-  distinct(species)
-
-# 1 Alburnus_arborella    
-# 2 Barbatula_sp_Lineage_I
-# 3 Cyprinus_carpio       
-# 4 Phoxinus_csikii       
-# 5 Salmo_trutta
-
-# NEW: upon inclusion of Lepomis gibbosus
-# above 90% = 3.013806e+01
-df_quantiles |> 
-  filter(.derivative > 3.013806e+01) |> 
-  distinct(species)
-
-# 1 Alburnus_arborella    
-# 2 Barbatula_sp_Lineage_I
-# 3 Phoxinus_csikii       
-# 4 Salmo_trutta
-
-# 5 species above 90% -> those will be simplified to binomial 
-# 1 Alburnus_arborella    
-# 2 Barbatula_sp_Lineage_I
-# 3 Cyprinus_carpio       
-# 4 Phoxinus_csikii       
-# 5 Salmo_trutta
-
-# Plotting Figure S3b
-
-# just keeping mean and maximum derivative values for each lake and species
-data_quantiles <- df_quantiles |> 
-  distinct(mean_derivative, max_derivative, fLake, species)
-
-# Breaking mean derivatives in the same percentiles as before
-
-data_quantiles$percentiles <- cut(data_quantiles$mean_derivative,              
-                     breaks = c(-1.579943e+06, -2.706252e+00, -1.269492e+00,
-                                -3.348207e-01, 1.844665e-03, 2.262231e-01,
-                                9.864616e-01, 2.018455e+01, 3.013806e+01, 1.648215e+06,
-                                1.303406e+07, 1.406101e+07),
-                     labels = c("0-1%", "1-5%", "5-10%", "10-25%", "25-50%", "50-75%", "75-85%", "85-90%", "90-95%",
-                                "95-99%", "99-100%"))
-
-# Plot S3b tile plot of mean derivatives per species in each lake
-mean_deriv_plot <- data_quantiles |> 
-  ggplot(aes(fLake, y = fct_reorder(species, mean_derivative), fill= percentiles)) + 
-  geom_tile() +
-  scale_fill_manual(breaks = levels(data_quantiles$percentiles),
-                    values = rev(brewer.pal(11, "BrBG")))
-
-derivative_percentiles_2 <- mean_deriv_plot +
-  guides(fill = guide_legend(title = "percentiles (mean derivative)", reverse = TRUE)) +
-  xlab("") +
-  ylab("") +
-  theme_classic(base_size = 16)
-
-derivative_percentiles_2
-
-# save plot as TIFF
-# tiff(paste("total_models/plots/plot_deriv_percentiles_2.tiff", sep = ""), units="in", width=19, height=12, res=300)
-# 
-# plot(derivative_percentiles_2)
-# 
-# dev.off()
-
-```
-
-##Fig. S2: Overview derivatives
-Overview of derivatives in each lake community
-
-```{r}
-# Read in derivative data
-all_models_derivatives <- readRDS("total_models/df_deriv_all.rds")
-all_derivatives <- as_tibble(all_models_derivatives)
-
-str(all_derivatives)
-
-# Highlighting species with interesting derivatives: 
-species_group <- all_derivatives |> 
-  filter(species %in% c("Phoxinus_csikii", "Barbatula_sp_Lineage_II", "Cottus_sp_Po_profundal")) |> 
-  distinct(species) |> 
-  pull(species)
-
-# Plot S2: derivatives included in each lake community
-overview_derivatives <- all_derivatives |>
-  arrange(fLake) |> 
-  ggplot(aes(temp, .derivative, color = species)) +
-  geom_line() +
-  facet_wrap(~fLake) +
-  theme_bw(base_size = 16) +
-  scale_color_viridis(discrete = TRUE, breaks = species_group, option = "G") +
-  xlab("Temperature") +
-  ylab("Derivatives") +
-  theme(panel.grid.major = element_blank(), 
-        panel.grid.minor = element_blank(), 
-        legend.position = "right",
-        legend.title = element_blank()) +
-  geom_hline(yintercept = 0, color = "#7A7A7A")
-
-overview_derivatives
-
-# save as TIFF
-# tiff(paste("total_models/plots/overview_derivatives_lake.tiff", sep = ""), units="in", width=11, height=7, res=300)
-# 
-# plot(overview_derivatives)
-# 
-# dev.off()
-```
-
-#Calculating thermal response diversity
-
-```{r}
-# Function provided by Ross et al. "How to measure response diversity" used to calculate
-# response diversity
-
-# read in function
-source(here("functions.R"))
-
-# load derivatives data
-all_models_derivatives <- readRDS("total_models/df_deriv_all.rds")
-
-all_deriv <- as_tibble(all_models_derivatives)
-
-str(all_deriv)
-
-levels(all_deriv$species)
-
-#Calculating response diversity for each lake community
-
-all_deriv$fLake <- as.character(all_deriv $fLake)
-
-lakes_list <- all_deriv |> 
-  distinct(fLake) |> 
-  pull(fLake)
-
-str(lakes_list)
-
-lakes_list <- sort(lakes_list)
-
-species_overview <- tibble()
-
-all_deriv$species <- as.character(all_deriv$species)
-all_deriv$species <- as.factor(all_deriv$species)
-
-str(all_deriv )
-
-# Loop to get response diversity measures for each lake 
-# loop through list of lakes and calculate response diversity in the lake communities
-
-for (i in lakes_list){
-
-  data <- all_deriv |>
-    select(.derivative, temp, species, fLake) |> 
-    filter(fLake == i)
-
-  df_resp_div <- data |>
-    pivot_wider(
-      names_from = species,
-      values_from = .derivative)
-
-  # apply the function from Ross et al. 
-  # Dissimilarity
-  df_resp_div$rdiv <- apply(df_resp_div[,-(1:2), drop = FALSE], 1, resp_div, sign_sens = F)
-  # Divergence
-  df_resp_div$sign <- apply(df_resp_div[,-(1:2), drop = FALSE], 1, resp_div, sign_sens = T)
-  # Median Dissimilarity
-  df_resp_div$Med <- median(df_resp_div$rdiv)
-
-  # save response diversity metrics per lake
-  saveRDS(df_resp_div, paste0("total_models/resp_div_lakes/df_resp_div_", i, ".rds"))
-
-}
-
-# save the response diversity per lake into one file
-resp_div_no_excl <- list.files(path = "total_models/resp_div_lakes", pattern = ".rds", full.names = TRUE) |>
-  map_dfr(readRDS) |>
-  relocate(rdiv, Med, sign, .after = temp)
-
-saveRDS(resp_div_no_excl,"total_models/resp_div_all.rds")
-```

### 2025-04-07 01_data_preparation.Rmd
folder creation

@@
   distinct(LakeBasin)
+
+# create folder needed in the directory
+
+folder_name <- "data_frame_models"
+if (!dir.exists(folder_name)){
+  dir.create(folder_name)}
+
 ```
@@
 We compiled four data frames for the four types of general additive models that 
-were run: Species with 1) binomial dataoccurring in one lake,
+were run: Species with 1) binomial data occurring in one lake,
 2) abundance data occurring in one lake, 3) binomial data 

### 2025-04-07 02_models_1.Rmd
first testing of the whole flow with all species with temp GAM

@@
 
-# function for the temperature only GAMs
-source(here("function_GAM.R"))
+# function for the temperature only GAMs with all run as expected
+source(here("predictions_function.R"))
+```
 
-# function for temp and depth models 
-source(here("function_temp_depth.R"))
+```{r}
+# prepare folders
+
+main_folder <- "gams_testing_1"
+if (!dir.exists(main_folder)){
+  dir.create(main_folder)}
+
+subfolders <- c("gam_check", "predictions", "derivatives", "resp_div")
+
+for (sub in subfolders){
+  sub_path <- file.path(main_folder, sub)
+  if(!dir.exists(sub_path)){
+    dir.create(sub_path)
+  }
+}
+
+# read in the dataframes
+df_1 <- readRDS("data_frame_models/df_binomial_gam")
+df_2 <- readRDS("data_frame_models/df_abundance_gam")
+df_3 <- readRDS("data_frame_models/df_binomial_re")
+df_4 <- readRDS("data_frame_models/df_abundance_re")
+
+predictions(df_1)
+predictions(df_2)
+predictions(df_3)
+predictions(df_4)
+
+total_models <- bind_rows(df_1, df_2, df_3, df_4)
+
+stats_predictions(total_models)
+
+df_predictions_all <- list.files(path = "gams_testing_1/predictions", pattern = ".rds", full.names = TRUE) |>
+  map_dfr(readRDS)
+
+# save total predictions as RDS
+saveRDS(df_predictions_all, "gams_testing_1/df_pred_all.rds")
+
 ```
+
+####Fig. 2
+Overview of all model predictions
+
+```{r, fig.show='hide'}
+# Load total predictions
+model_predictions <- readRDS("gams_testing_1/df_pred_all.rds")
+
+# check the species list
+model_predictions$species <- as.factor(model_predictions$species)
+levels(model_predictions$species)
+
+# read in the category table and rename categories
+species_across_lakes <-  read_xlsx("species_category_across_lakes.xlsx") |> 
+  select(-notes) |> 
+    mutate(category = ifelse(category == "native", "non_endemic_native", category)) |>
+  mutate(category = ifelse(category == "non_native_region", "non_endemic_native_and_translocated", category)) |> 
+  mutate(category2 = factor("all"))
+
+species_across_lakes$category <- as.factor(species_across_lakes$category)
+levels(species_across_lakes$category)
+
+#merge predictions with categories
+predictions_categories <- merge(model_predictions, species_across_lakes)
+
+# EVTL NEW PLOT 2
+# change category names for facet_wrap
+category_names <- c(
+  `endemic` = "endemic",
+  `endemic_and_translocated` = "translocated endemic",
+  `non_endemic_native` = "non-endemic native",
+  `non_endemic_native_and_translocated` = "translocated non-endemic native",
+  `non_native` = "non-native"
+)
+
+# colors for the plot
+mycolors1 <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
+                "endemic_and_translocated" = "#F2AD63", "non_endemic_native_and_translocated" = "#F2AD63")
+
+# Rescale the predictions to 0-1 
+predictions_categories_rescaled <- predictions_categories |>
+  group_by(species) |>
+  mutate(fit_rescaled = (fit - min(fit)) / (max(fit) - min(fit)))
+
+plot_pred <- predictions_categories_rescaled |>
+  ggplot(aes(temp, fit_rescaled, group = species, color = category)) +
+  geom_line() +
+  # how can I get the right color?
+   # geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), alpha = 0.3) +
+  theme_bw(base_size = 16) +
+  ylab("Abundance") +
+  xlab("Temperature") +
+  scale_color_manual(values = mycolors1, guide = NULL) +
+  theme(panel.grid.major = element_blank(), 
+        panel.grid.minor = element_blank(), 
+        legend.position = "bottom",
+        legend.title = element_blank())
+
+plot_category_predictions <- plot_pred + facet_grid(~category, labeller = as_labeller(category_names))
+
+plot_category_predictions
+
+# trying to group by category and color
+figure_test_2 <- predictions_categories |>
+  ggplot(aes(temp, fit)) +
+  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit), fill = factor(category)), alpha = 0.3) +
+  geom_line(aes(color = factor(category))) +
+  theme_bw() +
+  facet_wrap(~species + category, scale = "free_y", labeller = labeller(species = label_wrap_gen(width = 10))) +
+  theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
+  scale_color_manual(values = mycolors1, aesthetics = c("color", "fill")) +
+  xlab("Temperature") +
+  ylab("Abundance")
+
+figure_test_2
+# tiff(paste("total_models/plots/draft_all_predictions.tiff", sep = ""), units="in", width = 12, height=8, res=300)
+# plot(figure_test_2)
+# 
+# # Closing the graphical device
+# dev.off()
+
+```
+
+##Derivatives calculation
+
+We calculated derivatives for each species per lake.
+
+```{r}
+
+# Derivatives() will ignore any random effect smooths it encounters in object.
+
+# Table of pecies that were recorded with electro, fishbase or eawag, following Table 16 in Projet Lac Synthesis Report. 
+# Plus the minimum and maximum temp in each lake -> needed for the new data for 
+# calculating derivatives
+
+species_lake <- read_xlsx("species_occurrences_lakes.xlsx") 
+str(species_lake)
+
+species_lake$fLake <- as.factor(species_lake$Lake)
+species_lake$fProtocol <- as.factor(species_lake$Protocol)
+species_lake$Species <- as.factor(species_lake$Species)
+
+#Loading all four dfs to calculate derivatives
+
+df_binomial_gam <- readRDS("data_frame_models/df_binomial_gam")
+df_binomial_gam$fLake <- as.factor(df_binomial_gam$Lake)
+df_binomial_gam$fProtocol <- as.factor(df_binomial_gam$Protocol)
+
+df_abundance_gam <- readRDS("data_frame_models/df_abundance_gam")
+df_abundance_gam$fProtocol <- as.factor(df_abundance_gam$Protocol)
+df_abundance_gam$fLake <- as.factor(df_abundance_gam$Lake)
+
+df_binomial_re <- readRDS("data_frame_models/df_binomial_re")
+df_binomial_re$fLake <- as.factor(df_binomial_re$Lake)
+df_binomial_re$fProtocol <- as.factor(df_binomial_re$Protocol)
+
+df_abundance_re <- readRDS("data_frame_models/df_abundance_re")
+df_abundance_re$fLake <- as.factor(df_abundance_re$Lake)
+df_abundance_re$fProtocol <- as.factor(df_abundance_re$Protocol)
+
+```
+
+###Temp models
+```{r}
+
+# Now four different loops are run to calculate derivatives of each species 
+# in each lake
+
+# Model 1: Binomial data without random intercepts
+
+species_list <- df_binomial_gam |>
+  distinct(Species) |>
+  pull(Species)
+
+species_list <- sort(species_list)
+
+derivatives <- list()
+gam_output <- list()
+model_prediction <- list()
+
+# loop through the list of species
+for (i in species_list) {
+  data <- df_binomial_gam |>
+    filter(Species == i)
+
+  lake_data <- species_lake |>
+    filter(Species == i)
+
+  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'), family = binomial)
+
+  lake_list <- distinct(lake_data, Lake) |>
+    pull()
+
+  # loop through all lakes where the species is present, lake list is based on the information from
+  # Projet Lac
+  for (j in lake_list){
+
+    data_lake <- species_lake |>
+      filter(Species == i) |>
+      filter(Lake == j)
+
+    # using minimum and maximum temp in the lake for derivatives
+     newdata <- tibble(mean_last_7days = seq(from = min(data_lake$temp, na.rm = TRUE),
+                                             to = max(data_lake$temp, na.rm = TRUE), length = 200),
+                       fProtocol = factor("VERT"))
+
+     # calculate derivatives
+     derivatives <- derivatives(gam_output[[i]], data = newdata) |>
+       mutate(fLake = factor(j)) |>
+       mutate(species = factor(i))
+       # rename(temp = mean_last_7days)
+
+     saveRDS(derivatives, paste0("gams_testing_1/derivatives/derivatives_", i, "_",  j, ".rds"))
+
+  }
+}
+
+# Model 2.1: Abundance data and no random intercept for lake -> but binomial
+# Coregonus_sp_benthic_profundal did not run with ZIP and was thus
+# run with binomial.
+
+species_list <- df_abundance_gam |> 
+  # binomial one
+  filter(Species == "Coregonus_sp_benthic_profundal") |>
+  distinct(Species) |> 
+  pull(Species)
+
+species_list <- sort(species_list)
+
+derivatives <- list()
+gam_output <- list()
+model_prediction <- list()
+
+for (i in species_list) {
+
+  data <- df_abundance_gam |> 
+    filter(Species == i)
+
+  # binomial family instead of ZIP
+  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) +
+                           s(fProtocol, bs = 're'), family = binomial())
+
+  lake_data <- species_lake |>
+    filter(Species == i)
+
+  lake_list <- distinct(lake_data, Lake) |>
+    pull()
+
+  for (j in lake_list){
+
+    data_lake <- species_lake |>
+      filter(Species == i) |>
+      filter(Lake == j)
+
+    newdata <- tibble(mean_last_7days = seq(
+      from = min(data_lake$temp, na.rm = TRUE),
+      to = max(data_lake$temp, na.rm = TRUE), length = 200),
+      fProtocol = factor("VERT"))
+
+    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
+      mutate(fLake = factor(j)) |>
+      mutate(species = factor(i))
+
+    saveRDS(derivatives, paste0("gams_testing_1/derivatives/derivatives_", i, "_",  j, ".rds"))
+  }
+}
+
+# Model 2.2: Abundance data and no random intercept for lake -> ZIP
+# all other species in this group run with ZIP
+
+species_list <- df_abundance_gam |>
+  filter(!Species == "Coregonus_sp_benthic_profundal") |>
+  distinct(Species) |> 
+  pull(Species)
+
+species_list <- sort(species_list)
+
+derivatives <- list()
+gam_output <- list()
+model_prediction <- list()
+
+for (i in species_list) {
+
+  data <- df_abundance_gam |> 
+    filter(Species == i)
+
+  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'),
+                         family = ziP())
+
+  lake_data <- species_lake |>
+    filter(Species == i)
+
+  lake_list <- distinct(lake_data, Lake) |>
+    pull()
+
+  for (j in lake_list){
+
+    data_lake <- species_lake |>
+      filter(Species == i) |>
+      filter(Lake == j)
+
+    newdata <- tibble(mean_last_7days = seq(
+      from = min(data_lake$temp, na.rm = TRUE),
+      to = max(data_lake$temp, na.rm = TRUE), length = 200),
+      fProtocol = factor("VERT"))
+
+    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
+      mutate(fLake = factor(j)) |>
+      mutate(species = factor(i))
+
+    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
+  }
+}
+
+```
+
+```{r}
+
+# Model 3: Binomial data and random intercept for lake
+
+species_list <- df_binomial_re |> 
+  distinct(Species) |> 
+  pull(Species)
+
+species_list <- sort(species_list)
+
+derivatives <- list()
+gam_output <- list()
+model_prediction <- list()
+
+for (i in species_list) {
+
+  data <- df_binomial_re |> 
+    filter(Species == i)
+
+  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = "re")
+                         + s(fProtocol, bs = 're'), family = binomial)
+
+  lake_data <- species_lake |>
+    filter(Species == i)
+
+  lake_list <- distinct(lake_data, Lake) |>
+    pull()
+
+  unique_lakes <- distinct(data, Lake) |> 
+    pull()
+
+  for (j in lake_list){
+
+    data_lake <- species_lake |>
+      filter(Species == i) |>
+      filter(Lake == j)
+
+    # take all lakes where the species is present and take random one for 
+    # the data included in derivatives()
+    random_lake <- sample(unique_lakes, 1)
+
+    newdata <- tibble(mean_last_7days = seq(
+      from = min(data_lake$temp, na.rm = TRUE),
+      to = max(data_lake$temp, na.rm = TRUE), length = 200),
+      fProtocol = factor("VERT"), fLake = factor(random_lake))
+
+    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
+      mutate(fLake = factor(j)) |>
+      mutate(species = factor(i))
+
+    saveRDS(derivatives, paste0("gams_testing_1/derivatives/derivatives_", i, "_",  j, ".rds"))
+
+  }
+}
+
+```
+
+```{r}
+
+# Model 4.1: Abundance data and random intercept for lake, complex models
+
+species_list <- df_abundance_re |> 
+  distinct(Species) |> 
+  pull(Species)
+
+species_list <- sort(species_list)
+
+derivatives <- list()
+gam_output <- list()
+model_prediction <- list()
+
+for (i in species_list) {
+
+  data <- df_abundance_re |> 
+    filter(Species == i)
+
+  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
+                         +  s(fProtocol, bs = 're'), family = ziP())
+
+  lake_data <- species_lake |>
+    filter(Species == i)
+
+  lake_list <- distinct(lake_data, Lake) |>
+    pull()
+
+  unique_lakes <- distinct(data, Lake) |> 
+    pull()
+
+  for (j in lake_list){
+
+    data_lake <- species_lake |>
+      filter(Species == i) |>
+      filter(Lake == j)
+
+    random_lake <- sample(unique_lakes, 1)
+
+    newdata <- tibble(mean_last_7days = seq(
+      from = min(data_lake$temp, na.rm = TRUE),
+      to = max(data_lake$temp, na.rm = TRUE), length = 200),
+      fLake = factor(random_lake), fProtocol = factor("VERT"))
+
+    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
+      mutate(fLake = factor(j)) |>
+      mutate(species = factor(i))
+
+    saveRDS(derivatives, paste0("gams_testing_1/derivatives/derivatives_", i, "_",  j, ".rds"))
+
+  }
+}
+
+```
+
+```{r}
+
+# save all the derivatives in one file
+
+df_deriv <- list.files(path = "gams_testing_1/derivatives", pattern = ".rds", full.names = TRUE) |> 
+  map_dfr(readRDS)
+
+saveRDS(df_deriv, "gams_testing_1/df_deriv_all.rds")
+
+```
+
+##Calculating thermal response diversity
+
+```{r}
+# Function provided by Ross et al. "How to measure response diversity" used to calculate
+# response diversity
+
+# read in function
+source(here("functions.R"))
+
+# load derivatives data
+all_models_derivatives <- readRDS("gams_testing_1/df_deriv_all.rds")
+
+all_deriv <- as_tibble(all_models_derivatives)
+
+str(all_deriv)
+
+levels(all_deriv$species)
+
+#Calculating response diversity for each lake community
+
+all_deriv$fLake <- as.character(all_deriv $fLake)
+
+lakes_list <- all_deriv |> 
+  distinct(fLake) |> 
+  pull(fLake)
+
+str(lakes_list)
+
+lakes_list <- sort(lakes_list)
+
+species_overview <- tibble()
+
+all_deriv$species <- as.character(all_deriv$species)
+all_deriv$species <- as.factor(all_deriv$species)
+
+str(all_deriv )
+
+# Loop to get response diversity measures for each lake 
+# loop through list of lakes and calculate response diversity in the lake communities
+for (i in lakes_list){
+
+  data <- all_deriv |>
+    select(mean_last_7days, fLake, .derivative, species) |> 
+    filter(fLake == i)
+
+  df_resp_div <- data |>
+    pivot_wider(
+      names_from = species,
+      values_from = .derivative)
+
+  # apply the function from Ross et al. 
+  # Dissimilarity
+  df_resp_div$rdiv <- apply(df_resp_div[,-(1:2), drop = FALSE], 1, resp_div, sign_sens = F)
+  # Divergence
+  df_resp_div$sign <- apply(df_resp_div[,-(1:2), drop = FALSE], 1, resp_div, sign_sens = T)
+  # Median Dissimilarity
+  df_resp_div$Med <- median(df_resp_div$rdiv)
+
+  # save response diversity metrics per lake
+  saveRDS(df_resp_div, paste0("gams_testing_1/resp_div/df_resp_div_", i, ".rds"))
+
+}
+
+# save the response diversity per lake into one file
+resp_div_no_excl <- list.files(path = "gams_testing_1/resp_div", pattern = ".rds", full.names = TRUE) |>
+  map_dfr(readRDS) |>
+  relocate(rdiv, Med, sign, .after = mean_last_7days)
+
+saveRDS(resp_div_no_excl,"gams_testing_1/resp_div_all.rds")
+```
+
+#Drivers of thermal response diversity
+
+Preparing a df with the information on past eutrophication, species richness, and species categorization in each lake.
+```{r}
+# Df with properties of lakes
+lake_info <- read_xlsx("lake_info.xlsx")
+
+# select the columns we use
+lake_selection <- lake_info |> 
+  select(Lake, Phos_max, Max_depth, Lake_area, Species_loss, Phos_max_Vonlanthen)
+
+# information about categorization in each lake for the subgroups
+species_endemism <- read_excel("species_category_per_lake.xlsx") |> 
+  rename(endemism = detail_category)
+
+# re-calculate species richness per lake
+species_endemism_short <- species_endemism |> 
+  # filter(species != "Lepomis_gibbosus") |> 
+  select(-sum_species) |> 
+  group_by(fLake) |> 
+  mutate(sum_species = sum(num_species))
+
+species_endemism_short$endemism <- as.factor(species_endemism_short$endemism)
+
+str(species_endemism_short)
+
+# adding column with number of species per category in each lake
+df_species_endemism <- species_endemism_short |> 
+  select(-species, - num_species) |> 
+  add_count(fLake, endemism) |> 
+  rename(count = n) |> 
+  distinct(fLake, count, sum_species, endemism)
+
+str(df_species_endemism)
+
+df_species_endemism$count <- as.numeric(df_species_endemism$count)
+
+# making it wide 
+df_species_endemism_long <- df_species_endemism |> 
+  pivot_wider(names_from = endemism, values_from = count) |> 
+  mutate_if(is.numeric, ~replace(., is.na(.), 0)) |> 
+  rename(Lake = fLake)
+
+str(df_species_endemism_long)
+
+# merge endemism info and lake info
+
+endemism_lake <- merge(df_species_endemism_long, lake_selection)
+
+# Read in data on response diversity metrics in each lake
+resp_div_all <- readRDS("gams_testing_1/resp_div_all.rds") |> 
+  select(mean_last_7days:fLake)
+
+# select response diversity columns we need
+rdiv <- resp_div_all |>  
+  select(mean_last_7days, fLake, rdiv, sign) |>
+  drop_na()
+
+# calculate mean and maximum values and rename Lake
+df_rdiv <- rdiv |> 
+  group_by(fLake) |> 
+  mutate(mean_rdiv = mean(rdiv)) |> 
+  mutate(max_rdiv = max(rdiv)) |> 
+  mutate(mean_sign = mean(sign)) |> 
+  mutate(max_sign = max(sign)) |> 
+  distinct(fLake, mean_rdiv, max_rdiv, mean_sign, max_sign) |> 
+  rename(Lake = fLake)
+
+# combine lake properties, species subgroups and response diversity
+
+df_lm <- merge(df_rdiv, endemism_lake)
+
+str(df_lm)
+
+# phosphorus values as numeric
+df_lm$Phos_max <- as.numeric(df_lm$Phos_max)
+df_lm$Phos_max_Vonlanthen <- as.numeric(df_lm$Phos_max_Vonlanthen)
+
+```
+
+##Fig. 3a
+Dissimilarity, species richness and categories
+
+```{r}
+# Multiple linear regression for mean dissimilarity explained by species richness and past eutrophication: 
+# lm(mean_rdiv  ~ species richness + eutrophication)
+# Predict manually with new data eutrophication value fixed at mean and species richness min() to max()
+# do the same the other way around: mean_rdiv  ~  eutrophication + species richness, with species richness fixed at mean and eutrophication min() to max()species richness +
+#   prepare these two plots
+
+# Phosphorus data on Poschiavo is not available
+df_lm_excl <- df_lm |> 
+  filter(Lake != "Poschiavo")
+
+# first lm for species richness ~ maximum phosphorus
+
+lm1 <- lm(mean_rdiv ~ sum_species + Phos_max, data = df_lm_excl)
+
+# checking model assumptions
+summary(lm1)
+shapiro.test(resid(lm1))
+lmtest::bptest(lm1)
+plot(lm1)
+
+# model assumptions validated
+
+# manually predict data
+new_data <- tibble(sum_species = seq(from = min(df_lm_excl$sum_species), to = max(df_lm_excl$sum_species), length = 50), Phos_max = mean(df_lm_excl$Phos_max))
+
+# predict lm
+prediction_lm1 <- predict.lm(lm1, newdata = new_data, se.fit = TRUE, type = "response")
+
+# combine new data and predictions
+bind_model1 <- cbind(new_data, prediction_lm1) |> 
+  mutate(category ="color")
+
+mycolor = "black"
+
+# pltting the first linear regression
+lm1_plot <- bind_model1 |> 
+  ggplot(aes(sum_species, fit,  color = category, fill = category)) +
+  geom_line(linewidth = 1.2) +
+  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
+  theme_classic(base_size = 20) +
+  ylab("Mean dissimilarity") +
+  xlab("Species richness") +
+  # ylim(1,3) +
+  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
+
+lm1_plot
+
+```
+
+##Figure 3b
+```{r}
+# Lms for all categories: endemic, translcated, non_native and non_endemic_native
+# with data on all lakes
+# Endemic
+lm_endemic <- lm(mean_rdiv ~ endemic, data = df_lm)
+
+# checking model assumptions
+summary(lm_endemic)
+shapiro.test(resid(lm_endemic))
+lmtest::bptest(lm_endemic)
+plot(lm_endemic)
+
+new_data_a <- tibble(endemic = seq(from = min(df_lm$endemic), to = max(df_lm$endemic),length = 15))
+
+prediction_endemic <- predict.lm(lm_endemic, newdata = new_data_a, se.fit = TRUE, type = "response")
+
+df_lm_endemic <- cbind(new_data_a, prediction_endemic) |> 
+  mutate(category = factor("endemic")) |> 
+  rename(species_richness = endemic)
+
+# Non-native
+
+lm_nn <- lm(mean_rdiv ~ non_native, data = df_lm)
+
+# checking model assumptions
+summary(lm_nn)
+shapiro.test(resid(lm_nn))
+lmtest::bptest(lm_nn)
+plot(lm_nn)
+
+new_data_b <- tibble(non_native = seq(from = min(df_lm$non_native), to = max(df_lm$non_native),
+                                   length = 15))
+
+prediction_nn <- predict.lm(lm_nn, newdata = new_data_b, se.fit = TRUE, type = "response")
+
+df_lm_nn <- cbind(new_data_b, prediction_nn)  |> 
+  mutate(category = factor("non_native")) |> 
+  rename(species_richness = non_native)
+
+# Non-endemic native
+
+lm_nne <- lm(mean_rdiv ~ native, data = df_lm)
+
+# checking model assumptions
+summary(lm_nne)
+shapiro.test(resid(lm_nne))
+lmtest::bptest(lm_nne)
+plot(lm_nne)
+
+new_data_c <- tibble(native = seq(from = min(df_lm$native), to = max(df_lm$native),
+                                      length = 15))
+
+prediction_nne <- predict.lm(lm_nne, newdata = new_data_c, se.fit = TRUE, type = "response")
+
+df_lm_nne <- cbind(new_data_c, prediction_nne)  |> 
+  mutate(category = factor("non_endemic_native")) |> 
+  rename(species_richness = native)
+
+# Translocated (non-native to region) species
+
+lm_trans <- lm(mean_rdiv ~ non_native_region, data = df_lm)
+
+# checking model assumptions
+summary(lm_trans)
+shapiro.test(resid(lm_trans))
+lmtest::bptest(lm_trans)
+plot(lm_trans)
+
+new_data_d <- tibble(non_native_region = seq(from = min(df_lm$non_native_region), to = max(df_lm$non_native_region),
+                                              length = 15))
+
+prediction_trans <- predict.lm(lm_trans, newdata = new_data_d, se.fit = TRUE, type = "response")
+
+df_lm_trans <- cbind(new_data_d, prediction_trans)  |> 
+  mutate(category = factor("translocated")) |> 
+  rename(species_richness = non_native_region)
+
+# Combine all four lm predictions
+df_all_lms <- rbind(df_lm_endemic, df_lm_nn, df_lm_nne, df_lm_trans)
+
+mycolors <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
+               "translocated" = "#F2AD63")
+
+plot_all_lms <- df_all_lms|> 
+  ggplot(aes(species_richness, fit, color = category, fill = category, linetype = category)) +  
+  geom_line(linewidth = 1) +
+  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), color = NA, alpha = 0.3) +
+  theme_classic(base_size = 16) +
+  ylab("Mean dissimilarity") +
+  xlab("Species richness") +
+  scale_fill_manual(values = mycolors,
+                    aesthetics = c("color", "fill"),
+                    labels = c("endemic", "non-native", "non-endemic native", "translocated")) +
+   scale_linetype_manual(values = c("endemic" = "solid",
+                                   "non_native" = "dashed",
+                                   "non_endemic_native" = "solid",
+                                   "translocated" = "dashed"), labels = c("endemic", "non-native", "non-endemic native", "translocated"))
+
+plot_all_lms
+
+# save as TIFF
+# tiff(paste("total_models/plots/lm_all.tiff", sep = ""), units="in", width=11, height=7, res=300)
+# # plot(ggarrange(depth1, depth2, ncol = 2))
+# # plot science discussion
+# plot(plot_all_lms)
+# 
+# dev.off()
+
+```
+
+## Figure 3 Divergence
+
+```{r}
+
+# first lm for species richness ~ maximum phosphorus
+
+lm3 <- lm(mean_sign ~ sum_species + Phos_max, data = df_lm_excl)
+
+# checking model assumptions
+summary(lm3)
+shapiro.test(resid(lm3))
+lmtest::bptest(lm3)
+plot(lm3)
+
+# model assumptions validated
+
+# manually predict data
+new_data3 <- tibble(sum_species = seq(from = min(df_lm_excl$sum_species), to = max(df_lm_excl$sum_species), length = 50), Phos_max = mean(df_lm_excl$Phos_max))
+
+# predict lm
+prediction_lm3 <- predict.lm(lm3, newdata = new_data3, se.fit = TRUE, type = "response")
+
+# combine new data and predictions
+bind_model3 <- cbind(new_data3, prediction_lm3) |> 
+  mutate(category ="color")
+
+lm3_plot <- bind_model3 |> 
+  ggplot(aes(sum_species, fit,  color = category, fill = category)) +
+  geom_line(linewidth = 1.2) +
+  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
+  theme_classic(base_size = 20) +
+  ylab("Mean divergence") +
+  xlab("Species richness") +
+  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
+
+# Lms for all categories: endemic, translcated, non_native and non_endemic_native
+# with data on all lakes
+# Endemic
+deriv_lm1 <- lm(mean_sign ~ endemic, data = df_lm)
+
+# checking model assumptions
+summary(deriv_lm1)
+shapiro.test(resid(deriv_lm1))
+lmtest::bptest(deriv_lm1)
+plot(deriv_lm1)
+
+new_data_1 <- tibble(endemic = seq(from = min(df_lm$endemic), to = max(df_lm$endemic),
+                                     length = 15))
+
+prediction_endemic_1 <- predict.lm(deriv_lm1, newdata = new_data_1, se.fit = TRUE, type = "response")
+
+df_deriv_endemic <- cbind(new_data_1, prediction_endemic_1) |> 
+  mutate(category = factor("endemic")) |> 
+  rename(species_richness = endemic)
+
+# Non-native
+
+deriv_lm2 <- lm(mean_sign ~ non_native, data = df_lm)
+
+# checking model assumptions
+summary(deriv_lm2)
+shapiro.test(resid(deriv_lm2))
+lmtest::bptest(deriv_lm2)
+plot(deriv_lm2)
+
+new_data_2 <- tibble(non_native = seq(from = min(df_lm$non_native), to = max(df_lm$non_native),
+                                   length = 15))
+
+prediction_2 <- predict.lm(deriv_lm2, newdata = new_data_2, se.fit = TRUE, type = "response")
+
+df_deriv_nn <- cbind(new_data_2, prediction_2)  |> 
+  mutate(category = factor("non_native")) |> 
+  rename(species_richness = non_native)
+
+# Non-endemic native
+
+deriv_lm3 <- lm(mean_sign ~ native, data = df_lm)
+
+# checking model assumptions
+summary(deriv_lm3)
+shapiro.test(resid(deriv_lm3))
+lmtest::bptest(deriv_lm3)
+plot(deriv_lm3)
+
+new_data_3 <- tibble(native = seq(from = min(df_lm$native), to = max(df_lm$native),
+                                      length = 15))
+
+prediction_3 <- predict.lm(deriv_lm3, newdata = new_data_3, se.fit = TRUE, type = "response")
+
+df_deriv_nne <- cbind(new_data_3, prediction_3)  |> 
+  mutate(category = factor("non_endemic_native")) |> 
+  rename(species_richness = native)
+
+# Translocated (non-native to region) species
+
+deriv_lm4 <- lm(mean_sign ~ non_native_region, data = df_lm)
+
+# checking model assumptions
+summary(deriv_lm4)
+shapiro.test(resid(deriv_lm4))
+lmtest::bptest(deriv_lm4)
+plot(deriv_lm4)
+
+new_data_4 <- tibble(non_native_region = seq(from = min(df_lm$non_native_region), to = max(df_lm$non_native_region),
+                                              length = 15))
+
+prediction_4 <- predict.lm(deriv_lm4, newdata = new_data_4, se.fit = TRUE, type = "response")
+
+df_deriv_trans <- cbind(new_data_4, prediction_4)  |> 
+  mutate(category = factor("translocated")) |> 
+  rename(species_richness = non_native_region)
+
+# Combine all four lm predictions
+df_deriv_lms <- rbind(df_deriv_endemic, df_deriv_nn, df_deriv_nne, df_deriv_trans)
+
+mycolors <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
+               "translocated" = "#F2AD63")
+
+lm_divergence_all <- df_deriv_lms|> 
+  ggplot(aes(species_richness, fit, color = category, fill = category)) +
+  geom_line(linewidth = 1, linetype = "dashed") +
+  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), color = NA, alpha = 0.3) +
+  theme_classic(base_size = 16) +
+  ylab("Mean divergence") +
+  xlab("Species richness") +
+  scale_fill_manual(values = mycolors,
+                    aesthetics = c("color", "fill"), 
+                    labels = c("endemic", "non-native", "non-endemic native", "translocated"),
+                    name = NULL)
+
+lm_divergence_all
+
+# save as TIFF
+# tiff(paste("total_models/plots/lm_divergence_all.tiff", sep = ""), units="in", width=11, height=7, res=300)
+# # plot(ggarrange(depth1, depth2, ncol = 2))
+# # plot science discussion
+# plot(lm_divergence_all)
+# 
+# dev.off()
+```
+## Figure 4a Log
+```{r}
+
+# LOG
+df_lm_4 <- df_lm |> 
+  filter(!Lake %in% c("Poschiavo", "Joux", "Maggiore", "Lugano"))
+
+# add log 
+df_lm_4 <- df_lm_4 |> 
+  mutate(P_max_log = log(Phos_max_Vonlanthen))
+
+lm_species_loss_log <- lm(Species_loss ~ P_max_log, data = df_lm_4)
+
+# checking model assumptions
+summary(lm_species_loss_log)
+shapiro.test(resid(lm_species_loss_log))
+lmtest::bptest(lm_species_loss_log)
+plot(lm_species_loss_log)
+
+# new data
+new_data_species_loss_log <- tibble(P_max_log = seq(from = min(df_lm_4$P_max_log), to = max(df_lm_4$P_max_log), length = 15))
+
+# predict model
+prediction_species_loss_log <- predict.lm(lm_species_loss_log, newdata = new_data_species_loss_log, se.fit = TRUE, type = "response")
+
+# combine new data and predictions
+bind_model_species_loss_log <- cbind(new_data_species_loss_log, prediction_species_loss_log) |> 
+  mutate(category ="color")
+
+# Plot Fig. 3
+"#78BE63"
+mycolor <- c("color" = "black")
+# mycolor <- c("color" = "#A0E66A")
+"98D47C"
+
+test <- bind_model_species_loss_log|> 
+  ggplot(aes(P_max_log, fit, color = category, fill = category)) +
+  geom_line(linewidth = 1.2) +
+  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
+  theme_classic(base_size = 20) +
+  ylab("Species loss") +
+  xlab("Maximum total phosphorus (µg/L)") +
+  # ylim(1,3) +
+  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
+
+test
+```
+## Figure 4a
+```{r}
+
+df_lm_4 <- df_lm |> 
+  filter(!Lake %in% c("Poschiavo", "Joux", "Maggiore", "Lugano"))
+
+# add log 
+df_lm_4 <- df_lm_4 |> 
+  mutate(P_max_log = log(Phos_max_Vonlanthen))
+
+lm_species_loss <- lm(Species_loss ~ Phos_max_Vonlanthen, data = df_lm_4)
+
+# checking model assumptions
+summary(lm_species_loss)
+shapiro.test(resid(lm_species_loss))
+lmtest::bptest(lm_species_loss)
+plot(lm_species_loss)
+
+# new data
+new_data_species_loss <- tibble(Phos_max_Vonlanthen = seq(from = min(df_lm_4$Phos_max_Vonlanthen), to = max(df_lm_4$Phos_max_Vonlanthen), length = 15))
+
+# predict model
+prediction_species_loss <- predict.lm(lm_species_loss, newdata = new_data_species_loss, se.fit = TRUE, type = "response")
+
+# combine new data and predictions
+bind_model_species_loss <- cbind(new_data_species_loss, prediction_species_loss) |> 
+  mutate(category ="color")
+
+# Plot Fig. 3
+"#78BE63"
+mycolor <- c("color" = "black")
+# mycolor <- c("color" = "#A0E66A")
+"98D47C"
+
+bind_model_species_loss|> 
+  ggplot(aes(Phos_max_Vonlanthen, fit, color = category, fill = category)) +
+  geom_line(linewidth = 1.2) +
+  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
+  geom_point(data = df_lm_4, aes(x = Phos_max_Vonlanthen, y = Species_loss), inherit.aes = FALSE, color = "black", size = 3) +
+  theme_classic(base_size = 20) +
+  ylab("Species loss") +
+  xlab("Maximum total phosphorus (µg/L)") +
+  # ylim(1,3) +
+  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
+
+```
+
+## Figure 4b
+```{r}
+# second lm for maximum phosphorus ~ species richness
+lm2 <- lm(mean_rdiv ~ Phos_max + sum_species, data = df_lm_excl)
+
+# checking model assumptions
+summary(lm2)
+shapiro.test(resid(lm2))
+lmtest::bptest(lm2)
+plot(lm2)
+
+# new data
+new_data2 <- tibble(Phos_max = seq(from = min(df_lm_excl$Phos_max), to = max(df_lm_excl$Phos_max),length = 50), sum_species = mean(df_lm_excl$sum_species))
+
+# predict model
+prediction_lm2 <- predict.lm(lm2, newdata = new_data2, se.fit = TRUE, type = "response")
+
+# combine new data and predictions
+bind_model2 <- cbind(new_data2, prediction_lm2) |> 
+  mutate(category ="color")
+
+# Plot Fig. 3
+"#78BE63"
+mycolor <- c("color" = "black")
+# mycolor <- c("color" = "#A0E66A")
+"98D47C"
+
+lm2_plot <- bind_model2 |> 
+  ggplot(aes(Phos_max, fit, color = category, fill = category)) +
+  geom_line(linewidth = 1.2) +
+  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
+  theme_classic(base_size = 20) +
+  ylab("Mean dissimilarity") +
+  xlab("Maximum total phosphorus (µg/L)") +
+  # ylim(1,3) +
+  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
+
+lm2_plot
+# save as TIFF
+# tiff(paste("total_models/plots/lm_species_richness_eutroph.tiff", sep = ""), units="in", width=12, height=5, res=300)
+# 
+# plot(figure_3)
+# 
+# # Closing the graphical device
+# dev.off()
+
+```
+
+##Figure 4c
+```{r}
+
+lm_test <- lm(mean_rdiv ~ Species_loss, data = df_lm_4)
+
+# checking model assumptions
+summary(lm_test)
+shapiro.test(resid(lm_test))
+lmtest::bptest(lm_test)
+plot(lm_test)
+
+# new data
+new_data_test <- tibble(Species_loss = seq(from = min(df_lm_4$Species_loss), to = max(df_lm_4$Species_loss), length = 15))
+
+# predict model
+prediction_test <- predict.lm(lm_test, newdata = new_data_test, se.fit = TRUE, type = "response")
+
+# combine new data and predictions
+bind_model_test <- cbind(new_data_test, prediction_test) |> 
+  mutate(category ="color")
+
+# Plot Fig. 3
+"#78BE63"
+mycolor <- c("color" = "black")
+# mycolor <- c("color" = "#A0E66A")
+"98D47C"
+
+bind_model_test |> 
+  ggplot(aes(Species_loss, fit, color = category, fill = category)) +
+  geom_line(linewidth = 1.2) +
+  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
+  theme_classic(base_size = 20) +
+  ylab("Mean dissimilarity") +
+  xlab("Species loss") +
+  # ylim(1,3) +
+  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
+
+```
+
+## Figure 4 Divergence
+
+```{r}
+
+# second lm for maximum phosphorus ~ species richness
+lm4 <- lm(mean_sign ~ Phos_max + sum_species, data = df_lm_excl)
+
+# checking model assumptions
+summary(lm4)
+shapiro.test(resid(lm4))
+lmtest::bptest(lm4)
+plot(lm4)
+
+# new data
+new_data4 <- tibble(Phos_max = seq(from = min(df_lm_excl$Phos_max), to = max(df_lm_excl$Phos_max),length = 50), sum_species = mean(df_lm_excl$sum_species))
+
+# predict model
+prediction_lm4 <- predict.lm(lm4, newdata = new_data4, se.fit = TRUE, type = "response")
+
+# combine new data and predictions
+bind_model4 <- cbind(new_data4, prediction_lm4) |> 
+  mutate(category ="color")
+
+# Plot Fig. 3
+mycolor <- c("color" = "#88D44F")
+# mycolor <- c("color" = "#A0E66A")
+"98D47C"
+
+lm3_plot <- bind_model3 |> 
+  ggplot(aes(sum_species, fit,  color = category, fill = category)) +
+  geom_line(linewidth = 1.2) +
+  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
+  theme_classic(base_size = 20) +
+  ylab("Mean divergence") +
+  xlab("Species richness") +
+  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
+
+lm4_plot <- bind_model4 |> 
+  ggplot(aes(Phos_max, fit, color = category, fill = category)) +
+  geom_line(linewidth = 1.2) +
+  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
+  theme_classic(base_size = 20) +
+  ylab("Mean divergence") +
+  xlab("Maximum total phosphorus (µg/L)") +
+  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)
+
+# combine both plots
+figure_divergence <- ggarrange(lm3_plot, lm4_plot)
+
+figure_divergence
+# save as TIFF
+# tiff(paste("total_models/plots/lm_divergence_multiple_linear.tiff", sep = ""), units="in", width=12, height=5, res=300)
+# 
+# plot(figure_divergence)
+# 
+# # Closing the graphical device
+# dev.off()
+
+```

### 2025-04-07 predictions_function.R
new function for predictions for gam testing 1

@@ -1,0 +1,621 @@
+
+predictions <- function(df){
+  require(broom)
+  require(tidyverse)
+  require(DHARMa)
+  require(mgcv)
+  require(gratia)
+  require(mgcViz)
+  
+  species_list <- df |> 
+    distinct(Species) |> 
+    pull(Species)
+  
+  species_list <- sort(species_list)
+  
+  gam_output <- list()
+  model_prediction <- list()
+  derivatives <- list()
+  grid <- list()
+  pred_df <- list()
+  tiff_filename <- list()
+  tiff_file_2 <- list()
+  temp_data <- list()
+  summary <- list()
+  
+  df$fLake <- as.factor(df$Lake)
+  
+  df$fProtocol <- as.factor(df$Protocol)
+  
+  for (i in species_list) {
+    data <- df |> 
+      filter(Species == i) |> 
+      mutate(n_lake = n_distinct(Lake))
+    
+    unique_lakes <- unique(data$fLake)
+    
+    random_lake <- sample(levels(unique_lakes), 1)
+    
+    # species occurring in one lake -> net-type as random effect
+    if(max(data$n_lake) == 1) {
+      
+      # newdata for all species that fall into this category -> occuring
+      # in one lake only
+      grid <- expand.grid(mean_last_7days = seq(
+        from = min(data$mean_last_7days, na.rm = TRUE),
+        to = max(data$mean_last_7days, na.rm = TRUE), by = 0.02),
+        fProtocol = factor("VERT"))
+      
+      # special case, because model not running with ZIP
+      if (i == "Coregonus_sp_benthic_profundal")  { 
+        
+        gam_output <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) +
+                            s(fProtocol, bs = 're'), family = binomial)
+        # prepare residuals
+        simulationOutput <- simulateResiduals(fittedModel = gam_output, plot = F)
+        tiff_filename <- paste("gams_testing_1/gam_check/gam_check_", i, ".tiff", sep = "")
+        tiff(tiff_filename, width = 800, height = 600)
+        print(plot(simulationOutput))
+        dev.off()
+        
+        # Plotting standardized residuals against predictors
+        tiff_file_2 <- paste("gams_testing_1/gam_check/predictor_", i, ".tiff", sep = "")
+        tiff(tiff_file_2, width = 800, height = 600)
[…]

### 2025-04-07 stats_predictions_temp_function.R
stats with temp

@@ -1,0 +1,346 @@
+# statistics overview predictions_temp
+stats_predictions_temp <- function(df){
+  require(broom)
+  require(tidyverse)
+  require(mgcv)
+  require(modelsummary)
+  require(writexl)
+  
+  species_list <- df |> 
+    distinct(Species) |> 
+    pull(Species)
+  
+  species_list <- sort(species_list)
+  
+  gam_output <- list()
+  temp_data <- list()
+  summary <- list()
+  model_summary <- list()
+  total_summary <- list()
+  
+  df$fLake <- as.factor(df$Lake)
+  
+  df$fProtocol <- as.factor(df$Protocol)
+  
+  for (i in species_list) {
+    data <- df |> 
+      filter(Species == i) |> 
+      mutate(n_lake = n_distinct(Lake))
+    
+    # species occurring in one lake -> net-type as random effect
+    if(max(data$n_lake) == 1) {
+      
+      # special case, because model not running with ZIP
+      if (i == "Coregonus_sp_benthic_profundal")  { 
+        
+        gam_output <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) +
+                            s(fProtocol, bs = 're'), family = binomial)
+        
+        # stats
+        glance_summary <- glance(gam_output)
+        tidy_summary <- tidy(gam_output)
+        
+        # Add species column to both summaries
+        glance_summary <- glance_summary |> 
+          mutate(species = i)
+        
+        tidy_summary <- tidy_summary |> 
+          mutate(species = i)
+        
+        # abundance column
+        abundances <- data |> 
+          rename(species = Species) |> 
+          distinct(species, tot_abu)
+        
+        # combine both
+        pre_total_summary <- merge(tidy_summary, glance_summary)
+        
+        total_summary <- merge(pre_total_summary, abundances)
+        
+        # Store the summary in the list
+        summ <- summary(gam_output)
+        
+        dev_explained <- summ$dev.expl
+        total_summary$deviance_explained <- dev_explained
+        
+        chi_sq <- summ[["chi.sq"]][["s(mean_last_7days)"]]
+        
+        total_summary$chi_square <- chi_sq
+        
+        p_value <- summ[["s.pv"]][1]
+        total_summary$p_value <- p_value
+        
+        edf <- summ[["edf"]][1]
+        total_summary$edf <- edf
+        
+        r_sq <- summ[["r.sq"]]
+        total_summary$r_sq <- r_sq
+        
+        model_summary[[i]] <- total_summary
+        
+        
[…]
